---
title: "HUMAnN 3.0 functional and metabolic pathway analysis"
site: workflowr::wflow_site
cache: true
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
  html_document:
    fig_width: 10
    fig_height: 10
editor_options:
  chunk_output_type: console
---

```{r setwd, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

if (Sys.getenv("TERM_PROGRAM") == "vscode") {
  try(
    {
      .vsc.attach()
    },
    silent = TRUE
  )
  try(
    {
      setwd("workflowr")
    },
    silent = TRUE
  )
}
```

```{r report_template}
htmltools::includeHTML("analysis/index_for_workflowr.html")
```

```{r vars}
# Project root, relative to the knit working directory (workflowr/).
# Used to write interactive heatmaps into workflowr/docs/plots/.
base_path <- ".."

taxa.level <- "Species"
taxa.level.short <- "S"
```

```{r libraries}
library(tidyverse)
library(stringr)
library(purrr)
library(mia)
library(miaViz)
library(scater)
library(dplyr)

f <- stringr::str_glue

source("./analysis/rmarkdown_utils.R")
```

HUMAnN, like MetaPhlAn, maps reads to a reference database of microbial genomes. The abundance of known gene families (UniRef90) 
is quantified, and the abundance of _pathways_ of genes is estimated (using the MetaCyc pathways database).

Gene families and pathways can be viewed at the species level, or the community (whole-sample) level - we will focus on community level abundances here.
Pathway abundances are proportional to the number of complete "copies" of the pathway in the community.
Relative abundance values are shown here are copies-per-millon (CPM), where the abundances for any one sample sum to 1 million. This helps normalize for library size.

Additional gene family and pathway groupings are available but not shown - these are EggNOG Orthologies, EC numbers, Gene Ontology (GO) anntoations, KEGG Orthologies, PFAM protein families and [MetaCyc Reactions](https://biocyc.org/META/class-tree?object=Reactions).

```{r read_tsvs}
# We read in the main gene and pathway abundance tables, and cleanup some column names
# We drop all the taxonomic subdivision rows so we only have 'top level' gene and pathway abundance rows

gene.relab <- readr::read_tsv(
  # "../pipeline/humann/output/genefamilies.relab.tsv"
  "../pipeline/humann/output/genefamilies.cpm.tsv"
)

pathway.relab <- readr::read_tsv(
  # "../pipeline/humann/output/pathabundance.relab.tsv"
  "../pipeline/humann/output/pathabundance.cpm.tsv"
)

# Rename the ugly columns with # in front
colnames(gene.relab)[colnames(gene.relab) == "# Gene Family"] <- "Gene_Family"
colnames(pathway.relab)[colnames(pathway.relab) == "# Pathway"] <- "Pathway"

# Simplify sample (column) names
colnames(gene.relab) <- c(
  colnames(gene.relab[, 1]),
  str_extract(colnames(gene.relab[, -1]), "^[^_]+")
)
colnames(pathway.relab) <- c(
  colnames(pathway.relab[, 1]),
  str_extract(colnames(pathway.relab[, -1]), "^[^_]+")
)

# abundance & prevalence filter - drop pathway / family rows
# 20% of samples need to be greater than 0.0001 relative abudance for a given pathway, otherwise the row is dropped
print(f("Removing low abundance/low prevalence pathways/families (at least 20% of samples must have abundance > 0.0001)"))
print(f("Pathways before abundance/prevalence filter: {dim(pathway.relab)[1]}"))
print(f("Gene families before abundance/prevalence filter: {dim(gene.relab)[1]}"))
pathway.relab <- pathway.relab[apply(pathway.relab, 1, function(x) sum(x > 0.0001) > 0.2 * ncol(pathway.relab)), ]
gene.relab <- gene.relab[apply(gene.relab, 1, function(x) sum(x > 0.0001) > 0.2 * ncol(gene.relab)), ]
print(f("Pathways after abundance/prevalence filter: {dim(pathway.relab)[1]}"))
print(f("Gene families after abundance/prevalence filter: {dim(gene.relab)[1]}"))

# Make a long format of pathway abundances, Pathway, Organism, Sample, relabundance
pathway.relab.strat.long <- (
  pathway.relab
  %>% gather(Sample, relabundance, -Pathway)
    %>% separate(Pathway, c("Pathway", "Organism"), sep = "\\|", fill = "right")
    %>% mutate(Group = str_extract(Sample, "^(.*?)(?=-)"))
    %>% mutate(PathwayID = str_extract(Pathway, "^(.*?)(?=:)"))
)

# Make a long format of gene family abundances, Gene_Family, Organism, Sample, relabundance
gene.relab.strat.long <- (
  gene.relab
  %>% gather(Sample, relabundance, -Gene_Family)
    %>% separate(Gene_Family, c("Gene_Family", "Organism"), sep = "\\|", fill = "right")
    %>% mutate(Group = str_extract(Sample, "^(.*?)(?=-)"))
)

# Create long dataframes without the species-stratified values, no UNMAPPED/UNINTEGRATED
pathway.relab.long <- pathway.relab.strat.long %>%
  filter(is.na(Organism)) %>%
  select(-Organism) %>%
  filter(!(grepl("^UN", Pathway)))

gene.relab.long <- gene.relab.strat.long %>%
  filter(is.na(Organism)) %>%
  select(-Organism) %>%
  filter(!(grepl("^UN", Gene_Family)))

# Get rid of all the taxonomic subdivisions, keep just overall abundance
pathway.relab.strat <- data.frame(pathway.relab)
gene.relab.strat <- data.frame((gene.relab), row.names = NULL, check.rows = FALSE, check.names = TRUE, stringsAsFactors = default.stringsAsFactors())
gene.relab <- gene.relab %>% filter(!str_detect(Gene_Family, fixed("|")))
pathway.relab <- pathway.relab %>% filter(!str_detect(Pathway, fixed("|")))

path_prefix <- "../pipeline/humann/output/genefamilies"
uniref_level <- "uniref90"
# We make a named list of each of the regrouped classes
gene.relab.groups <- list()
for (grouping in c("eggnog", "go", "ko", "level4ec", "pfam", "rxn")) {
  g <- readr::read_tsv(
    f("{path_prefix}_{uniref_level}_{grouping}.relab.tsv")
  )
  colnames(g)[colnames(g) == "# Gene Family"] <- "Gene_Family"
  colnames(g[, -1]) <- c(colnames(g[, 1]), str_extract(colnames(g[, -1]), "^[^_]+"))
  g <- g %>% filter(!str_detect(Gene_Family, fixed("|")))

  gene.relab.groups[[grouping]] <- g
}
```

```{r include=FALSE, eval=FALSE}
# ## Testing gene family CPM filtering

# gene.relab <- readr::read_tsv(
#   #"../pipeline/humann/output/genefamilies.relab.tsv.gz"
#   "../pipeline/humann/output/genefamilies.cpm.tsv.gz"
# )

# print(f("Gene families before abundance/prevalence filter: {dim(gene.relab)[1]}"))
# gene.relab <- gene.relab[apply(gene.relab[, -1], 1, function(x) sum(x >= 0.0001) > 0.2 * (ncol(gene.relab) - 1)), ]
# print(f("Gene families after abundance/prevalence filter (0.0001 threshold): {dim(gene.relab)[1]}"))


# gene.relab <- readr::read_tsv(
#   #"../pipeline/humann/output/genefamilies.relab.tsv.gz"
#   "../pipeline/humann/output/genefamilies.cpm.tsv.gz"
# )

# print(f("Gene families before abundance/prevalence filter: {dim(gene.relab)[1]}"))
# gene.relab <- gene.relab[apply(gene.relab[, -1], 1, function(x) sum(x >= 0.01) > 0.2 * (ncol(gene.relab) - 1)), ]
# print(f("Gene families after abundance/prevalence filter (0.01 threshold): {dim(gene.relab)[1]}"))

# gene.relab <- readr::read_tsv(
#   #"../pipeline/humann/output/genefamilies.relab.tsv.gz"
#   "../pipeline/humann/output/genefamilies.cpm.tsv.gz"
# )
# print(f("Gene families before abundance/prevalence filter: {dim(gene.relab)[1]}"))
# gene.relab <- gene.relab[apply(gene.relab[, -1], 1, function(x) sum(x >= 10) > 0.2 * (ncol(gene.relab) - 1)), ]
# print(f("Gene families after abundance/prevalence filter (10 cpm threshold): {dim(gene.relab)[1]}"))
```

Also recorded are the abundance of `UNMAPPED` reads (those that failed to map to any reference genome in the database).
`UNGROUPED` or `UNINTEGRATED` features are those that aren't present in the grouped database 
(eg, the UniRef90 family has no annotated reaction in MetaCyc, for instance).
Reads that map but aren't associated with a gene that has a Uniref90 family are labelled `UniRef90_unknown`.

Percentage unmapped reads (for Uniref90 gene families):
```{r}
show_table(data.frame((gene.relab %>% filter(Gene_Family == "UNMAPPED"))[, -1] / 10000), digits = 1)
```

We can see between ~13% and 43% of reads do not map to the ChocoPhlAn (vOct22) database. 
This is not unusual, but worth noting this technique does not annotate the entire microbial diversity present in the sample due.

```{r make_sample_metadata_table}
sample_colnames <- colnames(gene.relab[, -1])

sample_metadata <- data.frame(
  sample = str_extract(sample_colnames, "^[^_]+"),
  group = str_extract(sample_colnames, "^[^-]+")
) %>% mutate(group = stringr::str_to_title(group))
rownames(sample_metadata) <- sample_metadata$sample
sample_metadata$group <- as.factor(sample_metadata$group)
```

```{r mia_loadFromHumann, include=FALSE, eval=FALSE}
gene.tse <- loadFromHumann(
  "../pipeline/humann/output/genefamilies.relab.tsv",
  # colData=sample_metadata,
  assay.type = "relabundance",
  removeTaxaPrefixes = TRUE,
  remove.suffix = TRUE
)

gene.tse$group <- sample_metadata$group

pathway.tse <- loadFromHumann(
  "../pipeline/humann/output/pathabundance.relab.tsv",
  # colData=sample_metadata,
  assay.type = "relabundance",
  removeTaxaPrefixes = TRUE,
  remove.suffix = TRUE
)

pathway.tse$group <- sample_metadata$group
```

```{r abundance_scatter, include=FALSE, eval=FALSE}
tse_glom <- mia::agglomerateByRank(gene.tse, rank = taxa.level, onRankOnly = TRUE)

plotAbundanceDensity(tse_glom,
  layout = "jitter", assay.type = "relabundance",
  n = 20, point_size = 1, point_shape = 19, point_alpha = 0.5,
  colour_by = "group"
) +
  scale_x_log10(label = scales::percent) +
  ggtitle(f("Relative abundance of top {taxa.level}"))
```

## Metabolic pathways

### Metabolic pathway (MetaCyc) heatmap

A hierarchical clustered heatmap of using the abundance of MetaCyc pathways for each sample.
To simplify the plot and only show pathways with larger between-sample diffferences, we only 
show pathways that have high variance between samples (greater than 75th percentle variance).
We show log2(cpm), where cpm is complete pathway copies per million.

- [plots/pathways_heatmap.html](plots/pathways_heatmap.html)

```{r}
library(pheatmap)
library(heatmaply)

pways <- pathway.relab %>% filter((Pathway != "UNMAPPED") & Pathway != "UNINTEGRATED")

# We remove the Abx condition since the pathways are driven almost entirely by Lactococcus specific metabolism
# This obscures interesting patterns between Het and Homo
pways <- pways %>% select(-starts_with("Abx"))

# pways <- pways %>% select(-starts_with("Homo-3"))

pways.mat <- as.matrix(pways[, -1])
rownames(pways.mat) <- pways$Pathway

pways.var <- apply(pways.mat, 1, var) # Calculate variance for each row
threshold <- quantile(pways.var, 0.75) # Set a threshold, (75th percentle)
pways.mat <- pways.mat[pways.var > threshold, ]

rownames(pways.mat) <- pways$Pathway[pways.var > threshold]

pathways.heatmap <- pheatmap(log(pways.mat + 1),
  clustering_distance_cols = "euclidean",
  clustering_distance_rows = "euclidean",
  # kmeans_k=4,
  fontsize_row = 8
)

hm <- heatmaply(
  log(pways.mat + 1),
  k_col = 2, k_row = 2,
  labels_row = rownames(pways.mat),
  labRow = rownames(pways.mat),
  labCol = colnames(pways.mat),
  fontsize_row = 8
)

htmlwidgets::saveWidget(hm, file.path(base_path, "workflowr/docs/plots/pathways_heatmap.html"))

# The non-interactive version has too many rows to interpret
pathways.heatmap
```

A multidimensional scaling (MDS) plot showing dimensionality reduction of the pathway relativate abundance for each sample.

```{r}
library(SingleCellExperiment())

pways <- pathway.relab %>% filter((Pathway != "UNMAPPED") & Pathway != "UNINTEGRATED")

pways.mat <- as.matrix(pways[, -1])
rownames(pways.mat) <- pways$Pathway

pway.se <- SingleCellExperiment(assays = list(relabundance = pways.mat))
pway.se$group <- str_extract(colnames(pways[, -1]), "^[^-]+")

# Bray-Curtis PCoA
# pway.mds <- runMDS(pway.se,
#   FUN = vegan::vegdist,
#   method = "bray",
#   name = "PCoA_BC",
#   assay.type = "relabundance"
# )

# # # Create ggplot object
# p <- plotReducedDim(pway.mds, "PCoA_BC",
#   colour_by = "group"
# )

# e <- attr(reducedDim(pway.mds, "PCoA_BC"), "eig")
# rel_eig <- e / sum(e[e > 0])

###

# Regular MDS
pway.mds <- runMDS(pway.se,
  name = "MDS",
  assay.type = "relabundance"
)

p <- plotReducedDim(pway.mds, "MDS",
  colour_by = "group"
)

# Calculate explained variance
e <- attr(reducedDim(pway.mds, "MDS"), "eig")
rel_eig <- e / sum(e[e > 0])

# Add explained variance for each axis
p <- p + labs(
  x = paste("dim 1 (", round(100 * rel_eig[[1]], 1), "%", ")", sep = ""),
  y = paste("dim 2 (", round(100 * rel_eig[[2]], 1), "%", ")", sep = ""),
  # title = "Bray-Curtis Beta Diversity of Pathways, PCoA",
  subtitle = "All samples"
)

p
```

```{r}
library(SingleCellExperiment())

pways <- pathway.relab %>% filter((Pathway != "UNMAPPED") & Pathway != "UNINTEGRATED")

# We remove the Abx condition since the pathways are driven almost entirely by Lactococcus specific metabolism
# This obscures interesting patterns between Het and Homo
pways <- pways %>% select(-starts_with("Abx"))

# pways <- pways %>% select(-starts_with("Homo-3"))

pways.mat <- as.matrix(pways[, -1])
rownames(pways.mat) <- pways$Pathway

pway.se <- SingleCellExperiment(assays = list(relabundance = pways.mat))
pway.se$group <- str_extract(colnames(pways[, -1]), "^[^-]+")

# Regular MDS
pway.mds <- runMDS(pway.se,
  name = "MDS",
  assay.type = "relabundance"
)

p <- plotReducedDim(pway.mds, "MDS",
  colour_by = "group"
)

# Calculate explained variance
e <- attr(reducedDim(pway.mds, "MDS"), "eig")
rel_eig <- e / sum(e[e > 0])

# Add explained variance for each axis
p <- p + labs(
  x = paste("dim 1 (", round(100 * rel_eig[[1]], 1), "%", ")", sep = ""),
  y = paste("dim 2 (", round(100 * rel_eig[[2]], 1), "%", ")", sep = ""),
  # title = "Bray-Curtis Beta Diversity of Pathways, PCoA",
  subtitle = "Het vs Homo samples"
)

p
```

The Abx samples are very different and cluster together (not surprising, they are mostly Lactococcus).
If we look at the Het/Homo samples with reveals more detail - the Het samples cluster together (more similar within-group pathway abundances), while the Homo samples are more dispersed.
This reflects the higher variance between biological replicates in the Homo samples, as seen in taxonomic abundance analysis previously.

### Differential pathway abundance

We investigate pathways with differential abundance in `Homo` relative to `Het`, using a linear model with log-transformation of CPM abundances (MaAsLin2 method).

```{r maaslin2, include=FALSE}
## MaAsLin2 generalized linear model analysis

# MaAsLin2 fits generalized linear models to estimate differentially abundance taxa between groups (Het/Homo).
# In this case, we use arcsine square-root transformed (AST) counts.

library(Maaslin2)

pways <- data.frame(pathway.relab %>% select(-starts_with("Abx")))
rownames(pways) <- pways$Pathway
metadata <- data.frame(sample_metadata %>% filter(group != "Abx"))
colnames(metadata)[colnames(metadata) == "sample"] <- "Sample"
colnames(metadata)[colnames(metadata) == "group"] <- "Group"
rownames(metadata) <- metadata$Sample

pathway_name_mappings <- createNameMapping(pways$Pathway)

maaslin2_out <- Maaslin2(
  input_data = pways[, -1],
  input_metadata = metadata,
  output = "docs/maaslin_hethomo_pathway_da",
  transform = "LOG",
  # AST is "arcsine square-root transformation"
  # transform = "AST",
  fixed_effects = "Group",
  # you can also fit MLM by specifying random effects
  # random_effects = c(...),
  reference = "Group,Het",
  normalization = "NONE",
  # normalization = "TSS", # if not already in relabundance units
  min_abundance = -Inf,
  standardize = FALSE,
  # filtering was previously performed
  min_prevalence = 0,
  max_pngs = 20,
  plot_heatmap = TRUE,
  heatmap_first_n = 20
)

maaslin2_results <- maaslin2_out$results %>%
  dplyr::rename(Pathway = feature)

maaslin2_results$Pathway <- convertToOriginalNames(
  maaslin2_results$Pathway, pathway_name_mappings
)

# maaslin2_results <- merge(
#   maaslin2_results,
#   refined %>%
#     dplyr::select(bin_id, Genus, Species) %>%
#     mutate(bin_id = str_replace_all(bin_id, "-", ".")), # maaslin munges feature names ...
#   by = "bin_id"
# ) %>%
#   relocate(bin_id, .after = last_col()) %>%
#   relocate(c("Species", "Genus"), .before = everything())
```

#### Full HUMAnN/MaAsLin2 association table

```{r}
# maaslin2_results <- maaslin2_results %>% mutate(PathwayID = str_split(Pathway, "\\.", simplify = TRUE)[, 1:2] %>% apply(1, paste, collapse = "-"))
maaslin2_results <- maaslin2_results %>% mutate(PathwayID = str_split(Pathway, ":", simplify = TRUE)[, 1])

# Turn pathways in table into hyperlinks to MetaCyc
maaslin2_results$PathwayLink <- paste0(
  "<a href=\"https://metacyc.org/META/NEW-IMAGE?object=",
  maaslin2_results$PathwayID, "\" target='_blank'>",
  maaslin2_results$Pathway,
  "</a>"
)
maaslin2_results <- move_column_to_index(maaslin2_results, "PathwayLink", 1)

maaslin2_results %>%
  select(-c(metadata, value, name, Pathway)) %>%
  # filter(qval < 0.05) %>%
  arrange(-coef) %>%
  arrange(qval) %>%
  show_table_extra(escape = FALSE) %>%
  DT::formatStyle(
    "qval",
    target = "row",
    backgroundColor = DT::styleInterval(c(0, 0.05001), c("white", "lightgrey", "white"))
  )
```

> - The `qval` is the adjusted p-value - rows with `qval <= 0.05` are highlighted grey.
> - The `coef` is the model coef, indicating the strength of the association between the pathway and the group (+ve means pathway is more abudant in `Homo`)
> - The `N` column is the total number of data points.
> - The `N.not.zero` column is the total of non-zero data points.

#### Significant HUMAnN/MaAsLin2 association table

Here is a filtered version of the same table, showing only associations with a adjusted p-value <= 0.05, and only pathways detected in 90% of samples.

```{r maaslin2_filtered_table}
maaslin2_results %>%
  select(-c(metadata, value, name, Pathway)) %>%
  filter(qval <= 0.05) %>%
  filter(N.not.zero / N >= 0.9) %>%
  arrange(-coef) %>%
  arrange(qval) %>%
  show_table_extra(escape = FALSE)
```

General themes for the identified differential pathways are basic energy metabolism (glycolsis and TCA cycle), pyrimidine and purine degradation, and of particular interest, on cysteine biosynthesis pathway.

[PWY-6292 - superpathway of L-cysteine biosynthesis (mammalian)](https://metacyc.org/META/NEW-IMAGE?object=PWY-6292) show significant (FDR=0.045) differential pathway abudance, with more complete pathways detected in the `Homo` microbial communities.
(Note that although this is annotated as mammalian, the pathway exists in bacteria and archea).

```{r}
cys.pathway.relab <- pathway.relab.long %>%
  filter(Group != "Abx") %>%
  filter(PathwayID == "PWY-6292")

set.seed(42)
cys.pathway.plot <- (ggplot(cys.pathway.relab, aes(x = Group, y = relabundance, color = Group)) +
  # geom_boxplot(width = 0.1) +
  # geom_jitter(width = 0.1) +
  geom_boxplot(outlier.shape = 2, outlier.size = 0) +
  geom_jitter(color = "black", width = 0, height = 0) +
  ylab("Pathway copies per million (CPM)") +
  ggtitle("Distribution of relative abundance (CPM) values for PWY-6292",
    subtitle = "superpathway of L-cysteine biosynthesis (mammalian)"
  ) +
  scale_fill_manual(values = scater:::.get_palette("tableau10medium")) +
  scale_colour_manual(values = scater:::.get_palette("tableau10medium")) +
  theme_minimal() +
  theme(
    # panel.grid.major.x = element_blank(),
    panel.grid = element_blank(),
  )
)

cys.pathway.plot

ggsave("docs/plots/humann_boxplot_PWY-6292_het_homo_plot.png", width = 11, height = 10, dpi = 600)
ggsave("docs/plots/humann_boxplot_PWY-6292_het_homo_plot.pdf", device = "pdf")
write.csv(cys.pathway.relab, "docs/plots/humann_boxplot_PWY-6292_het_homo_plot.csv", row.names = FALSE)
```

- PDF version: [plots/humann_boxplot_PWY-6292_het_homo_plot.pdf](plots/humann_boxplot_PWY-6292_het_homo_plot.pdf)

```{r}
# We also generate a table ofthe CPM values for each of the significant pathways, per sample
sig_pathways <- maaslin2_results %>%
  select(-c(metadata, value, name, Pathway)) %>%
  filter(qval <= 0.05)

sig_pathway_cpms <- data.frame()

for (pathway_id in sig_pathways$PathwayID) {
  pathway_cpms <- pathway.relab.long %>%
    filter(Group != "Abx") %>%
    filter(PathwayID == pathway_id)

  sig_pathway_cpms <- rbind(sig_pathway_cpms, pathway_cpms)
}

# Convert sig_pathway_cpms to wide format
sig_pathway_cpms_wide <- sig_pathway_cpms %>%
  select(Pathway, Sample, relabundance) %>%
  pivot_wider(names_from = Sample, values_from = relabundance)

# Ensure Pathway is the first column
sig_pathway_cpms_wide <- sig_pathway_cpms_wide %>%
  select(Pathway, everything())


# Display the significant pathway cpms
# sig_pathway_cpms_wide %>% show_table_extra(escape = FALSE)

write.csv(sig_pathway_cpms_wide, "docs/plots/humann_sig_pathways_cpms_by_sample.csv", row.names = FALSE)
```

Which organism(s) is PWY-6292 the pathway most associated with in Het/Homo ?

```{r}
# show_table(pathway.relab.strat.long %>% filter(Group != "Abx") %>% filter((grepl("PWY-6292", Pathway))))
show_table(pathway.relab.strat %>% filter((grepl("PWY-6292", Pathway))), digits = 9)
```

**Lactobacillus reuteri**, and unclassified species.

All siginficant pathways relative abundance box plots:
```{r}
sig <- maaslin2_results %>%
  filter(qval <= 0.05) %>%
  arrange(-coef)

apply(sig, 1, function(r) {
  (ggplot(pathway.relab.long
  %>% filter(Group != "Abx")
    %>% filter(PathwayID == r["PathwayID"]), aes(x = Group, y = relabundance)) +
    geom_boxplot() +
    geom_jitter() +
    ylab("Pathway copies per million (CPM)") +
    ggtitle(f("Distribution of relative abundance values (CPM) for {r['PathwayID']}"),
      subtitle = f("{r['Pathway']} \nFDR={r['qval']}")
    ) +
    theme_bw()
  )
})
```

```{r, pathway_density_plot_all, fig.width=11, fig.height=10, out.width="100%", include=FALSE, eval=FALSE}
# Create the density plot
ggplot(
  pathway.relab.long %>% filter(PathwayID %in% sig$PathwayID),
  aes(x = relabundance, fill = Group, color = Group)
) +
  geom_density(alpha = 0.5, adjust = 1) +
  facet_wrap(~Pathway, scales = "free_y", ncol = 1) +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_fill_manual(values = scater:::.get_palette("tableau10medium")) +
  scale_colour_manual(values = scater:::.get_palette("tableau10medium")) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    strip.background = element_blank(),
    strip.text.x = element_text(size = 10),
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(
    x = "Relative abundance", y = "Density", fill = "Group",
    title = "Relative abundance density of top Pathways (MetaCyc)"
  ) +
  guides(fill = guide_legend(title = "Group"))

ggsave("docs/plots/pathway_density_plot.png", width = 11, height = 10, dpi = 300)

# - [plots/pathway_density_plot.png](plots/pathway_density_plot.png)
```


```{r pathway_density_plot_hethomo, fig.width=11, fig.height=10, out.width="100%"}
# Create the density plot
ggplot(
  pathway.relab.long %>% filter(PathwayID %in% sig$PathwayID) %>% filter(Group != "Abx"),
  aes(x = relabundance, fill = Group, color = Group)
) +
  geom_density(alpha = 0.5, adjust = 1) +
  # geom_jitter(alpha = 0.8, size = 1, width = 0.02) +
  geom_rug(sides = "b", size = 1, length = unit(0.1, "npc")) +
  facet_wrap(~Pathway, scales = "free_y", ncol = 1) +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_fill_manual(values = scater:::.get_palette("tableau10medium")) +
  scale_colour_manual(values = scater:::.get_palette("tableau10medium")) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    strip.background = element_blank(),
    strip.text.x = element_text(size = 10),
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(
    x = "Relative abundance", y = "Density", fill = "Group",
    title = "Relative abundance density of top Pathways (MetaCyc)"
  ) +
  guides(fill = guide_legend(title = "Group"))

ggsave("docs/plots/pathway_density_het_homo_plot.png", width = 11, height = 10, dpi = 600)
```

- [plots/pathway_density_het_homo_plot.png](plots/pathway_density_het_homo_plot.png)

```{r, include=FALSE, eval=FALSE}
# Alternative - faceted box plot
ggplot(
  pathway.relab.long %>% filter(PathwayID %in% sig$PathwayID) %>% filter(Group != "Abx"),
  aes(x = relabundance, color = Group)
) +
  # geom_density(alpha = 0.5, adjust=1) +
  geom_boxplot() +
  geom_rug(sides = "b", size = 1, length = unit(0.1, "npc")) +
  facet_wrap(~Pathway, scales = "free_y", ncol = 1) +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_fill_manual(values = scater:::.get_palette("tableau10medium")) +
  scale_colour_manual(values = scater:::.get_palette("tableau10medium")) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    strip.background = element_blank(),
    strip.text.x = element_text(size = 10),
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(
    x = "Relative abundance", y = "Density", fill = "Group",
    title = "Relative abundance density of top Pathways (MetaCyc)"
  ) +
  guides(fill = guide_legend(title = "Group"))

ggsave("docs/plots/pathway_density_het_homo_plot.png", width = 11, height = 10, dpi = 600)
```


A heatmap of only the siginficant differentially abundant pathways.
We show log2(cpm), where cpm is complete pathway copies per million.

```{r pathway_heatmap, out.width="100%"}
library(pheatmap)
library(viridis)
library(RColorBrewer)
# library(heatmaply)

pways <- (
  pathway.relab %>%
    mutate(PathwayID = str_extract(Pathway, "^(.*?)(?=:)")) %>%
    filter(PathwayID %in% sig$PathwayID) %>%
    filter((Pathway != "UNMAPPED") & Pathway != "UNINTEGRATED") %>%
    # simplify long descriptions
    mutate(
      Pathway = str_remove(Pathway, "superpathway of\\s*"),
      Pathway = str_remove(Pathway, "\\s*\\([^\\)]+\\)"),
      Pathway = str_replace(Pathway, "ation", "'n"),
    )
)

# pways <- move_column_to_index(pways, "PathwayID", 1)
# Drop the PathwayID column we don't need anymore
pways <- pways %>% select(-PathwayID)

# We remove the Abx condition since the pathways are driven almost entirely by Lactococcus specific metabolism
# This obscures interesting patterns between Het and Homo
pways <- pways %>% select(-starts_with("Abx"))

# pways <- pways %>% select(-starts_with("Homo-3"))

pways.mat <- as.matrix(pways[, -1])
rownames(pways.mat) <- pways$Pathway

# pways.var <- apply(pways.mat, 1, var) # Calculate variance for each row
# threshold <- quantile(pways.var, 0.75) # Set a threshold, (75th percentle)
# pways.mat <- pways.mat[pways.var > threshold, ]
# rownames(pways.mat) <- pways$Pathway[pways.var > threshold]

pathways.sig.heatmap <- pheatmap(log(pways.mat + 1),
  clustering_distance_cols = "euclidean",
  clustering_distance_rows = "euclidean",
  # kmeans_k=4,
  cutree_cols = 2,
  treeheight_row = 20, treeheight_col = 20,
  fontsize = 8,
  angle_col = 90,
  breaks = seq(0, 6.0, length.out = 101),
  legend_breaks = seq(0, 6.0, by = 1),
  legend_labels = sprintf("%.1f", seq(0, 6.0, by = 1))
)

pathways.sig.heatmap.v <- pheatmap(log(pways.mat + 1),
  clustering_distance_cols = "euclidean",
  clustering_distance_rows = "euclidean",
  color = viridis(n = 64, option = "viridis"),
  # color = viridis(n = 64, option = "magma"),
  cutree_cols = 2,
  # kmeans_k=4,
  cellwidth = 10, cellheight = 10,
  treeheight_row = 20, treeheight_col = 20,
  fontsize = 8,
  angle_col = 90,
  breaks = seq(0, 6.0, length.out = 64),
  legend_breaks = seq(0, 6.0, by = 1),
  legend_labels = sprintf("%.1f", seq(0, 6.0, by = 1))
)

pathways.sig.heatmap

ggsave("docs/plots/Part_F_da_pathways_log2cpm_heatmap.tiff", pathways.sig.heatmap,
  device = "tiff", compression = "lzw", units = "mm", width = 180, height = 172.5, dpi = 600
)

ggsave("docs/plots/Part_F_da_pathways_log2cpm_heatmap.v.tiff", pathways.sig.heatmap.v,
  device = "tiff", compression = "lzw", units = "mm", width = 180, height = 172.5, dpi = 600
)

ggsave("docs/plots/Part_F_da_pathways_log2cpm_heatmap.svg", pathways.sig.heatmap,
  device = "svg", units = "mm", width = 180, height = 172.5, dpi = 600
)

ggsave("docs/plots/Part_F_da_pathways_log2cpm_heatmap.v.svg", pathways.sig.heatmap.v,
  device = "svg", units = "mm", width = 180, height = 172.5, dpi = 600
)

ggsave("docs/plots/Part_F_da_pathways_log2cpm_heatmap.png", pathways.sig.heatmap,
  device = "png", units = "mm", width = 180, height = 172.5, dpi = 600
)

ggsave("docs/plots/Part_F_da_pathways_log2cpm_heatmap.v.png", pathways.sig.heatmap.v,
  device = "png", units = "mm", width = 180, height = 172.5, dpi = 600
)

# hm <- heatmaply(
#   pways.mat,
#   k_col = 2, k_row = 2,
#   labels_row = rownames(pways.mat),
#   labRow = rownames(pways.mat),
#   labCol = colnames(pways.mat),
#   fontsize_row = 8
# )

# htmlwidgets::saveWidget(hm, file.path(base_path, "workflowr/docs/plots/da_pathways_heatmap.html"))
```


```{r genes_heatmap, eval=FALSE, include=FALSE}
# ## Gene families

# ### Gene family (UniRef90) heatmap

# A hierarchical clustered heatmap of using the abundance of UniRef90 protein families for each sample.
# To simplify plot we show only gene families with high variance between samples (greater than 99th percentile).

# - [plots/gene_family_heatmap.html](plots/gene_family_heatmap.html)


# library(pheatmap)
library(heatmaply)

genes <- gene.relab %>% filter((Gene_Family != "UNMAPPED") & Gene_Family != "UNINTEGRATED")

# We remove the Abx condition since the pathways are driven almost entirely by Lactococcus specific metabolism
# This obscures interesting patterns between Het and Homo
genes <- genes %>% select(-starts_with("Abx"))

# genes <- genes %>% select(-starts_with("Homo-3"))

genes.mat <- as.matrix(genes[, -1])
rownames(genes.mat) <- genes$Gene_Family

genes.var <- apply(genes.mat, 1, var) # Calculate variance for each row
threshold <- quantile(genes.var, 0.99) # Set a threshold, (95th percentle)
genes.mat <- genes.mat[genes.var > threshold, ]

rownames(genes.mat) <- genes$Gene_Family[genes.var > threshold]

hm <- heatmaply(
  genes.mat,
  k_col = 2, k_row = 2,
  labels_row = rownames(genes.mat),
  labRow = rownames(genes.mat),
  labCol = colnames(genes.mat),
  fontsize_row = 6
)

htmlwidgets::saveWidget(hm, file.path(base_path, "workflowr/docs/plots/gene_family_heatmap.html"))
```


```{r}
htmltools::includeHTML("analysis/acknowledging.html")
```
