---
title: "Home"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

```{r packages, include=FALSE}
# install.packages("renv")

# renv::install(c("languageserver")) # for vscode

# renv::install(c("tidyverse", "ggplot2", "DT", "kableExtra", "workflowr", "ggsignif", "RColorBrewer", "httpgd", "bioc::phyloseq", "bioc::mia", "bioc::miaViz"), prompt=FALSE)

# renv::install("FrederickHuangLin/ANCOMBC") # github
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

source("analysis/rmarkdown_utils.R")
source("analysis/tse_utils.R")

library(tidyverse)
library(mia)
# library(miaTime)
library(miaViz)
library(ggplot2)
library(BiocManager)
library(Biostrings)

f <- str_glue
```

```{r report_template}
htmltools::includeHTML("analysis/index_for_workflowr.html")
```

# Cystinuria Mouse Gut Microbiome
## Nirmal Bhatt, Malcolm Starkey

**Author:** Andrew Perry  
**BIP:** BIP-1131
**iLab:** MBP-MS-466
**starting date:** 14-Aug-2023   

---

### Summary
 
**Goal:** To explore the differences between faecal microbiomes of the homozygous cystinuria mice with heterozygotes.

**Data:** Mouse faecal microbiome, short read shotgun WGS.

Three conditions with 5 samples each:

- `Abx` (antibiotic treated)
- `Het` (heterozygous)
- `Homo` (homozygous)

Potential covariates:

- all mice are male (so no sex covariate)
- box locations (some mice were housed together in the same box)

```{r metadata_table}
metadata <- read.csv('data/metadata.csv')
show_table(metadata)
```

The cystinuria includes formation of cysteine stones in the predominantly in the kidney.
Stone count, mass (mg) and total stone surface area (mm^2) was also provided.


### Pipeline overview

Host (Mus musculus) and PhiX reads were depleted from quality-filtered paired-end reads using Bowtie2.
Taxonomic assignments of the remaining reads were made using MetaPhlAn4 against the ChocoPhlAn database (vOct22).

### Analysis

Data was analysed in R, primarily using the `mia` and `miaViz` Bioconductor pacakges.
Taxonomic abundance was assessed using MetaPhlAn4, with differential abundance testing via ANCOM-BC2.
Functional and pathway enrichment was assessed using HUMAnN, with differential pathway abundance testing via MaAsLin2.

- [Raw read QC](multiqc/run1/multiqc_report.html)
  - Raw sequencing data is of good quality. ~50-80M 100bp paired reads per sample. Most samples have ~3-5% mouse host reads (removed).
- [MetaPhlAn4 taxonomy and diversity](metaphlan.html)
- [HUMAnN functional and pathway enrichment](humann.html)
- [Summary notes & proposed manuscript figures](summary_notes.html)

Higher resolution versions of some key plots can be found in: [plots/](plots/)

```{markdown include=FALSE, eval=FALSE}
[This paper](https://www.nature.com/articles/s41591-019-0504-5), 16S study, 
found _Akkermansia muciniphila_ beneficial in mice, potentially helping 
modulate bile acids.  Uses the LEfSE linear discriminant analyst method for DA. 

- **TODO**: HUMAnN ?

```

```{r}
htmltools::includeHTML("analysis/acknowledging.html")
```


```{r include=FALSE, eval=FALSE}
# Never uncomment - infinite loop !
# rmarkdown::render("workflowr/analysis/index.Rmd", knit_root_dir="/home/andrewperry/projects/nirmal.bhatt/cystinuria_Mouse_microbiome_WGS/")

# or
# workflowr::wflow_build()
```

