---
title: "Methaphlan taxonomic abundance analysis"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

```{r setwd, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

if (Sys.getenv("TERM_PROGRAM") == "vscode") {
  try(
    {
      .vsc.attach()
    },
    silent = TRUE
  )
  try(
    {
      setwd("workflowr")
    },
    silent = TRUE
  )
}
```

```{r report_template}
htmltools::includeHTML("analysis/index_for_workflowr.html")
```

```{r data_import}
library(tidyverse)
library(purrr)
library(mia)
library(miaViz)
library(scater)
library(dplyr)
library(viridis)

f <- stringr::str_glue

source("analysis/rmarkdown_utils.R")
source("analysis/nfcore_mag_utils.R")
source("analysis/tse_utils.R")
```

MetaPhlAn uses (trimmed, host-filtered) reads aligned to a reference database of marker genes 
to assign taxonomy and estimate relative abundance of taxa.
Taxonomic labels were converted to the GTDB r202 taxonomy to allow comparison with other analyses.

(MetaPhlAn version 4.0.6, vOct22_CHOCOPhlAnSGB_202212 database).

```{r load_data, include=FALSE, eval=FALSE}
# # Loading merged tables, from `merge_metaphlan_tables.py --gtdb_profiles`
# # https://github.com/biobakery/MetaPhlAn/wiki/MetaPhlAn-4#merging-tables

# files <- Sys.glob("../pipeline/metaphlan/results-gtdb-nounclassified/*_metaphlan-gtdb.txt")
# tses <- c()
# for (fn in files) {
#   # print(fn)
#   sample_name = gsub("_metaphlan-gtdb\\.txt$", "", basename(fn))
#   tses <- c(tses, loadFromMetaphlanSingle(fn, sample_name, sep=";", skip=1))
# }

# # Merge single sample TreeSummarizedExperiments into a single one, all assays
# tse <- mergeSEs(tses, assay.type = assayNames(tses[[1]]))

# # Change NAs to zeros in assays
# for (a in assayNames(tse)) {
#   assay_data <- assay(tse, a)
#   assay_data[is.na(assay_data)] <- 0
#   assay(tse, a) <- assay_data
# }

# # tse <- loadFromMetaphlan4GTDB(
# #     "../pipeline/metaphlan/results/merged_abundance_table-gtdb.txt",
# #     #sep=";",
# #     column_suffix = "_metaphlan-gtdb"
# # )

# # Add the Abx,Het,Homo group names to the TreeSummarizedExperiment
# group <- gsub("-[1-9]", "", colnames(tse))
# colData(tse)$group <- group
```

```{r import_merged_data}
df <- readr::read_tsv("../pipeline/metaphlan/results-gtdb-nounclassified/merged-gtdb.tsv")

counts <- df %>%
  dplyr::select(c(ends_with("_counts"))) %>%
  dplyr::rename_with(~ gsub("_counts$", "", .), ends_with("_counts"))

counts <- as.matrix(counts)
rownames(counts) <- make.unique(df$clade_name)

relab <- df %>%
  dplyr::select(c(ends_with("_relative_abundance"))) %>%
  dplyr::rename_with(~ gsub("_relative_abundance$", "", .), ends_with("_relative_abundance"))

relab <- as.matrix(relab)
rownames(relab) <- make.unique(df$clade_name)

coverage <- df %>%
  dplyr::select(c(ends_with("_coverage"))) %>%
  dplyr::rename_with(~ gsub("_coverage$", "", .), ends_with("_coverage"))

coverage <- as.matrix(coverage)
rownames(coverage) <- make.unique(df$clade_name)

taxa_labels <- df %>% dplyr::select(
  clade_name,
  clade_taxid
)
# Split long d_Bacteria;p_Proteo .. etc strings into individual columns
taxa_labels <- taxa_labels %>%
  rowwise() %>%
  do(bind_cols(., split_taxonomy_str(.$clade_name)))

sample_metadata <- data.frame(
  sample = colnames(counts),
  group = gsub("-[0-9]+", "", colnames(counts))
) %>% mutate(group = stringr::str_to_title(group))
rownames(sample_metadata) <- sample_metadata$sample
sample_metadata$group <- as.factor(sample_metadata$group)

tse <- TreeSummarizedExperiment(
  assays = SimpleList(
    counts = counts
    # relab.metaphlan = relab,
    # coverage = coverage
  ),
  colData = sample_metadata,
  rowData = taxa_labels
)

# Drop Homo-3 ?
# tse <- tse[, !(tse$sample %in% c("Homo-3"))]

# Compute relabundance
tse <- transformCounts(tse,
  assay.type = "counts",
  method = "relabundance",
  # pseudocount = 1
)
```

```{r raw_table}
(
  df %>%
    mutate(mean_abundance = rowMeans(select(., ends_with("_relative_abundance")), na.rm = TRUE)) %>%
    arrange(desc(mean_abundance)) %>%
    show_table(pageLength = 20, escape = FALSE, rownames = FALSE)
)
```

```{r settings}
taxa.level <- "Species"
taxa.level.short <- "S"

# This sets the feature names to the species
# Mostly used for plotAbundanceDensity labels
# features <- rownames(tse)  # save original rownames
# rownames(tse) <- rowData(tse)[[taxa.level]]

# We can recover originals like:
# rownames(tse) <- features
```

## Taxonomy relative abundance

Barplot at `r taxa.level` level:

```{r taxa_barplot}
source("analysis/tse_utils.R")

height <- 172.5 * 0.75 * 0.7
height <- height * 0.95 # scale down by 5% after removing group legend

barplot <- (
  taxa_barplot(tse, taxa.level,
    n_top = 20,
    show_group_legend = FALSE
  )
  # uncomment if show_group_legend = TRUE
  #+ theme(legend.position = 'right')
  #+ theme(legend.position = c(1.2, 2))
)

barplot

ggsave("docs/plots/Part_A_composition_barplot.tiff", barplot,
  device = "tiff", compression = "lzw", units = "mm", width = 180,
  height = height, dpi = 600
)

ggsave("docs/plots/Part_A_composition_barplot.png", barplot,
  device = "png", units = "mm", width = 180, height = height, dpi = 600
)

ggsave("docs/plots/Part_A_composition_barplot.svg", barplot,
  device = "svg", units = "mm", width = 180, height = height, dpi = 600
)
```

Barplot, Abx/Homo and Het/Homo at `r taxa.level` level:

```{r taxa_barplot_hethomo}
barplot.hethomo <- (
  taxa_barplot(dropCondition(tse, "group", "Abx"), taxa.level,
    n_top = 20,
    show_group_legend = FALSE
  )
  # uncomment if show_group_legend = TRUE
  #+ theme(legend.position = 'right')
  #+ theme(legend.position = c(1.2, 2))
)

barplot.hethomo

ggsave("docs/plots/Part_Asplit_composition_barplot.hethomo.tiff", barplot.hethomo,
  device = "tiff", compression = "lzw", units = "mm", width = 130, height = height, dpi = 600
)

ggsave("docs/plots/Part_Asplit_composition_barplot.hethomo.png", barplot.hethomo,
  device = "png", units = "mm", width = 130, height = height, dpi = 600
)

ggsave("docs/plots/Part_Asplit_composition_barplot.hethomo.svg", barplot.hethomo,
  device = "svg", units = "mm", width = 130, height = height, dpi = 600
)
```

```{r taxa_barplot_abxhomo}
tse.nohet <- dropCondition(tse, "group", "Het")
colData(tse.nohet)$group <- factor(colData(tse.nohet)$group,
  levels = c("Homo", "Abx")
)

barplot.abxhomo <- (
  taxa_barplot(tse.nohet, taxa.level,
    n_top = 20,
    show_group_legend = FALSE
  )
  # uncomment if show_group_legend = TRUE
  #+ theme(legend.position = 'right')
  #+ theme(legend.position = c(1.2, 2))
)

barplot.abxhomo

ggsave("docs/plots/Part_Asplit_composition_barplot.abxhomo.tiff", barplot.abxhomo,
  device = "tiff", compression = "lzw", units = "mm", width = 130, height = height, dpi = 600
)

ggsave("docs/plots/Part_Asplit_composition_barplot.abxhomo.png", barplot.abxhomo,
  device = "png", units = "mm", width = 130, height = height, dpi = 600
)

ggsave("docs/plots/Part_Asplit_composition_barplot.abxhomo.svg", barplot.abxhomo,
  device = "svg", units = "mm", width = 130, height = height, dpi = 600
)
```


Distributions of taxonomic abundances for top taxa, by `r taxa.level`:

```{r abundance_scatter, include=FALSE, eval=FALSE}
# tse_glom <- mia::agglomerateByRank(tse, rank = taxa.level, onRankOnly = TRUE)

# plotAbundanceDensity(tse_glom,
#   layout = "jitter", assay.type = "relabundance",
#   n = 20, point_size = 1, point_shape = 19, point_alpha = 0.5,
#   colour_by = "group"
# ) +
#   scale_x_log10(label = scales::percent) +
#   ggtitle(f("Relative abundance of top {taxa.level}"))
```

Top 20 `r taxa.level` relatative abundance distributions:
```{r abundance_density}
tse_glom <- mia::agglomerateByRank(tse, rank = taxa.level, onRankOnly = TRUE)

# disable scientific notation (for the x-axis)
scipen.save <- options()$scipen
options(scipen = 999)

density.plot <- plotAbundanceDensity(tse_glom,
  layout = "density", assay.type = "relabundance",
  n = 20, colour_by = "group", point_alpha = 1 / 10,
) +
  # geom_point(aes(y='relabundance', color='group'), size=1, alpha=0.5) +
  scale_x_log10() +
  ggtitle(f("Relative abundance density of top {taxa.level}")) +
  theme(
    text = element_text(size = 8),
    axis.title = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    strip.text = element_text(size = 8),
    legend.text = element_text(size = 8),
  )

options(scipen = scipen.save)

density.plot

ggsave("docs/plots/Part_B_top_taxa_density_plot.tiff", density.plot,
  device = "tiff", compression = "lzw", units = "mm", width = 150, height = 150, dpi = 600
)

ggsave("docs/plots/Part_B_top_taxa_density_plot.png", density.plot,
  device = "png", units = "mm", width = 150, height = 150, dpi = 600
)

ggsave("docs/plots/Part_B_top_taxa_density_plot.svg", density.plot,
  device = "svg", units = "mm", width = 150, height = 150, dpi = 600
)
```

Top 20 `r taxa.level` relatative abundance distributions, `Het`/`Homo` and `Abx`/`Homo` only:
```{r mia_relabundance_density_hethomo}
# disable scientific notation (for the x-axis)
scipen.save <- options()$scipen
options(scipen = 999)

density.plot <- plotAbundanceDensity(dropCondition(tse_glom, "group", "Abx"),
  layout = "density", assay.type = "relabundance",
  n = 20, colour_by = "group", point_alpha = 1 / 10,
) +
  # geom_point(aes(y='relabundance', color='group'), size=1, alpha=0.5) +
  scale_x_log10() +
  ggtitle(f("Relative abundance density of top {taxa.level}")) +
  theme(
    text = element_text(size = 8),
    axis.title = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    strip.text = element_text(size = 8),
    legend.text = element_text(size = 8),
    plot.title = element_text(size = 8, hjust = 0.5)
  )

options(scipen = scipen.save)

w <- 120 # 150
h <- 100 # 150
ggsave("docs/plots/Part_Bsplit_top_taxa_density_plot.hethomo.tiff", density.plot,
  device = "tiff", compression = "lzw", units = "mm", width = w, height = h, dpi = 600
)

ggsave("docs/plots/Part_Bsplit_top_taxa_density_plot.hethomo.png", density.plot,
  device = "png", units = "mm", width = w, height = h, dpi = 600
)

ggsave("docs/plots/Part_Bsplit_top_taxa_density_plot.hethomo.svg", density.plot,
  device = "svg", units = "mm", width = w, height = h, dpi = 600
)
```

```{r mia_relabundance_density_abxhomo}
# disable scientific notation (for the x-axis)
scipen.save <- options()$scipen
options(scipen = 999)

density.plot <- plotAbundanceDensity(dropCondition(tse_glom, "group", "Het"),
  layout = "density", assay.type = "relabundance",
  n = 20, colour_by = "group", point_alpha = 1 / 10,
) +
  # geom_point(aes(y='relabundance', color='group'), size=1, alpha=0.5) +
  scale_x_log10() +
  ggtitle(f("Relative abundance density of top {taxa.level}")) +
  theme(
    text = element_text(size = 8),
    axis.title = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    strip.text = element_text(size = 8),
    legend.text = element_text(size = 8),
    plot.title = element_text(size = 8, hjust = 0.5)
  )

options(scipen = scipen.save)

w <- 120 # 150
h <- 100 # 150
ggsave("docs/plots/Part_Bsplit_top_taxa_density_plot.abxhomo.tiff", density.plot,
  device = "tiff", compression = "lzw", units = "mm", width = w, height = h, dpi = 600
)

ggsave("docs/plots/Part_Bsplit_top_taxa_density_plot.abxhomo.png", density.plot,
  device = "png", units = "mm", width = w, height = h, dpi = 600
)

ggsave("docs/plots/Part_Bsplit_top_taxa_density_plot.abxhomo.svg", density.plot,
  device = "svg", units = "mm", width = w, height = h, dpi = 600
)
```


Heatmap with hierarchical clustering based on 20 most abundanant taxa (all samples, and Het/Homo only):
Color intensity indicates the Z-score of the CLR (centered-log ratio) relative abundance.

```{r top_taxa_clr_z_heatmap}
# https://microbiome.github.io/OMA/microbiome-community.html#composition-heatmap
library(mia)
library(miaViz)
library(pheatmap)

# Take only Het/Homo - we do this FIRST so the CLR transformation is based on only Het/Homo
tse_glom.hethomo <- tse_glom[, tse_glom$group %in% c("Het", "Homo")]

# Add clr-transformation
tse_glom.hethomo <- mia::transformCounts(tse_glom.hethomo,
  # assay.type = "counts",
  # assay.type = "depth",
  assay.type = "relabundance",
  method = "clr",
  pseudocount = TRUE
)

tse_glom.hethomo <- mia::transformCounts(tse_glom.hethomo,
  assay.type = "clr",
  MARGIN = "features",
  method = "z", name = "clr_z"
)

# Get n most abundant taxa, and subsets the data by them
top_taxa <- getTopFeatures(
  tse_glom.hethomo,
  top = min(nrow(rowData(tse_glom)), 20),
  assay.type = "relabundance"
)

tse_glom_top <- tse_glom.hethomo[top_taxa, ]

# Get the assay table
mat <- assay(tse_glom_top, "clr_z")

# Het/Homo only
mat <- assay(tse_glom_top[, tse_glom_top$group %in% c("Het", "Homo")], "clr_z")
# Drop homo_3
# mat <- assay(tse_glom_top[, tse_glom_top$group %in% c('Het', 'Homo') & (tse_glom_top$sample != "Homo-3")], "clr_z")

set.seed(42)
# Create the heatmap
taxa.heatmap.top.clr_z <- pheatmap(mat,
  clustering_distance_cols = "euclidean",
  cluster_cols = FALSE,
  clustering_distance_rows = "euclidean",
  # kmeans_k=4,
  # main="Hierarchical clustering of top 20 species (z-transformed CLR)",
  # cutree_cols = 2,
  gaps_col = 5,
  legend = TRUE,
  cellwidth = 10, cellheight = 10,
  treeheight_row = 20, treeheight_col = 20,
  fontsize = 8,
  angle_col = 90,
  breaks = seq(-2.5, 2.5, length.out = 101),
  legend_breaks = seq(-2.5, 2.5, by = 0.5),
  legend_labels = sprintf("%.1f", seq(-2.5, 2.5, by = 0.5))
)

# Viridis version
taxa.heatmap.top.clr_z.v <- pheatmap(mat,
  clustering_distance_cols = "euclidean",
  cluster_cols = FALSE,
  clustering_distance_rows = "euclidean",
  # kmeans_k=4,
  # main="Hierarchical clustering of top 20 species (z-transformed CLR)",
  # cutree_cols = 2,
  gaps_col = 5,
  legend = TRUE,
  cellwidth = 10, cellheight = 10,
  fontsize = 8,
  treeheight_row = 20, treeheight_col = 20,
  angle_col = 90,
  color = viridis(n = 64, option = "viridis"),
  breaks = seq(-2.5, 2.5, length.out = 64),
  legend_breaks = seq(-2.5, 2.5, by = 0.5),
  legend_labels = sprintf("%.1f", seq(-2.5, 2.5, by = 0.5))
)

taxa.heatmap.top.clr_z

ggsave("docs/plots/Part_D_top_taxa_clr_z_heatmap.tiff", taxa.heatmap.top.clr_z,
  device = "tiff", compression = "lzw", units = "mm", width = 180 * 0.75, height = 172.5 * 0.75, dpi = 600
)

ggsave("docs/plots/Part_D_top_taxa_clr_z_heatmap.png", taxa.heatmap.top.clr_z,
  device = "png", units = "mm", width = 180 * 0.75, height = 172.5 * 0.75, dpi = 600
)

ggsave("docs/plots/Part_D_top_taxa_clr_z_heatmap.svg", taxa.heatmap.top.clr_z,
  device = "svg", units = "mm", width = 180 * 0.75, height = 172.5 * 0.75, dpi = 600
)

ggsave("docs/plots/Part_D_top_taxa_clr_z_heatmap.v.tiff", taxa.heatmap.top.clr_z.v,
  device = "tiff", compression = "lzw", units = "mm", width = 180 * 0.75, height = 172.5 * 0.75, dpi = 600
)

ggsave("docs/plots/Part_D_top_taxa_clr_z_heatmap.v.png", taxa.heatmap.top.clr_z.v,
  device = "png", units = "mm", width = 180 * 0.75, height = 172.5 * 0.75, dpi = 600
)

ggsave("docs/plots/Part_D_top_taxa_clr_z_heatmap.v.svg", taxa.heatmap.top.clr_z.v,
  device = "svg", units = "mm", width = 180 * 0.75, height = 172.5 * 0.75, dpi = 600
)
```


```{r filtered_counts}
# Make a new SummarizedExperiment where taxa must have non-zero counts for half the samples

counts.glom <- assay(tse_glom[, tse_glom_top$group %in% c("Het", "Homo")], "counts")
keep <- counts.glom >= 10
features_to_keep <- rowSums(keep) >= (ncol(counts.glom) / 2)
tse_glom.filtered <- tse_glom[features_to_keep, ]
```

Heatmap with hierarchical clustering based all taxa with at least 10 counts in half the samples (Het/Homo only):
Color intensity indicates the Z-score of the CLR (centered-log ratio) relative abundance.

```{r all_taxa_clrz_heatmap}
# https://microbiome.github.io/OMA/microbiome-community.html#composition-heatmap
library(mia)
library(miaViz)
library(pheatmap)

# Take only Het/Homo
# tse_glom <- tse_glom[, tse_glom$group %in% c("Het", "Homo")]

# Add clr-transformation
tse_glom.filtered <- mia::transformCounts(tse_glom.filtered,
  # assay.type = "counts",
  # assay.type = "depth",
  assay.type = "relabundance",
  method = "clr",
  pseudocount = TRUE
)

tse_glom.filtered <- mia::transformCounts(tse_glom.filtered,
  assay.type = "clr",
  MARGIN = "features",
  method = "z", name = "clr_z"
)

# Get n most abundant taxa, and subsets the data by them
top_taxa <- getTopFeatures(
  tse_glom.filtered,
  top = min(nrow(rowData(tse_glom.filtered)), 10000),
  assay.type = "relabundance"
)

tse_glom_top <- tse_glom.filtered[top_taxa, ]

# Get the assay table
mat <- assay(tse_glom_top, "clr_z")

# Het/Homo only
mat <- assay(tse_glom_top[, tse_glom_top$group %in% c("Het", "Homo")], "clr_z")
# Drop homo_3
# mat <- assay(tse_glom_top[, tse_glom_top$group %in% c('Het', 'Homo') & (tse_glom_top$sample != "Homo-3")], "clr_z")

# Create the heatmap
pheatmap(mat,
  clustering_distance_cols = "euclidean",
  clustering_distance_rows = "euclidean",
  # kmeans_k=4,
  main = "Hierarchical clustering (z-transformed CLR)",
  legend = TRUE,
  cellwidth = 12, cellheight = 12,
  fontsize = 8
)
```

Heatmap with hierarchical clustering using log10-relative abundance based the top 20 taxa (all samples, and Het/Homo only):
Color intensity indicates the log10 of relative abundance.

```{r top_taxa_abundance_heatmap_log}
logstyle <- "log10"

tse_glom.hethomo <- mia::transformCounts(tse_glom.hethomo,
  assay.type = "relabundance",
  method = logstyle,
  pseudocount = TRUE,
  name = f("{logstyle}_relabundance"),
)

top_taxa <- getTopFeatures(
  tse_glom.hethomo,
  top = min(nrow(rowData(tse_glom)), 20),
  method = "mean",
  assay.type = "relabundance"
)

tse_glom_top <- tse_glom.hethomo[top_taxa, ]

mat <- assay(tse_glom_top, f("{logstyle}_relabundance"))

# Het/Homo only
mat <- assay(tse_glom_top[, tse_glom_top$group %in% c("Het", "Homo")], f("{logstyle}_relabundance"))
# Drop homo_3
# mat <- assay(tse_glom[, tse_glom$group %in% c('Het', 'Homo') & (tse_glom$sample != "homo_3")], "clr_z")

# Create the heatmap
taxa.heatmap.log10relab <- pheatmap(mat,
  cluster_cols = FALSE,
  clustering_distance_cols = "euclidean",
  clustering_distance_rows = "euclidean",
  # kmeans_k=4,
  # main="Hierarchical clustering log10 relative abundance (top 20 taxa)",
  cellwidth = 10, cellheight = 10,
  treeheight_row = 20, treeheight_col = 20,
  fontsize = 8,
  angle_col = 90,
  breaks = seq(-5, 0, length.out = 101),
)

taxa.heatmap.log10relab.v <- pheatmap(mat,
  cluster_cols = FALSE,
  clustering_distance_cols = "euclidean",
  clustering_distance_rows = "euclidean",
  # kmeans_k=4,
  # main="Hierarchical clustering log10 relative abundance (top 20 taxa)",
  cellwidth = 10, cellheight = 10,
  treeheight_row = 20, treeheight_col = 20,
  fontsize = 8,
  angle_col = 90,
  color = viridis(n = 64, option = "viridis"),
  breaks = seq(-5, 0, length.out = 64),
)

taxa.heatmap.log10relab

ggsave("docs/plots/Part_Dalt_top_taxa_log_relab_heatmap.tiff", taxa.heatmap.log10relab,
  device = "tiff", compression = "lzw", units = "mm", width = 180 * 0.75, height = 172.5 * 0.75, dpi = 600
)

ggsave("docs/plots/Part_Dalt_top_taxa_log_relab_heatmap.png", taxa.heatmap.log10relab,
  device = "png", units = "mm", width = 180 * 0.75, height = 172.5 * 0.75, dpi = 600
)

ggsave("docs/plots/Part_Dalt_top_taxa_log_relab_heatmap.v.tiff", taxa.heatmap.log10relab.v,
  device = "tiff", compression = "lzw", units = "mm", width = 180 * 0.75, height = 172.5 * 0.75, dpi = 600
)

ggsave("docs/plots/Part_Dalt_top_taxa_log_relab_heatmap.v.png", taxa.heatmap.log10relab.v,
  device = "png", units = "mm", width = 180 * 0.75, height = 172.5 * 0.75, dpi = 600
)
```

Heatmap with hierarchical clustering using log10-relative abundance based all taxa with at least 10 counts in half the samples (all samples, and Het/Homo only):
Color intensity indicates the log10 of relative abundance.

```{r all_taxa_abundance_heatmap_log}
logstyle <- "log10"

tse_glom.filtered <- mia::transformCounts(tse_glom.filtered,
  assay.type = "relabundance",
  method = logstyle,
  pseudocount = TRUE,
  name = f("{logstyle}_relabundance"),
)

top_taxa <- getTopFeatures(
  tse_glom.filtered,
  top = min(nrow(rowData(tse_glom.filtered)), 10000),
  method = "mean",
  assay.type = "relabundance"
)

tse_glom_top <- tse_glom.filtered[top_taxa, ]

mat <- assay(tse_glom_top, f("{logstyle}_relabundance"))

# Het/Homo only
mat <- assay(tse_glom_top[, tse_glom_top$group %in% c("Het", "Homo")], f("{logstyle}_relabundance"))
# Drop homo_3
# mat <- assay(tse_glom[, tse_glom$group %in% c('Het', 'Homo') & (tse_glom$sample != "homo_3")], "clr_z")

# Create the heatmap
taxa.heatmap.log10relab <- pheatmap(mat,
  cluster_cols = FALSE,
  clustering_distance_cols = "euclidean",
  clustering_distance_rows = "euclidean",
  # kmeans_k=4,
  cellwidth = 10, cellheight = 10,
  fontsize = 8,
  cutree_cols = 2,
)

taxa.heatmap.log10relab.v <- pheatmap(mat,
  cluster_cols = FALSE,
  clustering_distance_cols = "euclidean",
  clustering_distance_rows = "euclidean",
  # kmeans_k=4,
  cellwidth = 10, cellheight = 10,
  fontsize = 8,
  color = viridis(n = 64, option = "viridis"),
  cutree_cols = 2,
)

taxa.heatmap.log10relab

ggsave("docs/plots/Part_D_taxa_log_relab_heatmap.tiff", taxa.heatmap.log10relab,
  device = "tiff", compression = "lzw", units = "mm", dpi = 600, height = 600
) # , width=180, height=172.5)

ggsave("docs/plots/Part_D_taxa_log_relab_heatmap.png", taxa.heatmap.log10relab,
  device = "png", units = "mm", dpi = 600, height = 600
)

ggsave("docs/plots/Part_D_taxa_log_relab_heatmap.v.tiff", taxa.heatmap.log10relab.v,
  device = "tiff", compression = "lzw", units = "mm", dpi = 600, height = 600
) # , width=180, height=172.5)

ggsave("docs/plots/Part_D_taxa_log_relab_heatmap.v.png", taxa.heatmap.log10relab.v,
  device = "png", units = "mm", dpi = 600, height = 600
)
```

Heatmap with hierarchical clustering using unnormalized raw counts based all taxa with at least 10 counts in half the samples (all samples, and Het/Homo only):
Color intensity indicates the counts.

```{r heatmap_counts, include=FALSE, eval=FALSE}
# top_taxa <- getTopFeatures(
#   tse_glom,
#   top = min(nrow(rowData(tse_glom)), 10000),
#   method = "mean",
#   assay.type = "counts")

tse_glom_top <- tse_glom.filtered # tse_glom[top_taxa, ]

mat <- assay(tse_glom_top, f("counts"))

# Het/Homo only
mat <- assay(tse_glom_top[, tse_glom_top$group %in% c("Het", "Homo")], f("counts"))
# Drop homo_3
# mat <- assay(tse_glom[, tse_glom$group %in% c('Het', 'Homo') & (tse_glom$sample != "homo_3")], "clr_z")

# Create the heatmap
pheatmap(mat,
  clustering_distance_cols = "euclidean",
  clustering_distance_rows = "euclidean",
  # kmeans_k=4,
  cellwidth = 12, cellheight = 12,
  fontsize = 8
)
```

Heatmap with hierarchical clustering based on rank of relative abundance based all taxa with at least 10 counts in half the samples (Het/Homo only):
Color intensity indicates the rank of the relative abundance for this species.

```{r all_taxa_heatmap_rank}
tse_glom.filtered <- mia::transformCounts(tse_glom,
  assay.type = "relabundance",
  method = "rank",
  pseudocount = TRUE,
  name = f("rank_relabundance"),
)

top_taxa <- getTopFeatures(
  tse_glom,
  top = min(nrow(rowData(tse_glom.filtered)), 10000),
  method = "mean",
  assay.type = "relabundance"
)

tse_glom_top <- tse_glom.filtered[top_taxa, ]

mat <- assay(tse_glom_top, f("rank_relabundance"))

# Het/Homo only
mat <- assay(tse_glom_top[, tse_glom_top$group %in% c("Het", "Homo")], f("rank_relabundance"))
# Drop homo_3
# mat <- assay(tse_glom[, tse_glom$group %in% c('Het', 'Homo') & (tse_glom$sample != "homo_3")], "clr_z")

# Create the heatmap
pheatmap(mat,
  clustering_distance_cols = "euclidean",
  clustering_distance_rows = "euclidean",
  # kmeans_k=4,
  cellwidth = 12, cellheight = 12,
  fontsize = 8
)
```

### Alpha diversity and richness measure
```{r alpha_diversity_observed}
tse <- mia::estimateRichness(tse_glom,
  assay.type = "counts",
  index = "observed",
  name = "observed"
)

plotColData(tse,
  "observed",
  "group",
  colour_by = "group"
) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle(f("Observed features alpha-diversity ({taxa.level})")) +
  ylab(expression(Richness[Observed]))
```

```{r}
metaphaln.alpha.diversity.df <- as.data.frame(colData(tse))
write.csv(metaphaln.alpha.diversity.df, "docs/plots/metaphaln_alpha_diversity_observed.csv", row.names = FALSE)
```

```{r alpha_diversity_chao1, include=FALSE, eval=FALSE}
# `Chao1` is a measure of richness that estimates the number of species (it gives more weight to low abundance species).
# Since `Chao1` is just an additive term to total observed features, and depends on singletons (taxa count = 1) and doubletons (taxa count = 2),
# in this case `Chao1` is identical to the number observed features (we have no singletons).

tse <- mia::estimateRichness(tse_glom,
  assay.type = "counts",
  index = "chao1",
  name = "chao1"
)

plotColData(tse,
  "chao1",
  "group",
  colour_by = "group"
) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle(f("Chao1 alpha-diversity ({taxa.level})")) +
  ylab(expression(Richness[Chao1]))
```

An another measure of diversity within each sample is the Shannon Index, plotted below. 
We can test if the Shannon Index is significantly different between groups.

```{r alpha_diversity_tests}
library(ggsignif)
library(ggplot2)
library(patchwork)

tse <- mia::estimateDiversity(tse_glom,
  assay.type = "counts",
  index = "shannon",
  name = "shannon"
)

# Subsets the data (if we only want to compare Homo and Het)
df <- as.data.frame(colData(tse)[tse$group %in%
  c("Abx", "Homo", "Het"), ])

df$group <- factor(df$group)

# For significance testing, all different combinations are determined
comb <- split(
  t(combn(levels(df$group), 2)),
  seq(nrow(t(combn(levels(df$group), 2))))
)

ggplot(df, aes(x = group, y = shannon)) +
  # Outliers are removed, because otherwise each data point would be plotted twice;
  # as an outlier of boxplot and as a point of dotplot.
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2) +
  geom_signif(comparisons = comb, map_signif_level = FALSE, test = "wilcox.test") +
  theme(text = element_text(size = 10)) +
  ggtitle(f("Shannon alpha-diversity ({taxa.level})"))

# Show as a table
# tse %>% dplyr::select("taxon", "shannon") %>% show_table()
```

```{r}
metaphaln.alpha.diversity.df <- as.data.frame(colData(tse))
write.csv(metaphaln.alpha.diversity.df, "docs/plots/metaphaln_alpha_diversity_shannon.csv", row.names = FALSE)
```

There doesn't appear to be and significant difference in alpha diversity (Shannon diversity, Wilcoxson Rank-Sum test) between `Het` and `Homo`. 
The variance in diversity between `Homo` samples is higher than `Het`.

### Beta diversity - Bray-Curtis PCoA

Bray-Curtis dissimilarity quantifies the difference between the composition of groups in the range 0 to 1, where 1 is most dissimilar.
If both samples share the same taxa at the same relative abundance, their Bray-Curtis dissimilarity is 0. 
If they share no common taxa, they will have a dissimilarity of 1.

```{r beta_diversity_bray_all}
# PCoA
tse_glom <- runMDS(tse_glom,
  FUN = vegan::vegdist,
  method = "bray",
  name = "PCoA_BC",
  assay.type = "relabundance"
)

# Create ggplot object
p <- plotReducedDim(tse_glom, "PCoA_BC",
  colour_by = "group"
)

# Calculate explained variance
e <- attr(reducedDim(tse_glom, "PCoA_BC"), "eig")
rel_eig <- e / sum(e[e > 0])

# Add explained variance for each axis
p <- p + labs(
  x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%", ")", sep = ""),
  y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%", ")", sep = ""),
  title = "Bray-Curtis index, PCoA",
  subtitle = "Beta diversity of all samples"
)

p
```


```{r beta_diversity_bray_het_homo_only}
# We use Guila's theme for this plot - this requires a bit of messing about
# since scater::plotReducedDim renames the 'group' to 'colour_by'
# guila_palette <- c('#7D83F8', '#7DB8F3', 'grey')
guila_palette <- c("#7DB8F3", "grey")
names(guila_palette) <- c("Homo", "Het")
gulia_theme <- theme_classic() +
  theme(
    plot.title = element_text(face = "bold"),
    strip.background = element_rect(fill = "lightgrey", linewidth = 0),
    strip.text = element_text(face = "bold", colour = "black"),
    legend.position = "right", legend.justification = "left",
    text = element_text(size = 8)
  )

# Drop the Abx condition (and any features that were unique to that condition)
tse.hh <- dropCondition(tse_glom, group, "Abx")

# PCoA
tse.hh <- runMDS(tse.hh,
  FUN = vegan::vegdist,
  method = "bray",
  name = "PCoA_BC",
  assay.type = "relabundance"
)

permanova.adonis2 <- vegan::adonis2(t(assay(tse.hh, "relabundance")) ~ group,
  data = colData(tse.hh),
  permutations = 9999
)
# We run adonis2 here, just to get the relevant stats to associate with the PCoA
# (adonis [deprecated] is run again below, since I've not found an easy way to extract the
# coefficients from adonis2)
# Extract stats from PERMANOVA (adonis2) output
annotation <- paste(paste("SumOfSqs =", round(permanova.adonis2["group", ]$SumOfSqs, )),
  paste("R2 =", round(permanova.adonis2["group", ]$R2, 3)),
  paste("p =", round(permanova.adonis2["group", ]$Pr, 3)),
  sep = " ; "
)

# Create ggplot object
p <- scater::plotReducedDim(tse.hh, "PCoA_BC",
  colour_by = "group"
)

# Calculate explained variance
e <- attr(reducedDim(tse.hh, "PCoA_BC"), "eig")
rel_eig <- e / sum(e[e > 0])

# Add explained variance for each axis
p <- p + labs(
  x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%", ")", sep = ""),
  y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%", ")", sep = ""),
  title = "Bray-Curtis index, PCoA",
  subtitle = "Beta diversity of Het vs Homo samples"
) +
  # Additional theming, point styles to follow Guila's theme
  geom_point(aes(fill = colour_by), shape = 21, color = "black") +
  scale_fill_manual(values = guila_palette) +
  ggforce::geom_mark_hull(aes(fill = colour_by), 
                expand = unit(1.9, "mm"),
                radius = unit(2, "mm"),
                concavity = 4,
                colour = NA,
                alpha = 0.2) +
  # ggforce::geom_mark_ellipse(aes(fill = colour_by),
  #   expand = unit(2, "mm"),
  #   radius = unit(1, "mm"),
  #   n=100,
  #   tol=0.01,
  #   colour = NA,
  #   alpha = 0.2
  # ) +
  labs(
    caption = annotation,
  ) +
  guides(
    colour = "none",
    fill = guide_legend(title = "Group")
  ) +
  gulia_theme

p

# Save the plot in multiple formats
plot.size <- c(80, 80)
ggsave("docs/plots/Bray_Curtis_PCoA_HetHomo.tiff", p,
  device = "tiff", compression = "lzw",
  units = "mm", width = plot.size[1], height = plot.size[2], dpi = 600
)

ggsave("docs/plots/Bray_Curtis_PCoA_HetHomo.png", p,
  device = "png",
  units = "mm", width = plot.size[1], height = plot.size[2], dpi = 600
)

ggsave("docs/plots/Bray_Curtis_PCoA_HetHomo.svg", p,
  device = "svg",
  units = "mm", width = plot.size[1], height = plot.size[2], dpi = 600
)
```

```{r beta_diversity_jaccard_het_homo_only, include=FALSE, eval=FALSE}
# Drop the Abx condition (and any features that were unique to that condition)
tse.hh <- dropCondition(tse_glom, group, "Abx")

# PCoA
tse.hh <- runMDS(tse.hh,
  FUN = vegan::vegdist,
  method = "jaccard",
  name = "PCoA_Jaccard",
  assay.type = "relabundance"
)

# Create ggplot object
p <- plotReducedDim(tse.hh, "PCoA_Jaccard",
  colour_by = "group"
)

# Calculate explained variance
e <- attr(reducedDim(tse.hh, "PCoA_Jaccard"), "eig")
rel_eig <- e / sum(e[e > 0])

# Add explained variance for each axis
p <- p + labs(
  x = paste("PCoA 1 (", round(100 * rel_eig[[1]], 1), "%", ")", sep = ""),
  y = paste("PCoA 2 (", round(100 * rel_eig[[2]], 1), "%", ")", sep = ""),
  title = "Jaccard Beta Diversity, PCoA",
  subtitle = "Het vs Homo samples"
)

p
```

When we exclude the `Abx` samples, the `Het` and `Homo` samples show clear seperation based on the first two PCoA dimensions.

### PERMANOVA

PERMANOVA can be used to test which taxa most contribute to the differences in overall relative abundance between `Het` and `Homo` (based on the centroid and dispersion of all taxonomic abundances at the `r taxa.level` level).

```{r permanova}
library(vegan)

permanova <- vegan::adonis(t(assay(tse.hh, "relabundance")) ~ group,
  data = colData(tse.hh),
  permutations = 9999
)

# P-value
permanova.p <- as.data.frame(permanova$aov.tab)["group", "Pr(>F)"]
print(paste0("PERMANOVA composition differences p-value, Het vs Homo: ", permanova.p))

coef <- coefficients(permanova)["group1", ]
top.coef <- sort(head(coef[rev(order(abs(coef)))], 20))

# Check the homogeneity condition (PERMDISP2)
tse.hh <- runNMDS(
  tse.hh,
  FUN = vegan::vegdist,
  name = "NMDS_BC",
  nmdsFUN = "monoMDS",
  exprs_values = "relabundance",
  keep_dist = TRUE
)

# perma.df <- data.frame(
#   feature=factor(names(coef),unique(names(coef)),
#   coefficients=coef)
# )

show_table(data.frame(top.coef) %>% arrange(-top.coef))

ggplot(
  data.frame(
    x = top.coef,
    y = factor(
      names(top.coef),
      unique(names(top.coef))
    )
  ),
  aes(x = x, y = y)
) +
  geom_bar(stat = "identity") +
  labs(x = "", y = "coefficient", title = "Top Taxa") +
  theme_bw()
```

A PERMANOVA p-value <0.05 (p=0.0071, 9999 permutations) suggests that the centroids and/or dispersion for the 
Bray-Curis dissimilarity between the groups is different.

The plot of PERMANOVA coefficients, which can be interpreted as an effect size, 
give an indication of the taxa contributing to the difference between groups 
(in this case, based on Bray-Curtis dissimilarity).

PERMDISP2 test for homogeneity of group dispersions:
```{r permadisp2}
print("PERMDISP2 test for homogeneity of group dispersions: ")
permadisp <- vegan::betadisper(
  attr(reducedDim(tse.hh, "NMDS_BC"), "dist"),
  colData(tse.hh)$group
)
anova(permadisp)

permadisp.scores <- scores(permadisp)
permadisp.scores

TukeyHSD(permadisp)

mean_distances <- permadisp$group.distances
print("Mean distances to group centroids (as in box plot):")
print(mean_distances)

png("docs/plots/permadisp2_boxplot_hethomo.png",
  width = 100, height = 100, units = "mm", res = 300
)
boxplot(permadisp, xlab = "Group")
dev.off()

png("docs/plots/permadisp2_plot_hethomo.png",
  width = 100, height = 100, units = "mm", res = 300
)
plot(permadisp)
dev.off()
```

```{r permdisp2_pcoa_to_centroid_plot}
library(ggplot2)
library(tidyr)
library(dplyr)

# Create dataframe for sites with Group column
sites_df <- as.data.frame(permadisp.scores$sites)
colnames(sites_df) <- c("PCoA1", "PCoA2") # Explicitly name columns
sites_df$Group <- c(rep("Het", 5), rep("Homo", 5))

# Create dataframe for centroids
centroids_df <- as.data.frame(permadisp.scores$centroids)
colnames(centroids_df) <- c("PCoA1", "PCoA2") # Explicitly name columns
centroids_df$Group <- rownames(centroids_df)

# Create segments dataframe by joining sites with their corresponding centroids
segments_df <- sites_df %>%
  as.data.frame() %>% # Ensure it's a data frame
  dplyr::left_join(centroids_df %>% as.data.frame(), by = "Group", suffix = c(".site", ".centroid"))

# Create the plot
p <- ggplot() +
  # Add segments from points to centroids
  geom_segment(
    data = segments_df,
    aes(
      x = PCoA1.site, y = PCoA2.site,
      xend = PCoA1.centroid, yend = PCoA2.centroid,
      color = Group
    ),
    alpha = 0.5
  ) +
  # Add points for sites
  geom_point(
    data = sites_df,
    aes(x = PCoA1, y = PCoA2, color = Group),
    size = 3
  ) +
  # Add points for centroids, with show.legend=FALSE
  geom_point(
    data = centroids_df,
    aes(x = PCoA1, y = PCoA2, color = Group),
    shape = 8, size = 4, show.legend = FALSE
  ) +
  # Customize appearance
  theme_bw() +
  ggtitle("PERMADISP2 scores from Bray-Curtis PCoA") +
  theme(plot.title = element_text(hjust = 0.5))

p

# Save the plot in multiple formats
ggsave("docs/plots/permadisp2_pcoa_to_centroid_plot.tiff", p,
  device = "tiff", compression = "lzw",
  units = "mm", width = 100, height = 100, dpi = 600
)

ggsave("docs/plots/permadisp2_pcoa_to_centroid_plot.png", p,
  device = "png",
  units = "mm", width = 100, height = 100, dpi = 600
)

ggsave("docs/plots/permadisp2_pcoa_to_centroid_plot.svg", p,
  device = "svg",
  units = "mm", width = 100, height = 100, dpi = 600
)
```

```{r permdisp2_boxplot}
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggpubr) # for stat_compare_means

# Create dataframe with distances and groups
distances_df <- data.frame(
  Distance = permadisp$distances,
  Group = c(rep("Het", 5), rep("Homo", 5))
)

# Create the boxplot with consistent colors
p <- ggplot(distances_df, aes(x = Group, y = Distance, color = Group)) +
  geom_boxplot(width = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 3, alpha = 0.6) +
  stat_compare_means(
    method = "anova",
    label.y.npc = 0.95,
    label = "p.format", # Show actual p-value
    bracket.size = 0.5,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c("Het" = "#729ECE", "Homo" = "#FF9E4A")) +
  theme_bw() +
  ggtitle("PERMADISP2 distances to centroid") +
  ylab("Distance to Centroid") +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none"
  ) # Remove legend as it's redundant

p

# Save the plot
ggsave("docs/plots/permadisp2_distances_boxplot.tiff", p,
  device = "tiff", compression = "lzw",
  units = "mm", width = 100, height = 100, dpi = 600
)

ggsave("docs/plots/permadisp2_distances_boxplot.png", p,
  device = "png",
  units = "mm", width = 100, height = 100, dpi = 600
)

ggsave("docs/plots/permadisp2_distances_boxplot.svg", p,
  device = "svg",
  units = "mm", width = 100, height = 100, dpi = 600
)

# Print the actual ANOVA results
print(summary(aov(Distance ~ Group, data = distances_df)))

# Write distances dataframe to CSV
write.csv(distances_df, "docs/plots/permadisp2_distances.csv", row.names = FALSE)
```

**However**, the PERMDISP2 test for homogeneity of group dispersions fails (p = 0.012), so the dispesions of the groups are different 
enough that we should interpret the PERMANOVA results with caution.

### ANCOM-BC2 for differential abundance

ANCOM-BC2 fits linear mixed-models to estimate the log-fold change of taxa between groups, 
and includes terms to account for sampling fraction and per-taxa sequencing efficiency. 

We drop the `Abx` condition - this is both not of interest for the main comparison,
very different to the other conditions giving many 'significant' differences.

We exclude any taxa that aren't detected in at least 90 % of samples.

#### Species level differential abundance

```{r ancombc2_species}
# https://microbiome.github.io/OMA/differential-abundance.html#ancom-bc

nocache <- FALSE
# nocache <- TRUE

library(ANCOMBC)

tse.hh <- tse[, tse$group %in% c("Het", "Homo")]
tse.hh$group <- droplevels(tse.hh$group)

# Drop the Abx condition (and any features that were unique to that condition)
# tse.hh <- dropCondition(tse, group, "Abx")
# tse.hh <- tse.brkn[rowSums(assay(tse.brkn, "counts")) > 0, ]
# tse.hh <- minCounts(tse.hh, 10)

# Run ANCOM-BC2 on only Het/Homo, excluding Abx
# We set prv_cut = 0.9 (taxon must be present in 90% of samples) otherwise
# we get errors related to zero-variance taxa across groups.
# We can optionally use 'box' as a random effect to account for variation due
# to 'cage effect', (all `Het` mice were in the same box, others in various boxes).
ancomb_datafile <- f("data/ancombc2.metaphlan.{taxa.level.short}.rds")
if (!file.exists(ancomb_datafile) || nocache) {
  ancombc2_out <- ancombc2(
    data = tse.hh,
    assay_name = "counts",
    fix_formula = "group",
    # rand_formula = "(1|box)",
    p_adj_method = "fdr",
    prv_cut = 0.9,
    group = "group",
    tax_level = taxa.level,
    struc_zero = TRUE,
    neg_lb = TRUE,
    global = TRUE
  )
  saveRDS(ancombc2_out, file = ancomb_datafile)
} else {
  ancombc2_out <- readRDS(ancomb_datafile)
}
```

Log-fold changes estimated by ANCOM-BC2.
We exclude and taxa that failed sensitivity analysis of pseudo-count addition for zero count measurements.

```{r ancombc2_species_lfc_table}
taxa_lfc <- ancombc2_out$res %>% filter(passed_ss_groupHomo == TRUE)

rownames(taxa_lfc) <- taxa_lfc$taxon

taxa_lfc[[taxa.level]] <- paste0(
  "<a href=\"https://gtdb.ecogenomic.org/tree?r=",
  tolower(taxa.level.short), "__", taxa_lfc$taxon, "\" target='_blank'>",
  taxa_lfc$taxon,
  "</a>"
)

taxa_lfc %>%
  dplyr::select(taxa.level, lfc_groupHomo, q_groupHomo) %>%
  filter(q_groupHomo < 0.05) %>%
  arrange(-lfc_groupHomo) %>%
  show_table(pageLength = 100, escape = FALSE, rownames = FALSE)
```

The full ANCOM-BC2 result table, including species where FDR > 0.05.
```{r}
taxa_lfc %>%
  dplyr::select(taxa.level, lfc_groupHomo, q_groupHomo, p_groupHomo) %>%
  arrange(q_groupHomo) %>%
  show_table(pageLength = 10, escape = FALSE, rownames = FALSE)
```

```{r mia_relabundance_species_top_lfc}
# disable scientific notation (for the x-axis)
scipen.save <- options()$scipen
options(scipen = 999)

relab_cutoff <- 0.005
sig <- taxa_lfc %>% filter(q_groupHomo < 0.05)
sig.taxa <- sapply(sig$taxon, function(x) {
  str_split(x, ":", simplify = TRUE)[2]
})
# n_lines <- nrow(taxa_lfc %>% filter(Het_relabundance_mean > relab_cutoff | Homo_relabundance_mean > relab_cutoff))
tse.sig.abund <- tse.hh[rowData(tse)[[taxa.level]] %in% sig.taxa, ]
rownames(tse.sig.abund) <- rowData(tse.sig.abund)[[taxa.level]]

plotAbundanceDensity(tse.sig.abund,
  layout = "density", assay.type = "relabundance",
  colour_by = "group", point_alpha = 0.4,
  # n = length(sig$taxon),
) +
  scale_x_log10()

plotAbundanceDensity(tse.sig.abund,
  layout = "point", assay.type = "relabundance",
  colour_by = "group", point_alpha = 0.8,
  # n = length(sig$taxon),
) +
  scale_x_log10()

options(scipen = scipen.save)
```

```{r methaphlan_top_lfc_boxplot, fig.width=8, fig.height=8}
# sig$species <- sapply(rownames(sig), function (x) { str_split(x, ":", simplify = TRUE)[2]})

sig.long <- meltAssay(tse.sig.abund,
  assay.type = "relabundance",
  add_row_data = TRUE, add_col_data = TRUE,
  sample_name = "sample", feature_name = "species"
)

sig.long$species <- (
  factor(sig.long$species,
    levels = unique(sig.long$species[order(sig.long$relabundance, decreasing = TRUE)])
))

(
  ggplot(
    sig.long,
    aes(x = relabundance, color = group)
  ) +
    # geom_density(alpha = 0.5, adjust=1) +
    geom_boxplot(width = 0.1, position = position_nudge(y = 1.2), outlier.shape = NA) +
    # geom_boxplot() +
    # geom_rug(sides = "b", size = 1, length = unit(0.1, "npc")) +
    geom_point(aes(y = species), alpha = 0.8, size = 1) +
    facet_wrap(~species, scales = "free_y", ncol = 1) +
    # facet_grid(species ~ ., scales = "free_y") +
    scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
    scale_fill_manual(values = scater:::.get_palette("tableau10medium")) +
    scale_colour_manual(values = scater:::.get_palette("tableau10medium")) +
    theme_minimal() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.major.x = element_line(color = "grey", linetype = "dotted"),
      panel.grid.minor.x = element_line(color = "grey", , linetype = "dotted"),
      panel.grid.minor = element_blank(),
      panel.background = element_blank(),
      strip.background = element_blank(),
      strip.text.x = element_text(size = 10),
      # strip.text = element_blank(), # remove facet labels
      axis.text.x = element_text(angle = 90, hjust = 1),
      axis.text.y = element_blank(), # remove y labels
      # axis.text.y = element_text(angle = 90, vjust=1, hjust=0.5), # y labels
      panel.spacing.y = unit(-1, "lines"),
      axis.line = element_line(color = "black"),
      axis.line.y = element_line(color = "black"),
      legend.position = "bottom"
    ) +
    labs(
      x = "Relative abundance",
      y = "",
      fill = "group",
      # title = "Differentially abundant species") +
    ) +
    guides(fill = guide_legend(title = "Group")) +
    ggprism::theme_prism() +
    theme(axis.text.y = element_blank())
)

ggsave("docs/plots/Part_E_metaphlan_boxplot_het_homo_plot.tiff",
  units = "mm", device = "tiff", compression = "lzw",
  width = 100, height = 100, dpi = 600
)

ggsave("docs/plots/Part_E_metaphlan_boxplot_het_homo_plot.pdf",
  device = "pdf", units = "mm",
  width = 100, height = 100, dpi = 600
)

write.csv(sig.long, "docs/plots/metaphlan_boxplot_het_homo_plot.csv", row.names = FALSE)
```

- PDF version: [plots/metaphlan_boxplot_het_homo_plot.pdf](plots/metaphlan_boxplot_het_homo_plot.pdf)

All species with at least 10 counts in 50% of samples, log relative abudance box plots.

```{r methaphlan_all_lfc_boxplot, fig.width=8, fig.height=16}
# sig$species <- sapply(rownames(sig), function (x) { str_split(x, ":", simplify = TRUE)[2]})

sig.long <- meltAssay(tse_glom.filtered[, tse_glom.filtered$group %in% c("Het", "Homo")],
  assay.type = "relabundance",
  add_row_data = TRUE, add_col_data = TRUE,
  sample_name = "sample", feature_name = "species"
)

sig.long$species <- (
  factor(sig.long$species,
    levels = unique(sig.long$species[order(sig.long$relabundance, decreasing = TRUE)])
))

ggplot(
  sig.long,
  aes(x = relabundance, color = group)
) +
  geom_boxplot(width = 0.1, position = position_nudge(y = 1.2), outlier.shape = NA) +
  geom_point(aes(y = species), alpha = 0.8, size = 1) +
  facet_wrap(~species, scales = "free_y", ncol = 2, strip.position = "left") +
  scale_x_log10(labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  scale_fill_manual(values = scater:::.get_palette("tableau10medium")) +
  scale_colour_manual(values = scater:::.get_palette("tableau10medium")) +
  theme_minimal() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.major.x = element_line(color = "grey", linetype = "dotted"),
    panel.grid.minor.x = element_line(color = "grey", , linetype = "dotted"),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    strip.background = element_blank(),
    strip.text.x = element_text(size = 10),
    strip.text.y.left = element_text(size = 10, angle = 0, hjust = 1),
    strip.placement = "outside",
    # strip.text = element_blank(), # remove facet labels
    axis.text.x = element_text(angle = 90, hjust = 1),
    axis.text.y = element_blank(), # remove y labels
    # axis.text.y = element_text(angle = 90, vjust=1, hjust=0.5), # y labels
    panel.spacing.y = unit(-1, "lines"),
    axis.line = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    legend.position = "bottom"
  ) +
  labs(
    x = "Relative abundance",
    y = "",
    fill = "group",
    title = "Differentially abundant species"
  ) +
  guides(fill = guide_legend(title = "Group"))
```

#### Genus level differential abundance

When counts are agglomerated to the Genus level, **no significant differental abundances are found** after multiple testing correction.

Top differentially abundance Genus with `p <= 0.05` (but NOT `FDR <= 0.05`) are are shown for reference.

```{r ancombc2_genus}
taxa.level <- "Genus"
taxa.level.short <- "G"

nocache <- FALSE
# nocache <- TRUE

library(ANCOMBC)

tse.hh.G <- tse[, tse$group %in% c("Het", "Homo")]
tse.hh.G$group <- droplevels(tse.hh.G$group)

# Agglomerate to Genus level
tse.hh.G <- mia::agglomerateByRank(tse.hh.G, rank = taxa.level, onRankOnly = TRUE)

# Drop the Abx condition (and any features that were unique to that condition)
# tse.hh.G <- dropCondition(tse, group, "Abx")
# tse.hh.G <- tse.brkn[rowSums(assay(tse.brkn, "counts")) > 0, ]
# tse.hh.G <- minCounts(tse.hh.G, 10)

# Run ANCOM-BC2 on only Het/Homo, excluding Abx
# We set prv_cut = 0.9 (taxon must be present in 90% of samples) otherwise
# we get errors related to zero-variance taxa across groups.
# We can optionally use 'box' as a random effect to account for variation due
# to 'cage effect', (all `Het` mice were in the same box, others in various boxes).
ancomb_datafile <- f("data/ancombc2.metaphlan.{taxa.level.short}.rds")
if (!file.exists(ancomb_datafile) || nocache) {
  ancombc2_out <- ancombc2(
    data = tse.hh.G,
    assay_name = "counts",
    fix_formula = "group",
    # rand_formula = "(1|box)",
    p_adj_method = "fdr",
    prv_cut = 0.9,
    group = "group",
    tax_level = taxa.level,
    struc_zero = TRUE,
    neg_lb = TRUE,
    global = TRUE
  )
  saveRDS(ancombc2_out, file = ancomb_datafile)
} else {
  ancombc2_out <- readRDS(ancomb_datafile)
}
```

```{r ancombc2_genus_lfc_table}
taxa_lfc <- ancombc2_out$res %>% filter(p_groupHomo <= 0.05)

rownames(taxa_lfc) <- taxa_lfc$taxon

taxa_lfc[[taxa.level]] <- paste0(
  "<a href=\"https://gtdb.ecogenomic.org/tree?r=",
  tolower(taxa.level.short), "__", taxa_lfc$taxon, "\" target='_blank'>",
  taxa_lfc$taxon,
  "</a>"
)

taxa_lfc %>%
  dplyr::select(taxa.level, lfc_groupHomo, q_groupHomo, p_groupHomo) %>%
  arrange(-lfc_groupHomo) %>%
  show_table(pageLength = 100, escape = FALSE, rownames = FALSE)
```

```{r mia_relabundance_genus_top_lfc, eval=FALSE, include=FALSE}
# disable scientific notation (for the x-axis)

scipen.save <- options()$scipen

options(scipen = 999)

relab_cutoff <- 0.005
sig <- taxa_lfc
tse.sig.abund <- tse.hh.G[rowData(tse.hh.G)[[taxa.level]] %in% sig$taxon, ]
rownames(tse.sig.abund) <- rowData(tse.sig.abund)[[taxa.level]]

plotAbundanceDensity(tse.sig.abund,
  layout = "density", assay.type = "relabundance",
  colour_by = "group", point_alpha = 0.4,
  n = length(sig$taxon),
) +
  scale_x_log10()

plotAbundanceDensity(tse.sig.abund,
  layout = "point", assay.type = "relabundance",
  colour_by = "group", point_alpha = 0.8,
  n = length(sig$taxon),
) +
  scale_x_log10()

options(scipen = scipen.save)
```

#### Differential abundance at Order level

After some exploration of DA at different levels, Order seems the most interesting alternative taxonomic level to examine.

```{r ancombc2_order}
taxa.level <- "Order"
taxa.level.short <- "O"

nocache <- FALSE
# nocache <- TRUE

library(ANCOMBC)

tse.hh.O <- tse[, tse$group %in% c("Het", "Homo")]
tse.hh.O$group <- droplevels(tse.hh.O$group)

# Agglomerate to Genus level
tse.hh.O <- mia::agglomerateByRank(tse.hh.O, rank = taxa.level, onRankOnly = TRUE)

# Drop the Abx condition (and any features that were unique to that condition)
# tse.hh.O <- dropCondition(tse, group, "Abx")
# tse.hh.O <- tse.brkn[rowSums(assay(tse.brkn, "counts")) > 0, ]
# tse.hh.O <- minCounts(tse.hh.O, 10)

# Run ANCOM-BC2 on only Het/Homo, excluding Abx
# We set prv_cut = 0.9 (taxon must be present in 90% of samples) otherwise
# we get errors related to zero-variance taxa across groups.
# We can optionally use 'box' as a random effect to account for variation due
# to 'cage effect', (all `Het` mice were in the same box, others in various boxes).
ancomb_datafile <- f("data/ancombc2.metaphlan.{taxa.level.short}.rds")
if (!file.exists(ancomb_datafile) || nocache) {
  ancombc2_out <- ancombc2(
    data = tse.hh.O,
    assay_name = "counts",
    fix_formula = "group",
    # rand_formula = "(1|box)",
    p_adj_method = "fdr",
    prv_cut = 0.9,
    group = "group",
    tax_level = taxa.level,
    struc_zero = TRUE,
    neg_lb = TRUE,
    global = TRUE
  )
  saveRDS(ancombc2_out, file = ancomb_datafile)
} else {
  ancombc2_out <- readRDS(ancomb_datafile)
}
```

```{r ancombc2_order_lfc_table}
taxa_lfc <- ancombc2_out$res %>% filter(p_groupHomo <= 0.05)

rownames(taxa_lfc) <- taxa_lfc$taxon

taxa_lfc[[taxa.level]] <- paste0(
  "<a href=\"https://gtdb.ecogenomic.org/tree?r=",
  tolower(taxa.level.short), "__", taxa_lfc$taxon, "\" target='_blank'>",
  taxa_lfc$taxon,
  "</a>"
)

taxa_lfc %>%
  dplyr::select(taxa.level, lfc_groupHomo, q_groupHomo, p_groupHomo) %>%
  arrange(-lfc_groupHomo) %>%
  show_table(pageLength = 100, escape = FALSE, rownames = FALSE)
```

```{r mia_relabundance_order_top_lfc}
# disable scientific notation (for the x-axis)

scipen.save <- options()$scipen

options(scipen = 999)

relab_cutoff <- 0.005
sig <- taxa_lfc
tse.sig.abund <- tse.hh.O[rowData(tse.hh.O)[[taxa.level]] %in% sig$taxon, ]
rownames(tse.sig.abund) <- rowData(tse.sig.abund)[[taxa.level]]

plotAbundanceDensity(tse.sig.abund,
  layout = "density", assay.type = "relabundance",
  colour_by = "group", point_alpha = 0.4,
  n = length(sig$taxon),
) +
  scale_x_log10()

plotAbundanceDensity(tse.sig.abund,
  layout = "point", assay.type = "relabundance",
  colour_by = "group", point_alpha = 0.8,
  n = length(sig$taxon),
) +
  scale_x_log10()


options(scipen = scipen.save)
```

*Christensenellales* (in the class *Clostridia*) is underrepresented in Homo samples, while *Verrucomicrobiales* (parent of family *Akkermansiaceae*) is overrepresented in Het samples.

```{r}
htmltools::includeHTML("analysis/acknowledging.html")
```
