---
title: "Cysteinuria Project"
author: "Giulia Iacono"
date: "Last knitted on `r format(Sys.time(), '%d %B, %Y')`"
output: 
  
---

# *Differential abundance
## URINE

```{r}
# limma_list <- list()
#ur <- SummarizedExperiment(assays = metab_ur_df, colData = metadata_urine, rowData = metab_info_ur)

design <- model.matrix(~0+Class, metadata_urine) 
colnames(design) <- gsub("Class|-", "", colnames(design))
contr_matrix <- makeContrasts(HomoALA_vs_HomoCtrl = HomoALA-HomoCtrl, HomoCtrl_vs_HeteroCtrl = HomoCtrl-HeteroCtrl, levels = colnames(design))
vfit <- lmFit(metab_ur_df, design)
vfit <- contrasts.fit(vfit, contrasts = contr_matrix)
efit <- eBayes(vfit)

summary(decideTests(efit, p.value = 0.05, lfc = 0.5)) 
# topTreat(efit, coef = "HomoALA_vs_HomoCtrl")

test_variable <- "urine"
folder_id <- paste0(drive_path, 'limma/', test_variable)
dir.create(folder_id)

limma_list <- limma_molecules_process(list = limma_list, 
                                    data = metab_ur_df, 
                                    metadata = metadata_urine,
                                    coef_name = c("HomoALA_vs_HomoCtrl", "HomoCtrl_vs_HeteroCtrl"), 
                                    title = test_variable, 
                                    efit = efit, 
                                    folder_id = folder_id, 
                                    logFC_threshold = 0.5)

urine_molecules_list <- list("HomoALA_vs_HomoCtrl" = limma_list$urine$HomoALA_vs_HomoCtrl$sig_hits$Molecule,
                       "HomoCtrl_vs_HeteroCtrl" = limma_list$urine$HomoCtrl_vs_HeteroCtrl$sig_hits$Molecule) 

length(urine_molecules_list$HomoCtrl_vs_HeteroCtrl) # 195
length(urine_molecules_list$HomoALA_vs_HomoCtrl) # 563

limma_list <- molecules_pathway_analysis(list = limma_list, 
                                         data_info = metab_info_ur, 
                                         coef_name = "HomoALA_vs_HomoCtrl", 
                                         title = test_variable, 
                                         molecules = urine_molecules_list$HomoALA_vs_HomoCtrl, 
                                         folder_id = folder_id, fella = mmu.hsa)


limma_list <- molecules_pathway_analysis(list = limma_list, 
                                         data_info = metab_info_ur, 
                                         coef_name = "HomoCtrl_vs_HeteroCtrl", 
                                         title = test_variable, 
                                         molecules = urine_molecules_list$HomoCtrl_vs_HeteroCtrl, 
                                         folder_id = folder_id, fella = mmu.hsa)

saveRDS(limma_list, paste0(drive_path, "limma/limma_list.rds"))


limma_list$urine$HomoALA_vs_HomoCtrl$sig_hits %>% group_by(direction) %>% dplyr::count()
```

```{r}
# Merge old urine table with new urine table
## HomoCtrl vs HeteroCtrl
read.csv(paste0(drive_path, "limma/urine/old/HomoCtrl_vs_HeteroCtrl_da_features_urine_old.csv")) %>%
  dplyr::select(!direction) %>%
  right_join(limma_list$urine$HomoCtrl_vs_HeteroCtrl$sig_hits) %>%
  dplyr::select(!c(AveExpr, t, P.Value, adj.P.Val, B)) %>%
  write.csv(paste0(drive_path, "limma/HomoCtrl_vs_HeteroCtrl_da_features_urine.csv"), row.names = FALSE)

## HomoALA vs HomoCtrl
read.csv(paste0(drive_path, "limma/urine/old/HomoALA_vs_HomoCtrl_da_features_urine_old.csv")) %>%
  dplyr::select(!direction) %>%
  right_join(limma_list$urine$HomoALA_vs_HomoCtrl$sig_hits) %>%
  dplyr::select(!c(AveExpr, t, P.Value, adj.P.Val, B)) %>%
  write.csv(paste0(drive_path, "limma/HomoALA_vs_HomoCtrl_da_features_urine.csv"), row.names = FALSE)
```


## SERUM

```{r}
design <- model.matrix(~0+Class, metadata_serum) 
colnames(design) <- gsub("Class|-", "", colnames(design))
contr_matrix <- makeContrasts(HomoALA_vs_HomoCtrl = HomoALA-HomoCtrl, HomoCtrl_vs_HeteroCtrl = HomoCtrl-HeteroCtrl, levels = colnames(design))
vfit <- lmFit(metab_se_df, design)
vfit <- contrasts.fit(vfit, contrasts = contr_matrix)
efit <- eBayes(vfit)

summary(decideTests(efit, p.value = 0.05, lfc = 0.5)) 
# topTreat(efit, coef = "HomoALA_vs_HomoCtrl")

test_variable <- "serum"
folder_id <- paste0(drive_path, 'limma/', test_variable)
dir.create(folder_id)

limma_list <- limma_molecules_process(list = limma_list, 
                                    data = metab_se_df, 
                                    metadata = metadata_serum,
                                    coef_name = c("HomoALA_vs_HomoCtrl", "HomoCtrl_vs_HeteroCtrl"), 
                                    title = test_variable, 
                                    efit = efit, 
                                    folder_id = folder_id, 
                                    logFC_threshold = 0.5)

serum_molecules_list <- list("HomoALA_vs_HomoCtrl" = limma_list$serum$HomoALA_vs_HomoCtrl$sig_hits$Molecule,
                             "HomoCtrl_vs_HeteroCtrl" = limma_list$serum$HomoCtrl_vs_HeteroCtrl$sig_hits$Molecule) 

length(serum_molecules_list$HomoCtrl_vs_HeteroCtrl) # 50
length(serum_molecules_list$HomoALA_vs_HomoCtrl) # 114

limma_list <- molecules_pathway_analysis(list = limma_list, 
                                         data_info = metab_info_se, 
                                         coef_name = "HomoALA_vs_HomoCtrl", 
                                         title = test_variable, 
                                         molecules = serum_molecules_list$HomoALA_vs_HomoCtrl, 
                                         folder_id = folder_id, fella = mmu.hsa)


limma_list <- molecules_pathway_analysis(list = limma_list, 
                                         data_info = metab_info_se, 
                                         coef_name = "HomoCtrl_vs_HeteroCtrl", 
                                         title = test_variable, 
                                         molecules = serum_molecules_list$HomoCtrl_vs_HeteroCtrl, 
                                         folder_id = folder_id, fella = mmu.hsa)

saveRDS(limma_list, paste0(drive_path, "limma/limma_list.rds"))

limma_list$serum$HomoALA_vs_HomoCtrl$sig_hits %>% group_by(direction) %>% dplyr::count()
```

```{r}
# Not used
data <- as.data.frame(limma_list$urine$HomoALA_vs_HomoCtrl$heat_sig_hits)
#identical(rownames(ms_combined_heat_intervals), biomarker_Classes$Molecule) # TRUE
#identical(colnames(ms_combined_heat_intervals), rownames(df))
#a <- a[rownames(biomarker_heat_classes), ]

colAnn <- HeatmapAnnotation(Class = metadata_urine$Class, 
                            col = list(Class = group_palette), 
                            annotation_name_side = "right"#,
                              #height = unit(0.1, "cm"),
                              #annotation_legend_param = list(direction = "vertical")
                            ) 

colHm <- colorRamp2(breaks = seq(min(data, na.rm = T), max(data, na.rm = T), length = 7), colors=c('#2166ac','#67a9cf','#d1e5f0', '#f7f7f7', '#fddbc7', '#ef8a62', '#b2182b'))

p <- 
  
      draw(Heatmap(data, 
        name = "z-score",
        column_title = "HomoALA vs HomoCtrl",
        column_title_gp = gpar(fontface = "bold"),
        row_title = "Metabolites",  
        show_column_names = FALSE,
        show_row_names = TRUE,
        row_names_gp = gpar(fontsize = 6),
        #column_order = colnames(data[, biomarker_small_molecules_metadata %>% arrange(Group, Days) %>% pull(Patient_days)]),
        top_annotation = colAnn, 
        col = colHm,
        #right_annotation = annotate_heatmap_row(metadata = clad_ms_heat_path, 
        #                                        row_id = Feature, 
        #                                        row_or = clad_ms_heat_path$Feature, 
        #                                        annotations = c("Class", "Significant"), 
        #                                        path_palette = c("#00AFB5", "#184E77","#7AE7C7", "#D81159")), #"#FF6700", 
        heatmap_legend_param = list(direction = "horizontal")),
        padding = unit(c(0, 0, 0, 0), "cm"),
     merge_legend = TRUE,
     heatmap_legend_side = "left")

pdf(paste0(drive_path, "limma/urine/heat-sig.pdf"), width = 14, height = 10)
p
invisible(dev.off()) 
```

### DA features

```{r}
features <- rbind(read.csv("HomoALA_vs_HomoCtrl_da_features - urine.csv") %>% 
                    dplyr::select(Molecule, Rename, Metabolism) %>% 
                    mutate(Dataset = "Urine", Contrast = "HomoALA_vs_HomoCtrl"),
                  read.csv("HomoALA_vs_HomoCtrl_da_features - serum.csv") %>% 
                    dplyr::select(Molecule, Rename, Metabolism) %>%
                    mutate(Dataset = "Serum", Contrast = "HomoALA_vs_HomoCtrl"),
                  read.csv("HomoCtrl_vs_HeteroCtrl_da_features - urine.csv") %>% 
                    dplyr::select(Molecule, Rename, Metabolism)%>%
                    mutate(Dataset = "Urine", Contrast = "HomoCtrl_vs_HeteroCtrl"),
                  read.csv("HomoCtrl_vs_HeteroCtrl_da_features - serum.csv") %>% 
                    dplyr::select(Molecule, Rename, Metabolism) %>%
                    mutate(Dataset = "Serum", Contrast = "HomoCtrl_vs_HeteroCtrl") ) %>% 
  distinct() %>%
  mutate(Rename = gsub("*", "", Rename, fixed = TRUE),
         Rename = ifelse(Rename == "", Molecule, Rename)) #%>%
  #mutate(Rename = ifelse(Rename == "2-Hydroxybutyric acid", "4-Hydroxybutyric acid", Rename))

length(unique(features$Molecule)) #780 

# features[duplicated(features$Molecule), ]

unique_features <- features %>% dplyr::select(Molecule, Rename, Metabolism) %>% distinct() # 780
# unique_features[duplicated(unique_features$Molecule), ]
```

# ----------------------
## **HomoCtrl vs HetCtrl 
### PCA

```{r}
# Urine
m <- metadata_urine %>% filter(Class != "Homo-ALA")
d <- metab_ur_df[, m$SampleNumber]

set.seed(3)
a <- data.frame(adonis2(formula = as.formula(paste('t(d) ~', "Class")), 
                      data = m, 
                      distance = "eu", 
                      permutations = 999,
                      na.action = na.omit,
                      parallel = 6))

annotation <- paste(paste("SumOfSqs =", round(a["Class",]$SumOfSqs, 3)),
              paste("R2 =", round(a["Class",]$R2, 3)),
              paste("p =", round(a["Class",]$Pr..F., 3)), sep = " ; " )

pca <- pca_process(data = d, metadata = m)

p <- 

  ggplot(pca$data, aes(x = PC1, y = PC2)) +
  ggforce::geom_mark_ellipse(aes(fill = Class), expand = unit(2.5, "mm"), colour = NA, alpha = 0.2) +
  geom_point(aes(fill = Class), shape = 21, alpha = 0.8) +
  scale_fill_manual(values = group_palette) +
  theme +
  theme(legend.position = "none") +
  labs(title = "Urine metabolomics",
       caption = annotation,
       fill = "Group",
       x = paste0('PC1 ', round(pca$varexp[1], 2), '%'),
       y = paste0('PC2 ', round(pca$varexp[2], 2), '%')) +
  scale_y_continuous(limits = c(-55, 50)) +
  scale_x_continuous(limits = c(-40, 80))

ggsave(paste0(drive_path, "figure3/homo-het-urine-pca.tiff"), p, width = 2, height = 2, dpi = 600)
```


```{r}
# Serum
m <- metadata_serum %>% filter(Class != "Homo-ALA")
d <- metab_se_df[, m$SampleNumber]

set.seed(3)
a <- data.frame(adonis2(formula = as.formula(paste('t(d) ~', "Class")), 
                      data = m, 
                      distance = "eu", 
                      permutations = 999,
                      na.action = na.omit,
                      parallel = 6))

annotation <- paste(paste("SumOfSqs =", round(a["Class",]$SumOfSqs, 3)),
              paste("R2 =", round(a["Class",]$R2, 3)),
              paste("p =", round(a["Class",]$Pr..F., 3)), sep = " ; " )

pca <- pca_process(data = d, metadata = m)

p <- 

  ggplot(pca$data, aes(x = PC1, y = PC2)) +
  #stat_ellipse(aes(fill = Class), geom = 'polygon', type = 't', level = 0.9, alpha = 0.2) +
  ggforce::geom_mark_ellipse(aes(fill = Class), expand = unit(2.5, "mm"), colour = NA, alpha = 0.2) +
  geom_point(aes(fill = Class), shape = 21) +
  scale_fill_manual(values = group_palette) +
  theme +
  theme(legend.position = "none") +
  labs(title = "Serum metabolomics",
       caption = annotation,
       fill = "Group",
       x = paste0('PC1 ', round(pca$varexp[1], 2), '%'),
       y = paste0('PC2 ', round(pca$varexp[2], 2), '%')) +
  scale_y_continuous(limits = c(-30, 30)) +
  scale_x_continuous(limits = c(-45, 45))

ggsave(paste0(drive_path, "figure3/homo-het-serum-pca.tiff"), p, width = 2, height = 2, dpi = 600)

# Legend
pca$data <- pca$data %>%
  mutate(Class = ifelse(Class == "Homo-Ctrl", "Homozygote", "Heterozygote"))

p <- 

  ggplot(pca$data, aes(x = PC1, y = PC2)) +
  geom_point(aes(fill = Class), shape = 21) +
  ggforce::geom_mark_ellipse(aes(fill = Class), expand = unit(2.5, "mm"), colour = NA, alpha = 0.2) +
  scale_fill_manual(values = c("Homozygote" = "#7DB8F3", "Heterozygote" = "grey")) +
  theme +
  theme(legend.position = "bottom") +
  labs(fill = "Group")
ggsave(paste0(drive_path, "figure3/homo-het-pca-legend.tiff"), p, width = 3, height = 2, dpi = 600)
```


### Barplot

```{r}
u <- limma_list$urine$HomoCtrl_vs_HeteroCtrl$sig_hits %>%
  full_join(read.csv("HomoCtrl_vs_HeteroCtrl_da_features - urine.csv")[, c("Molecule", "Rename", "Metabolism", "Relevant")] ) %>%
  mutate(Dataset = "Urine",
         Rename = ifelse(Rename == "", Molecule, Rename),
         Rename = gsub("*", "", Rename, fixed = TRUE)) %>%
  dplyr::select(!Molecule) %>%
  rename(Rename = "Molecule", direction = "Direction") %>%
  mutate(Direction = gsub("HomoCtrl", "Increased in Homozygote", Direction),
         Direction = gsub("HeteroCtrl", "Decreased in Homozygote", Direction)) %>%
  relocate(Molecule) #%>%
  #mutate(Molecule = ifelse(Molecule == "2-Hydroxybutyric acid", "4-Hydroxybutyric acid", Molecule))
dim(u)[1] # 195

write.csv(u %>% dplyr::select(!Relevant), 
          paste0(drive_path, "tables/Suppl_table_HomoCtrl_vs_HeteroCtrl_da_features_urine.csv"), row.names = FALSE )

s <- limma_list$serum$HomoCtrl_vs_HeteroCtrl$sig_hits %>%
  full_join(read.csv("HomoCtrl_vs_HeteroCtrl_da_features - serum.csv")[, c("Molecule", "Rename", "Metabolism", "Relevant")] ) %>%
  mutate(Dataset = "Serum",
         Rename = ifelse(Rename == "", Molecule, Rename),
         Rename = gsub("*", "", Rename, fixed = TRUE)) %>%
  dplyr::select(!Molecule) %>%
  rename(Rename = "Molecule", direction = "Direction") %>%
  mutate(Direction = gsub("HomoCtrl", "Increased in Homozygote", Direction),
         Direction = gsub("HeteroCtrl", "Decreased in Homozygote", Direction)) %>%
  relocate(Molecule) #%>%
  #mutate(Molecule = ifelse(Molecule == "2-Hydroxybutyric acid", "4-Hydroxybutyric acid", Molecule))
dim(s)[1] #50

write.csv(s %>% dplyr::select(!Relevant), 
          paste0(drive_path, "tables/Suppl_table_HomoCtrl_vs_HeteroCtrl_da_features_serum.csv"), row.names = FALSE )

data <- rbind(u, s) %>%
  filter(Relevant == TRUE) %>%
  mutate(Significance = case_when(adj.P.Val >= 0.05 ~ "",
                                      adj.P.Val < 0.05 & adj.P.Val >= 0.01 ~ "*",
                                      adj.P.Val < 0.01 & adj.P.Val >= 0.001 ~ "**",
                                      adj.P.Val < 0.001 & adj.P.Val >= 0.0001 ~ "***",
                                      adj.P.Val < 0.0001 ~ "****") ) %>%
  mutate(Metabolism = gsub("and", "\nand", Metabolism),
         Metabolism = gsub(",", ",\n", Metabolism),
         Metabolism = factor(Metabolism, levels = c("Arginine", "Cysteine \nand methionine", "Histidine", 
                                                "Lysine \nand carnitine","Tryptophan", "Other",
                                                "Nucleotide", "Vitamin,\n flavonoid \nand folate"))) %>%
  group_by(Metabolism) %>%
  arrange(logFC) %>%
  mutate(Molecule = factor(Molecule, levels = unique(.$Molecule)))
dim(data) #40 + 10
  
p <- 
  ggplot(data, aes(x = logFC, y = Molecule)) +
    geom_col(aes(fill = Direction), colour = "black", size = 0.2) +
    geom_text(aes(label = Significance), nudge_x = 0.1, nudge_y = -0.2, size = 4) +
    theme +
    theme(strip.text.y = element_text(angle = 0),
          legend.position = "bottom") +
    geom_vline(xintercept = 0, lty = "dashed") +
    labs(x = "Log2FC", fill = "", y = "", title = "Differentially abundant metabolites") +
    scale_fill_manual(values = c("Increased in Homozygote" = "#7DB8F3", "Decreased in Homozygote" = "#CECECE")) +
    facet_grid(Metabolism~Dataset, scales = "free_y", space = "free_y") +
    scale_x_continuous(limits = c(-4, 6))
ggsave(paste0(drive_path, "figure3/homo-het-urine-serum-da-list_2.tiff"), p, width = 5, height = 6, dpi = 600)
```


### Heatmap

```{r}
# Urine
colHm <- colorRamp2(breaks = seq(-2, 2, length = 7), colors=c('#2166ac','#67a9cf','#d1e5f0', '#f7f7f7', '#fddbc7', '#ef8a62', '#b2182b'))

homo_vs_het_da_features_urine <- read.csv("HomoCtrl_vs_HeteroCtrl_da_features - urine.csv") %>% 
  filter(Relevant == "TRUE") %>%
  mutate(Rename = gsub("*", "", Rename, fixed = TRUE))
dim(homo_vs_het_da_features_urine)[1] # 40
m <- metadata_urine %>% filter(Class != "Homo-ALA") %>% 
  mutate(Class = as.character(Class),
         Class = ifelse(Class == "Hetero-Ctrl", "Heterozygote", "Homozygote"),
         Class = factor(Class, levels = c("Heterozygote", "Homozygote")))

data_urine <- as.data.frame(limma_list$urine$HomoCtrl_vs_HeteroCtrl$heat_sig_hits) %>% 
  rownames_to_column("Molecule") %>%
  right_join(homo_vs_het_da_features_urine) %>%
  dplyr::select(Rename, m$SampleNumber) %>%
  dplyr::slice(match(Rename, homo_vs_het_da_features_urine$Rename)) %>%
  column_to_rownames("Rename") %>%
  as.matrix()

colAnn <- HeatmapAnnotation(Group = m$Class, 
                            col = list(Group = c("Homozygote" = "#7DB8F3", "Heterozygote" = "grey")), 
                            annotation_name_side = "right") 
identical(rownames(data_urine), homo_vs_het_da_features_urine$Rename) #TRUE

rowAnn <- rowAnnotation(Metabolism = homo_vs_het_da_features_urine$Metabolism,
                        col = list(Metabolism = metabolism_palette))

tiff(paste0(drive_path, "supplementary/homo-het-urine-heatmap.tiff"), width = 16, height = 19, units = "cm", res = 600)
pushViewport(viewport(gp = gpar(fontfamily = "Arial", fontsize = 8)))
p <- 
      Heatmap(data_urine, 
        name = "Intensity",
        column_title = "Urine",
        column_title_gp = gpar(fontface = "bold"),
        row_title = "Metabolite",  
        show_column_names = FALSE, show_row_names = TRUE,
        show_row_dend = FALSE, show_column_dend = FALSE,
        row_names_gp = gpar(fontsize = 8),
        top_annotation = colAnn, 
        col = colHm,
        right_annotation = rowAnn,
        row_km = 2, row_gap = unit(1, "mm"), show_parent_dend_line = FALSE,
        column_km = 2, column_gap = unit(1, "mm"),
        border_gp = gpar(col = "black", lwd = 0.5),
        heatmap_legend_param = list(direction = "vertical"))
        
draw(p, padding = unit(c(0, 0, 0, 0), "cm"), merge_legend = TRUE, heatmap_legend_side = "left", newpage = FALSE)
popViewport()
invisible(dev.off()) 
```

```{r}
# Serum
homo_vs_het_da_features_serum <- read.csv("HomoCtrl_vs_HeteroCtrl_da_features - serum.csv") %>% 
  filter(Relevant == "TRUE") %>%
  mutate(Rename = gsub("*", "", Rename, fixed = TRUE))
dim(homo_vs_het_da_features_serum)[1] # 10
m <- metadata_serum %>% filter(Class != "Homo-ALA") %>% 
  mutate(Class = as.character(Class),
         Class = ifelse(Class == "Hetero-Ctrl", "Heterozygote", "Homozygote"),
         Class = factor(Class, levels = c("Heterozygote", "Homozygote")))

data_serum <- as.data.frame(limma_list$serum$HomoCtrl_vs_HeteroCtrl$heat_sig_hits) %>% 
  rownames_to_column("Molecule") %>%
  right_join(homo_vs_het_da_features_serum) %>%
  dplyr::select(Rename, m$SampleNumber) %>%
  column_to_rownames("Rename") %>%
  as.matrix()

colAnn <- HeatmapAnnotation(Group = m$Class, 
                            col = list(Group = c("Homozygote" = "#7DB8F3", "Heterozygote" = "grey")), 
                            annotation_name_side = "right") 
identical(rownames(data_serum), homo_vs_het_da_features_serum$Rename)

rowAnn <- rowAnnotation(Metabolism = homo_vs_het_da_features_serum$Metabolism,
                        col = list(Metabolism = metabolism_palette))

tiff(paste0(drive_path, "supplementary/homo-het-serum-heatmap.tiff"), width = 16, height = 8, units = "cm", res = 600)
pushViewport(viewport(gp = gpar(fontfamily = "Arial", fontsize = 8)))
p <- 
      Heatmap(data_serum, 
        name = "z-score",
        column_title = "Serum",
        column_title_gp = gpar(fontface = "bold"),
        row_title = "Metabolite",  
        show_column_names = FALSE, show_row_names = TRUE,
        show_row_dend = FALSE, show_column_dend = FALSE,
        row_names_gp = gpar(fontsize = 8),
        top_annotation = colAnn, 
        col = colHm,
        right_annotation = rowAnn,
        row_km = 2, row_gap = unit(1, "mm"), show_parent_dend_line = FALSE,
        column_km = 2, column_gap = unit(1, "mm"),
        border_gp = gpar(col = "black", lwd = 0.5),
        heatmap_legend_param = list(direction = "vertical"))
        
draw(p, padding = unit(c(0, 0, 0, 0), "cm"), merge_legend = TRUE, heatmap_legend_side = "left", newpage = FALSE)
popViewport()
invisible(dev.off()) 
```

```{r}
# Row annotation only
u <- unique(c(homo_vs_het_da_features_serum$Metabolism, homo_vs_het_da_features_urine$Metabolism))
m <- c("Arginine" = "#5F72A1", 
       "Cysteine and methionine" = "#1c97ba",
       "Histidine" = "#083D77",
       "Lysine and carnitine" = "#995de3",
       "Nucleotide" = "#209179",
       "Tryptophan" = "#A8DCD1", 
       "Vitamin, flavonoid and folate" = "#89e2fa", 
       "Other" = "grey")

lgd <- Legend(at = names(m), 
              title = "Metabolism", 
              legend_gp = gpar(fill = m))

tiff(paste0(drive_path, "supplementary/homo-het-heatmap-legend.tiff"), width = 8, height = 8, units = "cm", res = 600)
draw(lgd, x = unit(1, "npc") - unit(1, "cm"), y = unit(1, "npc") - unit(1, "cm"), just = c("right", "top"))
invisible(dev.off()) 
```



### Overlapping metabolites

```{r}
# Common to serum and urine
length(intersect(limma_list$serum$HomoCtrl_vs_HeteroCtrl$sig_hits$Molecule, limma_list$urine$HomoCtrl_vs_HeteroCtrl$sig_hits$Molecule) ) # 14
```

```{r}
# Venn diagram
l <- list("Urine" = features %>% filter(Dataset == "Urine", Contrast == "HomoCtrl_vs_HeteroCtrl") %>% pull(Rename), # checked same as Molecule
          "Serum" = features %>% filter(Dataset == "Serum", Contrast == "HomoCtrl_vs_HeteroCtrl") %>% pull(Rename))

p <- 
  ggVennDiagram(l, label_alpha = 0, label_size = 6, label = "count", set_size = 6, set_name_position = c(0.5, 1.5)) + 
    theme(plot.title = element_text(hjust = 0.5, vjust = 1, face = "bold"), 
          legend.position = "none") +
    labs(title = "Homozygote vs Heterozygote\nDA molecules overlap") +
    scale_fill_gradient(low = "white", high = "lightgrey") +
    scale_x_continuous(expand = expansion(mult = c(0.4, 0.4))) +
    scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) #Mol 36 | 14 | 181 same as rename

ggsave(paste0(drive_path, "figure3/homo-het-venn.tiff"), p, width = 10, height = 6, units = 'cm', dpi = 600)  

l_p <- process_region_data(Venn(l))$item 
names(l_p) <- process_region_data(Venn(l))$name
```

#### Boxplots

```{r}
# Boxplots
data_urine <- metab_ur_df %>% 
  rownames_to_column("Molecule") %>% 
  full_join(unique_features[, c("Molecule", "Rename")]) %>%
  filter(Rename %in% unique(c(l_p$`Urine/Serum`))) %>% 
  #mutate(Molecule = ifelse(Molecule == "2-Hydroxybutyric acid", "4-Hydroxybutyric acid", Molecule)) %>%
  dplyr::select(!Molecule) %>%
  pivot_longer(!Rename, names_to = "SampleNumber", values_to = "Intensity") %>%
  right_join(metadata_urine[, c("SampleNumber", "Class")]) %>%
  filter(Class != "Homo-ALA") %>% 
  mutate(SampleType = "Urine")

data_serum <- metab_se_df %>% 
  rownames_to_column("Molecule") %>% 
  full_join(unique_features[, c("Molecule", "Rename")]) %>%
  filter(Rename %in% unique(c(l_p$`Urine/Serum`))) %>% 
  #mutate(Molecule = ifelse(Molecule == "2-Hydroxybutyric acid", "4-Hydroxybutyric acid", Molecule)) %>%
  dplyr::select(!Molecule) %>%
  pivot_longer(!Rename, names_to = "SampleNumber", values_to = "Intensity") %>%
  right_join(metadata_serum[, c("SampleNumber", "Class")]) %>%
  filter(Class != "Homo-ALA") %>% 
  mutate(SampleType = "Serum")

data <- rbind(data_urine, data_serum) %>% 
  mutate(Class = as.character(Class),
         Class = ifelse(Class == "Hetero-Ctrl", "Heterozygote", "Homozygote"),
         Class = factor(Class, levels = c("Heterozygote", "Homozygote")))

for (i in 1:length(unique(c(l_p$`Urine/Serum`)))){
  
  title <- unique(c(l_p$`Urine/Serum`))[i]
  test <- "t.test"
  df <- data %>% filter(Rename %in% title) 
  
  ## Serum
  d <- df %>% filter(SampleType == "Serum")
    
  p1 <- 
    ggplot(d, aes(x = Class, y = Intensity)) +
      geom_boxplot(aes(fill = Class), outlier.shape = NA) +
      stat_compare_means(label = "p.signif", method = test, paired = FALSE, label.x.npc = "centre", label.y = max(d$Intensity)-0.02, size = 3) + 
      geom_jitter(aes(fill = Class), width = 0.1, shape = 21) +
      theme +
      theme(legend.position = "none") +
      labs(fill = "Group", x = "Group") +
      facet_grid(~SampleType) +
      scale_fill_manual(values = c("Homozygote" = "#7DB8F3", "Heterozygote" = "grey")) +
      scale_y_continuous(limits = c(min(d$Intensity)-1, max(d$Intensity)+1))
  
  ## Urine
  d <- df %>% filter(SampleType == "Urine")
    
  p2 <- 
    ggplot(d, aes(x = Class, y = Intensity)) +
      geom_boxplot(aes(fill = Class), outlier.shape = NA) +
      stat_compare_means(label = "p.signif", method = test, paired = FALSE, label.x.npc = "centre", label.y = max(d$Intensity)-0.02, size = 3) + 
      geom_jitter(aes(fill = Class), width = 0.1, shape = 21) +
      theme +
      theme(legend.position = "none") +
      labs(fill = "Group", x = "Group") +
      facet_grid(~SampleType) +
      scale_fill_manual(values = c("Homozygote" = "#7DB8F3", "Heterozygote" = "grey")) +
      scale_y_continuous(limits = c(min(d$Intensity)-1, max(d$Intensity)+1))
  
  p <- p1 + p2 + plot_annotation(title = title, 
                           theme = theme(title = element_text(face = "bold", size = 8), 
                                         plot.subtitle = element_blank(),
                                         plot.margin = margin(t = 0, r = 0, b = 0, l = 0)))
  
  ggsave(paste0(drive_path, "figure3/boxplots/homo-het-", gsub("/", "-", title, fixed = TRUE), "-", test, ".tiff"), p, width = 9, height = 5, units = 'cm', dpi = 600)  
}
```

```{r}
extra_metabolites <- c("(-)-Epigallocatechin", "1-Pyrroline-5-carboxylic acid", "Homocysteine-cysteine disulfide", "Ascorbic acid", "Bicarbonate carbon", "Epicatechin", "Ferulic acid", "Formononetin", "Homocysteine thiolactone", "Indole-3-acetylglycine", "L-Cystathionine", "Pelargonidin 3-glucoside", "Phenol sulphate", "Diaminopimelic acid")

# Boxplots
data_urine <- metab_ur_df %>% 
  rownames_to_column("Molecule") %>% 
  full_join(unique_features[, c("Molecule", "Rename")]) %>%
  filter(Rename %in% extra_metabolites) %>%
  dplyr::select(!Molecule) %>%
  pivot_longer(!Rename, names_to = "SampleNumber", values_to = "Intensity") %>%
  right_join(metadata_urine[, c("SampleNumber", "Class")]) %>%
  filter(Class != "Homo-ALA") %>% 
  mutate(SampleType = "Urine")

data_serum <- metab_se_df %>% 
  rownames_to_column("Molecule") %>% 
  full_join(unique_features[, c("Molecule", "Rename")]) %>%
  filter(Rename %in% extra_metabolites) %>%
  dplyr::select(!Molecule) %>%
  pivot_longer(!Rename, names_to = "SampleNumber", values_to = "Intensity") %>%
  right_join(metadata_serum[, c("SampleNumber", "Class")]) %>%
  filter(Class != "Homo-ALA") %>% 
  mutate(SampleType = "Serum")

data <- rbind(data_urine, data_serum) %>%
  mutate(Class = as.character(Class),
         Class = ifelse(Class == "Hetero-Ctrl", "Heterozygote", "Homozygote"),
         Class = factor(Class, levels = c("Heterozygote", "Homozygote")))

for (i in 1:length(extra_metabolites)){
  
  title <- extra_metabolites[i]
  test <- "t.test"
  df <- data %>% filter(Rename %in% title) 
  
  ## Serum
  d <- df %>% filter(SampleType == "Serum")
    
  p1 <- 
    ggplot(d, aes(x = Class, y = Intensity)) +
      geom_boxplot(aes(fill = Class), outlier.shape = NA) +
      stat_compare_means(label = "p.signif", method = test, paired = FALSE, label.x.npc = "centre", label.y = max(d$Intensity)-0.02, size = 3) + 
      geom_jitter(aes(fill = Class), width = 0.1, shape = 21) +
      theme +
      theme(legend.position = "none") +
      labs(fill = "Group", x = "Group") +
      facet_grid(~SampleType) +
      scale_fill_manual(values = c("Homozygote" = "#7DB8F3", "Heterozygote" = "grey")) +
      scale_y_continuous(limits = c(min(d$Intensity)-1, max(d$Intensity)+1))
  
  ## Urine
  d <- df %>% filter(SampleType == "Urine")
    
  p2 <- 
    ggplot(d, aes(x = Class, y = Intensity)) +
      geom_boxplot(aes(fill = Class), outlier.shape = NA) +
      stat_compare_means(label = "p.signif", method = test, paired = FALSE, label.x.npc = "centre", label.y = max(d$Intensity)-0.02, size = 3) + 
      geom_jitter(aes(fill = Class), width = 0.1, shape = 21) +
      theme +
      theme(legend.position = "none") +
      labs(fill = "Group", x = "Group") +
      facet_grid(~SampleType) +
      scale_fill_manual(values = c("Homozygote" = "#7DB8F3", "Heterozygote" = "grey")) +
      scale_y_continuous(limits = c(min(d$Intensity)-1, max(d$Intensity)+1))
  
  p <- p1 + p2 + plot_annotation(title = title, 
                           theme = theme(title = element_text(face = "bold", size = 8), 
                                         plot.subtitle = element_blank(),
                                         plot.margin = margin(t = 0, r = 0, b = 0, l = 0)))
  
  ggsave(paste0(drive_path, "figure3/boxplots/selected/homo-het-", gsub("/", "-", title, fixed = TRUE), "-", test, ".tiff"), p, width = 9, height = 5, units = 'cm', dpi = 600)  
}
```


# ---------------------
## **Homo-ALA & Homo-Ctrl 
### PCA

```{r}
# Urine
m <- metadata_urine %>% filter(Class != "Hetero-Ctrl")
d <- metab_ur_df[, m$SampleNumber]

set.seed(3)
a <- data.frame(adonis2(formula = as.formula(paste('t(d) ~', "Class")), 
                      data = m, 
                      distance = "eu", 
                      permutations = 999,
                      na.action = na.omit,
                      parallel = 6))

annotation <- paste(paste("SumOfSqs =", round(a["Class",]$SumOfSqs, 3)),
              paste("R2 =", round(a["Class",]$R2, 3)),
              paste("p =", round(a["Class",]$Pr..F., 3)), sep = " ; " )

pca <- pca_process(data = d, metadata = m)

p <- 

  ggplot(pca$data, aes(x = PC1, y = PC2)) +
  ggforce::geom_mark_ellipse(aes(fill = Class), expand = unit(2.5, "mm"), colour = NA, alpha = 0.2) +
  geom_point(aes(fill = Class), shape = 21) +
  scale_fill_manual(values = group_palette) +
  theme +
  theme(legend.position = "none") +
  labs(title = "Urine metabolomics",
       caption = annotation,
       fill = "Group",
       x = paste0('PC1 ', round(pca$varexp[1], 2), '%'),
       y = paste0('PC2 ', round(pca$varexp[2], 2), '%')) +
  scale_y_continuous(limits = c(-45, 45)) +
  scale_x_continuous(limits = c(-50, 60))

ggsave(paste0(drive_path, "figure7/ala-homo-urine-pca.tiff"), p, width = 2, height = 2, dpi = 600)
```

```{r}
# Serum
m <- metadata_serum %>% filter(Class != "Hetero-Ctrl")
d <- metab_se_df[, m$SampleNumber]

set.seed(3)
a <- data.frame(adonis2(formula = as.formula(paste('t(d) ~', "Class")), 
                      data = m, 
                      distance = "eu", 
                      permutations = 999,
                      na.action = na.omit,
                      parallel = 6))

annotation <- paste(paste("SumOfSqs =", round(a["Class",]$SumOfSqs, 3)),
              paste("R2 =", round(a["Class",]$R2, 3)),
              paste("p =", round(a["Class",]$Pr..F., 3)), sep = " ; " )

pca <- pca_process(data = d, metadata = m)

p <- 

  ggplot(pca$data, aes(x = PC1, y = PC2)) +
  ggforce::geom_mark_ellipse(aes(fill = Class), expand = unit(2.5, "mm"), colour = NA, alpha = 0.2) +
  geom_point(aes(fill = Class), shape = 21) +
  scale_fill_manual(values = group_palette) +
  theme +
  theme(legend.position = "none") +
  labs(title = "Serum metabolomics",
       caption = annotation,
       fill = "Group",
       x = paste0('PC1 ', round(pca$varexp[1], 2), '%'),
       y = paste0('PC2 ', round(pca$varexp[2], 2), '%')) +
  scale_y_continuous(limits = c(-35, 30)) +
  scale_x_continuous(limits = c(-30, 30))

ggsave(paste0(drive_path, "figure7/ala-homo-serum-pca.tiff"), p, width = 2, height = 2, dpi = 600)

# Legend
#pca$data <- pca$data %>%
#  mutate(Class = ifelse(Class == "Homo-Ctrl", "Ctrl", "ALA"))

p <- 

  ggplot(pca$data, aes(x = PC1, y = PC2)) +
  geom_point(aes(fill = Class), shape = 21) +
  ggforce::geom_mark_ellipse(aes(fill = Class), expand = unit(2.5, "mm"), colour = NA, alpha = 0.2) +
  #scale_fill_manual(values = c("Ctrl" = "#7DB8F3", "ALA" = "#7D83F8")) +
  scale_fill_manual(values = group_palette) +
  theme +
  labs(fill = "Group") +
  theme(legend.position = "bottom")
ggsave(paste0(drive_path, "figure7/ala-homo-pca-legend2.tiff"), p, width = 3, height = 2, dpi = 600)
```

### Barplot

```{r}
u <- limma_list$urine$HomoALA_vs_HomoCtrl$sig_hits %>%
  full_join(read.csv("HomoALA_vs_HomoCtrl_da_features - urine.csv")[, c("Molecule", "Rename", "Metabolism", "Relevant")] ) %>%
  mutate(Dataset = "Urine",
         Rename = ifelse(Rename == "", Molecule, Rename),
         Rename = gsub("*", "", Rename, fixed = TRUE)) %>%
  dplyr::select(!Molecule) %>%
  rename(Rename = "Molecule", direction = "Direction") %>%
  mutate(Direction = gsub("HomoALA", "Increased in Homo-ALA", Direction),
         Direction = gsub("HomoCtrl", "Decreased in Homo-ALA", Direction)) %>%
  relocate(Molecule) #%>%
  #mutate(Molecule = ifelse(Molecule == "2-Hydroxybutyric acid", "4-Hydroxybutyric acid", Molecule))
dim(u)[1] #563

write.csv(u %>% dplyr::select(!Relevant), 
          paste0(drive_path, "tables/Suppl_table_HomoALA_vs_HomoCtrl_da_features_urine.csv"), row.names = FALSE )

s <- limma_list$serum$HomoALA_vs_HomoCtrl$sig_hits %>%
  full_join(read.csv("HomoALA_vs_HomoCtrl_da_features - serum.csv")[, c("Molecule", "Rename", "Metabolism", "Relevant")] ) %>%
  mutate(Dataset = "Serum",
         Rename = ifelse(Rename == "", Molecule, Rename),
         Rename = gsub("*", "", Rename, fixed = TRUE)) %>%
  dplyr::select(!Molecule) %>%
  rename(Rename = "Molecule", direction = "Direction") %>%
  mutate(Direction = gsub("HomoALA", "Increased in Homo-ALA", Direction),
         Direction = gsub("HomoCtrl", "Decreased in Homo-ALA", Direction)) %>%
  relocate(Molecule) #%>%
  #mutate(Molecule = ifelse(Molecule == "2-Hydroxybutyric acid", "4-Hydroxybutyric acid", Molecule))
dim(s)[1] # 114

write.csv(s %>% dplyr::select(!Relevant), 
          paste0(drive_path, "tables/Suppl_table_HomoALA_vs_HomoCtrl_da_features_serum.csv"), row.names = FALSE )


data <- rbind(u, s) %>%
  filter(Relevant == TRUE) %>%
  mutate(Significance = case_when(adj.P.Val >= 0.05 ~ "",
                                      adj.P.Val < 0.05 & adj.P.Val >= 0.01 ~ "*",
                                      adj.P.Val < 0.01 & adj.P.Val >= 0.001 ~ "**",
                                      adj.P.Val < 0.001 & adj.P.Val >= 0.0001 ~ "***",
                                      adj.P.Val < 0.0001 ~ "****") ) %>%
  mutate(Metabolism = gsub("and", "\nand", Metabolism),
         Metabolism = gsub(",", ",\n", Metabolism),
         Metabolism = factor(Metabolism, levels = c("Arginine", "Cysteine \nand methionine","Histidine",
                                                "Lysine \nand carnitine","Tryptophan","Other",
                                                "Nucleotide", "Vitamin,\n flavonoid \nand folate","Glycolysis"))) %>%
  group_by(Metabolism) %>%
  arrange(logFC) %>%
  mutate(Molecule = factor(Molecule, levels = unique(.$Molecule)))
dim(data)[1] # 98 = 68 + 30

p <- 
  ggplot(data, aes(x = logFC, y = Molecule)) +
    geom_col(aes(fill = Direction), colour = "black", size = 0.2) +
    geom_text(aes(label = Significance), nudge_x = 0.1, nudge_y = -0.2, size = 5) +
    theme +
    theme(strip.text.y = element_text(angle = 0),
          legend.position = "bottom") +
    geom_vline(xintercept = 0, lty = "dashed") +
    labs(x = "Log2FC", fill = "", y = "", title = "Differentially abundant metabolites") +
    scale_fill_manual(values = c("Increased in Homo-ALA" = "#7D83F8", "Decreased in Homo-ALA" = "#CECECE")) +
    facet_grid(Metabolism~Dataset, scales = "free_y", space = "free_y")
ggsave(paste0(drive_path, "figure7/ala-homo-urine-serum-da-list.tiff"), p, width = 7, height = 12, dpi = 600)
```


### Heatmap

```{r}
# Urine 
m <- metadata_urine %>% filter(Class != "Hetero-Ctrl") #%>%
  #mutate(Class = as.character(Class),
  #       Class = ifelse(Class == "Homo-ALA", "ALA", "Ctrl"),
  #       Class = factor(Class, levels = c("Ctrl", "ALA")))

ala_vs_homo_da_features_urine <- read.csv("HomoALA_vs_HomoCtrl_da_features - urine.csv") %>% 
  filter(Relevant == "TRUE") %>%
  mutate(Rename = gsub("*", "", Rename, fixed = TRUE))
dim(ala_vs_homo_da_features_urine)[1] # 68

data_urine <- as.data.frame(limma_list$urine$HomoALA_vs_HomoCtrl$heat_sig_hits) %>% 
  rownames_to_column("Molecule") %>%
  right_join(ala_vs_homo_da_features_urine) %>%
  dplyr::select(Rename, m$SampleNumber) %>%
  column_to_rownames("Rename") %>%
  as.matrix()
data_urine <- data_urine[ala_vs_homo_da_features_urine$Rename, ]
identical(rownames(data_urine), ala_vs_homo_da_features_urine$Rename)  #TRUE

colAnn <- HeatmapAnnotation(Group = m$Class, 
                            col = list(Group = c("Homo-Ctrl" = "#7DB8F3", "Homo-ALA" = "#7D83F8")), 
                            annotation_name_side = "right") 

rowAnn <- rowAnnotation(Metabolism = ala_vs_homo_da_features_urine$Metabolism,
                        col = list(Metabolism = metabolism_palette))

tiff(paste0(drive_path, "supplementary/ala-homo-urine-heatmap.tiff"), width = 18, height = 20, units = "cm", res = 600)
pushViewport(viewport(gp = gpar(fontfamily = "Arial", fontsize = 8)))
p <- 
      Heatmap(data_urine, 
        name = "z-score",
        column_title = "Urine",
        column_title_gp = gpar(fontface = "bold"),
        row_title = "Metabolite",  
        show_column_names = FALSE, show_row_names = TRUE,
        show_row_dend = FALSE, show_column_dend = FALSE,
        row_names_gp = gpar(fontsize = 8),
        top_annotation = colAnn, 
        col = colHm,
        right_annotation = rowAnn,
        row_km = 2, row_gap = unit(1, "mm"), show_parent_dend_line = FALSE,
        column_km = 2, column_gap = unit(1, "mm"),
        border_gp = gpar(col = "black", lwd = 0.5),
        heatmap_legend_param = list(direction = "vertical"))
        
draw(p, padding = unit(c(0, 0, 0, 0), "cm"), merge_legend = TRUE, heatmap_legend_side = "left", newpage = FALSE)
popViewport()
invisible(dev.off()) 
```

```{r}
# Serum 
m <- metadata_serum %>% filter(Class != "Hetero-Ctrl") #%>%
  #mutate(Class = as.character(Class),
  #       Class = ifelse(Class == "Homo-ALA", "ALA", "Ctrl"),
  #       Class = factor(Class, levels = c("Ctrl", "ALA")))

ala_vs_homo_da_features_serum <- read.csv("HomoALA_vs_HomoCtrl_da_features - serum.csv") %>% 
  filter(Relevant == "TRUE") %>%
  mutate(Rename = gsub("*", "", Rename, fixed = TRUE))
dim(ala_vs_homo_da_features_serum)[1] # 30

data_serum <- as.data.frame(limma_list$serum$HomoALA_vs_HomoCtrl$heat_sig_hits) %>% 
  rownames_to_column("Molecule") %>%
  right_join(ala_vs_homo_da_features_serum) %>%
  dplyr::select(Rename, m$SampleNumber) %>%
  #mutate(Rename = ifelse(Rename == "2-Hydroxybutyric acid", "4-Hydroxybutyric acid", Rename)) %>%
  column_to_rownames("Rename") %>%
  as.matrix()

colAnn <- HeatmapAnnotation(Group = m$Class, 
                            col = list(Group = c("Homo-Ctrl" = "#7DB8F3", "Homo-ALA" = "#7D83F8")), 
                            annotation_name_side = "right") 
identical(rownames(data_serum), ala_vs_homo_da_features_serum$Rename) #TRUE

rowAnn <- rowAnnotation(Metabolism = ala_vs_homo_da_features_serum$Metabolism,
                        col = list(Metabolism = metabolism_palette))

tiff(paste0(drive_path, "supplementary/ala-homo-serum-heatmap.tiff"), width = 18, height = 15, units = "cm", res = 600)

pushViewport(viewport(gp = gpar(fontfamily = "Arial", fontsize = 8)))
p <- 
      Heatmap(data_serum, 
        name = "z-score",
        column_title = "Serum",
        column_title_gp = gpar(fontface = "bold"),
        show_row_dend = FALSE, show_column_dend = FALSE,
        row_title = "Metabolites",  
        show_column_names = FALSE, show_row_names = TRUE,
        row_names_gp = gpar(fontsize = 8),
        top_annotation = colAnn, 
        col = colHm,
        right_annotation = rowAnn,
        row_km = 2, row_gap = unit(1, "mm"), show_parent_dend_line = FALSE,
        column_km = 2, column_gap = unit(1, "mm"),
        border_gp = gpar(col = "black", lwd = 0.5),
        heatmap_legend_param = list(direction = "vertical"))
        
draw(p, padding = unit(c(0, 0, 0, 0), "cm"), merge_legend = TRUE, heatmap_legend_side = "left", newpage = FALSE)
popViewport()

invisible(dev.off()) 
```

### Overlapping metabolites

```{r}
# Common to serum and urine
length(intersect(limma_list$serum$HomoALA_vs_HomoCtrl$sig_hits$Molecule, limma_list$urine$HomoALA_vs_HomoCtrl$sig_hits$Molecule) ) # 24

read.csv("HomoALA_vs_HomoCtrl_da_features - urine.csv")[, c("Molecule", "direction")] %>%
  inner_join(read.csv("HomoALA_vs_HomoCtrl_da_features - serum.csv")[, c("Molecule", "direction")], by = "Molecule", suffix = c("_urine", "_serum") ) %>%
  mutate(Different = ifelse(direction_urine != direction_serum, TRUE, FALSE)) %>%
  write.csv(., "HomoALA_vs_HomoCtrl_molecule_overlap.csv")
```

```{r}
# Venn diagram
l <- list("Urine" = features %>% filter(Dataset == "Urine", Contrast == "HomoALA_vs_HomoCtrl") %>% pull(Rename), # checked same as Molecule
          "Serum" = features %>% filter(Dataset == "Serum", Contrast == "HomoALA_vs_HomoCtrl") %>% pull(Rename))

p <- 
  ggVennDiagram(l, label_alpha = 0, label_size = 6, label = "count", set_size = 6, set_name_position = c(0.5, 1.5)) + 
    theme(plot.title = element_text(hjust = 0.5, vjust = 1, face = "bold"), 
          legend.position = "none") +
    labs(title = "ALA Homozygote vs Homozygote Control\nDA molecules overlap") +
    scale_fill_gradient(low = "white", high = "lightgrey") +
    scale_x_continuous(expand = expansion(mult = c(0.4, 0.4))) +
    scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) #90 | 24 | 539 same 

ggsave(paste0(drive_path, "figure7/ala-homo-venn.tiff"), p, width = 10, height = 6, units = 'cm', dpi = 600)  

l_p <- process_region_data(Venn(l))$item 
names(l_p) <- process_region_data(Venn(l))$name
```

#### Boxplots

```{r}
# Boxplots
data_urine <- metab_ur_df %>% 
  rownames_to_column("Molecule") %>% 
  full_join(unique_features[, c("Molecule", "Rename")]) %>%
  #filter(Rename %in% extra_metabolites) %>%
  filter(Rename %in% unique(c(l_p$`Urine/Serum`))) %>% 
  dplyr::select(!Molecule) %>%
  pivot_longer(!Rename, names_to = "SampleNumber", values_to = "Intensity") %>%
  right_join(metadata_urine[, c("SampleNumber", "Class")]) %>%
  filter(Class != "Hetero-Ctrl") %>% 
  mutate(SampleType = "Urine")

data_serum <- metab_se_df %>% 
  rownames_to_column("Molecule") %>% 
  full_join(unique_features[, c("Molecule", "Rename")]) %>%
  #filter(Rename %in% extra_metabolites) %>%
  filter(Rename %in% unique(c(l_p$`Urine/Serum`))) %>% 
  dplyr::select(!Molecule) %>%
  pivot_longer(!Rename, names_to = "SampleNumber", values_to = "Intensity") %>%
  right_join(metadata_serum[, c("SampleNumber", "Class")]) %>%
  filter(Class != "Hetero-Ctrl") %>% 
  mutate(SampleType = "Serum")

data <- rbind(data_urine, data_serum) #%>% 
  #mutate(Class = as.character(Class),
  #       Class = ifelse(Class == "Homo-ALA", "ALA", "Ctrl"),
  #       Class = factor(Class, levels = c("Ctrl", "ALA")))

for (i in 1:length(unique(c(l_p$`Urine/Serum`)))){
  
  title <- unique(c(l_p$`Urine/Serum`))[i]
  test <- "t.test"
  df <- data %>% filter(Rename %in% title) 
  
  ## Serum
  d <- df %>% filter(SampleType == "Serum")
    
  p1 <- 
    ggplot(d, aes(x = Class, y = Intensity)) +
      geom_boxplot(aes(fill = Class), outlier.shape = NA) +
      stat_compare_means(label = "p.signif", method = test, paired = FALSE, label.x.npc = "centre", label.y = max(d$Intensity)-0.02, size = 3) + 
      geom_jitter(aes(fill = Class), width = 0.1, shape = 21) +
      theme +
      theme(legend.position = "none") +
      labs(fill = "Group", x = "Group") +
      facet_grid(~SampleType) +
      scale_fill_manual(values = c("Homo-Ctrl" = "#7DB8F3", "Homo-ALA" = "#7D83F8")) +
      scale_y_continuous(limits = c(min(d$Intensity)-1, max(d$Intensity)+1))
  
  ## Urine
  d <- df %>% filter(SampleType == "Urine")
    
  p2 <- 
    ggplot(d, aes(x = Class, y = Intensity)) +
      geom_boxplot(aes(fill = Class), outlier.shape = NA) +
      stat_compare_means(label = "p.signif", method = test, paired = FALSE, label.x.npc = "centre", label.y = max(d$Intensity)-0.02, size = 3) + 
      geom_jitter(aes(fill = Class), width = 0.1, shape = 21) +
      theme +
      theme(legend.position = "none") +
      labs(fill = "Group", x = "Group") +
      facet_grid(~SampleType) +
      scale_fill_manual(values = c("Homo-Ctrl" = "#7DB8F3", "Homo-ALA" = "#7D83F8")) +
      scale_y_continuous(limits = c(min(d$Intensity)-1, max(d$Intensity)+1))
  
  p <- p1 + p2 + plot_annotation(title = title, 
                           theme = theme(title = element_text(face = "bold", size = 8), 
                                         plot.subtitle = element_blank(),
                                         plot.margin = margin(t = 0, r = 0, b = 0, l = 0)))
  
  ggsave(paste0(drive_path, "figure7/boxplots/ala-ctrl-", gsub("/", "-", title, fixed = TRUE), "-", test, ".tiff"), p, width = 9, height = 5, units = 'cm', dpi = 600)  
}
```

```{r}
# Boxplots
extra_metabolites <- c("(-)-Epigallocatechin", "1-Pyrroline-5-carboxylic acid", "Homocysteine-cysteine disulfide", "Ascorbic acid", "Bicarbonate carbon", "Epicatechin", "Ferulic acid", "Formononetin", "Homocysteine thiolactone", "Indole-3-acetylglycine", "L-Cystathionine", "Pelargonidin 3-glucoside", "Phenol sulphate", "Diaminopimelic acid", "Pomiferin")

data_urine <- metab_ur_df %>% 
  rownames_to_column("Molecule") %>% 
  full_join(unique_features[, c("Molecule", "Rename")]) %>%
  filter(Rename %in% extra_metabolites) %>%
  dplyr::select(!Molecule) %>%
  pivot_longer(!Rename, names_to = "SampleNumber", values_to = "Intensity") %>%
  right_join(metadata_urine[, c("SampleNumber", "Class")]) %>%
  filter(Class != "Hetero-Ctrl") %>% 
  mutate(SampleType = "Urine")

data_serum <- metab_se_df %>% 
  rownames_to_column("Molecule") %>% 
  full_join(unique_features[, c("Molecule", "Rename")]) %>%
  filter(Rename %in% extra_metabolites) %>%
  dplyr::select(!Molecule) %>%
  pivot_longer(!Rename, names_to = "SampleNumber", values_to = "Intensity") %>%
  right_join(metadata_serum[, c("SampleNumber", "Class")]) %>%
  filter(Class != "Hetero-Ctrl") %>% 
  mutate(SampleType = "Serum")

data <- rbind(data_urine, data_serum) #%>% 
  #mutate(Class = as.character(Class),
  #       Class = ifelse(Class == "Homo-ALA", "ALA", "Ctrl"),
  #       Class = factor(Class, levels = c("Ctrl", "ALA")))

for (i in 1:length(extra_metabolites)){
  
  title <- extra_metabolites[i]
  test <- "t.test"
  df <- data %>% filter(Rename %in% title) 
  
  ## Serum
  d <- df %>% filter(SampleType == "Serum")
    
  p1 <- 
    ggplot(d, aes(x = Class, y = Intensity)) +
      geom_boxplot(aes(fill = Class), outlier.shape = NA) +
      stat_compare_means(label = "p.signif", method = test, paired = FALSE, label.x.npc = "centre", label.y = max(d$Intensity)-0.02, size = 3) + 
      geom_jitter(aes(fill = Class), width = 0.1, shape = 21) +
      theme +
      theme(legend.position = "none") +
      labs(fill = "Group", x = "Group") +
      facet_grid(~SampleType) +
      scale_fill_manual(values = c("Homo-Ctrl" = "#7DB8F3", "Homo-ALA" = "#7D83F8")) +
      scale_y_continuous(limits = c(min(d$Intensity)-1, max(d$Intensity)+1))
  
  ## Urine
  d <- df %>% filter(SampleType == "Urine")
    
  p2 <- 
    ggplot(d, aes(x = Class, y = Intensity)) +
      geom_boxplot(aes(fill = Class), outlier.shape = NA) +
      stat_compare_means(label = "p.signif", method = test, paired = FALSE, label.x.npc = "centre", label.y = max(d$Intensity)-0.02, size = 3) + 
      geom_jitter(aes(fill = Class), width = 0.1, shape = 21) +
      theme +
      theme(legend.position = "none") +
      labs(fill = "Group", x = "Group") +
      facet_grid(~SampleType) +
      scale_fill_manual(values = c("Homo-Ctrl" = "#7DB8F3", "Homo-ALA" = "#7D83F8")) +
      scale_y_continuous(limits = c(min(d$Intensity)-1, max(d$Intensity)+1))
  
  p <- p1 + p2 + plot_annotation(title = title, 
                           theme = theme(title = element_text(face = "bold", size = 8), 
                                         plot.subtitle = element_blank(),
                                         plot.margin = margin(t = 0, r = 0, b = 0, l = 0)))
  
  ggsave(paste0(drive_path, "figure7/boxplots/selected/ala-ctrl-", gsub("/", "-", title, fixed = TRUE), "-", test, ".tiff"), p, width = 9, height = 5, units = 'cm', dpi = 600)  
}
```


# ------------------------
# Stone correlations

```{r}
correlation_list <- list()

# Urine
df1 <- metadata_urine %>% 
  dplyr::select(SampleNumber, Total_stone_volume, Stone_number, Stone_surface_area) %>% column_to_rownames("SampleNumber")
df2 <- metab_ur_df %>% filter(rownames(.) %in% unique_features$Molecule)

correlation_list[["Urine"]] <- correlate_datasets(df1 = df1, 
                     df2 = df2, 
                     df1_type = "Urine", df2_type = "Molecule", 
                     plots = FALSE, 
                     coef_threshold = 0, correlation = "pearson", pval_threshold = 1)

# Serum
df1 <- metadata_serum %>% 
  dplyr::select(SampleNumber, Total_stone_volume, Stone_number, Stone_surface_area) %>% column_to_rownames("SampleNumber")
df2 <- metab_se_df %>% filter(rownames(.) %in% unique_features$Molecule)
correlation_list[["Serum"]] <- correlate_datasets(df1 = df1, 
                     df2 = df2, 
                     df1_type = "Serum", df2_type = "Molecule", 
                     plots = FALSE, 
                     coef_threshold = 0, correlation = "pearson", pval_threshold = 1)
```

```{r}
a <- correlation_list$Urine$coef_df %>% 
  dplyr::rename(Molecule = df2_name) 

b <- correlation_list$Serum$coef_df %>% 
  dplyr::rename(Molecule = df2_name) 

corr_table <- rbind(a, b) %>% 
  mutate(df1_name = gsub("Total_stone_volume", "Volume (mm\u00B3)", df1_name),
         df1_name = gsub("Stone_number", "Number", df1_name),
         df1_name = gsub("Stone_surface_area", "Surface area (mm\u00B2)", df1_name),
         Correlation2 = ifelse(pval >= 0.05, 0, Correlation),
         Sign = case_when(Correlation2 > 0 ~ "+", 
                          Correlation2 == 0 ~ "", 
                          Correlation2 < 0 ~ "-"), 
         Significance = case_when(pval >= 0.05 ~ "",
                                      pval < 0.05 & pval >= 0.01 ~ "*",
                                      pval < 0.01 & pval >= 0.001 ~ "**",
                                      pval < 0.001 & pval >= 0.0001 ~ "***",
                                      pval < 0.0001 ~ "****") ) #%>%
  #mutate(Molecule = ifelse(Molecule == "2-hydroxybutyric acid", "4-Hydroxybutyric acid", Molecule))

write.csv(corr_table %>% 
            filter(df1_name == "Volume (mm\u00B3)") %>% 
            left_join(unique_features) %>%
            dplyr::select(!c("Correlation2", "df2_type", "Molecule", "df1_name")) %>% 
            dplyr::rename(Molecule = Rename) %>%
            relocate(Molecule, Metabolism), 
          paste0(drive_path, "tables/stone-volume-correlations.csv"), row.names = FALSE)

write.csv(corr_table %>% 
            filter(df1_name == "Number") %>% 
            left_join(unique_features) %>%            
            dplyr::select(!c("Correlation2", "df2_type", "Molecule", "df1_name")) %>% 
            dplyr::rename(Molecule = Rename) %>%
            relocate(Molecule, Metabolism), 
          paste0(drive_path, "tables/stone-number-correlations.csv"), row.names = FALSE)

write.csv(corr_table %>% 
            filter(df1_name == "Surface area (mm\u00B2)") %>% 
            left_join(unique_features) %>%           
            dplyr::select(!c("Correlation2", "df2_type", "Molecule", "df1_name")) %>% 
            dplyr::rename(Molecule = Rename) %>%
            relocate(Molecule, Metabolism), 
          paste0(drive_path, "tables/stone-area-correlations.csv"), row.names = FALSE)

sig_features <- corr_table %>%
  filter(pval < 0.05) %>%
  pull(Molecule) %>%
  unique()

dim(corr_table) #2032   10 with area 3048
length(sig_features) # 109 with area 112

saveRDS(corr_table, paste0(drive_path, "tables/correlation-table.rds")) # the one without area was moved to an archive folder
```

```{r}
## Volume correlations
data <- corr_table %>% 
  filter(df1_name == "Volume (mm\u00B3)", Molecule %in% sig_features) %>%
  left_join(unique_features) %>%
  filter(Metabolism != "") %>%
  mutate(Rename = gsub("*", "", Rename, fixed = TRUE),
         Metabolism = factor(Metabolism, levels = c("Arginine", "Cysteine and methionine","Histidine",
                                                "Lysine and carnitine","Tryptophan","Other",
                                                "Nucleotide", "Vitamin, flavonoid and folate","Glycolysis")))
length(unique(data$Molecule) ) #19

p <- 
  ggplot(data, aes(x = df1_type, y = Rename)) +
    geom_point(aes(size = -log(pval)), colour = "grey", stroke = 1) +
    geom_point(aes(colour = Correlation2, size = -log(pval))) +
    geom_text(aes(size = -log(pval), label = Sign), nudge_y = 0.1) +
    geom_text(aes(label = Significance), nudge_x = 0.15, nudge_y = 0.1) +
    facet_grid(Metabolism~., scales = "free", space = "free") +
    labs(title = "Urinary bladder stone volume correlations", x = "", colour = "Pearson (r)", y = "Metabolite", size = "-Log(p-val)") +
    theme +
    theme(strip.text.y = element_text(angle = 0, size = 6),
          legend.key.size = unit(0.4, 'cm'),
          legend.title = element_text(size = 8),
          legend.text = element_text(size = 6)) +
    guides(fill = "none") +
    scale_colour_gradient2(low = "#7DB8F3" , mid = "white", high = "#b08802")

ggsave(paste0(drive_path, "figure8/stone-volume-with-metabolism.tiff"), p, width = 15, height = 11, units = 'cm', dpi = 600)  
```

```{r}
## Number correlations
data <- corr_table %>% 
  filter(df1_name == "Number", Molecule %in% sig_features) %>%
  left_join(unique_features) %>%
  filter(Metabolism != "") %>%
  mutate(Rename = gsub("*", "", Rename, fixed = TRUE),
         Metabolism = factor(Metabolism, levels = c("Arginine", "Cysteine and methionine","Histidine",
                                                "Lysine and carnitine","Tryptophan","Other",
                                                "Nucleotide", "Vitamin, flavonoid and folate","Glycolysis")))
length(unique(data$Molecule) ) #19

p <- 
  ggplot(data, aes(x = df1_type, y = Rename)) +
    geom_point(aes(size = -log(pval)), colour = "grey", stroke = 1) +
    geom_point(aes(colour = Correlation2, size = -log(pval))) +
    geom_text(aes(size = -log(pval), label = Sign), nudge_y = 0.1) +
    geom_text(aes(label = Significance), nudge_x = 0.15, nudge_y = 0.1) +
    facet_grid(Metabolism~., scales = "free", space = "free") +
    labs(title = "Urinary bladder stone number correlations", x = "", colour = "Pearson (r)", y = "Metabolite", size = "-Log(p-val)") +
    theme +
    theme(strip.text.y = element_text(angle = 0, size = 6),
          legend.key.size = unit(0.4, 'cm'),
          legend.title = element_text(size = 8),
          legend.text = element_text(size = 6)) +
    guides(fill = "none") +
    scale_colour_gradient2(low = "#7DB8F3" , mid = "white", high = "#b08802")

ggsave(paste0(drive_path, "figure8/stone-number-with-metabolism.tiff"), p, width = 15, height = 11, units = 'cm', dpi = 600)  
```

```{r}
## Area correlations
data <- corr_table %>% 
  filter(df1_name == "Surface area (mm\u00B2)", Molecule %in% sig_features) %>%
  left_join(unique_features) %>%
  filter(Metabolism != "") %>%
  mutate(Rename = gsub("*", "", Rename, fixed = TRUE),
         Metabolism = factor(Metabolism, levels = c("Arginine", "Cysteine and methionine","Histidine",
                                                "Lysine and carnitine","Tryptophan","Other",
                                                "Nucleotide", "Vitamin, flavonoid and folate","Glycolysis")))
length(unique(data$Molecule) ) #19

p <- 
  ggplot(data, aes(x = df1_type, y = Rename)) +
    geom_point(aes(size = -log(pval)), colour = "grey", stroke = 1) +
    geom_point(aes(colour = Correlation2, size = -log(pval))) +
    geom_text(aes(size = -log(pval), label = Sign), nudge_y = 0.1) +
    geom_text(aes(label = Significance), nudge_x = 0.15, nudge_y = 0.1) +
    facet_grid(Metabolism~., scales = "free", space = "free") +
    labs(title = "Urinary bladder stone surface area correlations", x = "", colour = "Pearson (r)", y = "Metabolite", size = "-Log(p-val)") +
    theme +
    theme(strip.text.y = element_text(angle = 0, size = 6),
          legend.key.size = unit(0.4, 'cm'),
          legend.title = element_text(size = 8),
          legend.text = element_text(size = 6)) +
    guides(fill = "none") +
    scale_colour_gradient2(low = "#7DB8F3" , mid = "white", high = "#b08802")

ggsave(paste0(drive_path, "figure8/stone-area-with-metabolism.tiff"), p, width = 15, height = 11, units = 'cm', dpi = 600)  
```

```{r}
# Unique in area
d1 <- corr_table %>% 
  filter(df1_name == "Number", pval < 0.05) 
d2 <- corr_table %>% 
  filter(df1_name == "Volume (mm\u00B3)", pval < 0.05) 

corr_table %>% 
  filter(df1_name == "Surface area (mm\u00B2)", pval < 0.05, !Molecule %in% unique(c(d1$Molecule, d2$Molecule)) )
```


### Overlapping correlations

Common metabolites that correlate with stone volume, number and area.

```{r}
## Selected overlapping
selected_overlapping <- corr_table %>% 
  filter(pval < 0.05) %>% 
  arrange(Molecule) %>% 
  group_by(Molecule) %>% 
  dplyr::count(df1_name) %>%
  group_by(Molecule) %>% 
  dplyr::count() %>% 
  filter(n > 1) %>% 
  filter(Molecule %in% c("Ornithine", "Indole-3-acetylglycine", "Imidazolepropionic acid", "Formiminoglutamic acid", 
                         "2-amino-2-thiazoline-4-carboxylic acid", "L-serine", "Glutathione thiol", 
                         "Cysteineglutathione disulfide", "Cysteic acid", "Ascorbic acid", "2-hydroxybutyric acid")) %>% 
  pull(Molecule)
length(selected_overlapping) # 11

data <- corr_table %>%
    left_join(unique_features) %>%  filter(Molecule %in% selected_overlapping) %>%
  mutate(Rename = gsub("*", "", Rename, fixed = TRUE),
         df1_name = factor(df1_name, levels = c("Number", "Surface area (mm\u00B2)", "Volume (mm\u00B3)")),
         Metabolism = factor(Metabolism, levels = c("Arginine", "Cysteine and methionine","Histidine",
                                                "Lysine and carnitine","Tryptophan","Other",
                                                "Nucleotide", "Vitamin, flavonoid and folate","Glycolysis", "")))

p <- 
  ggplot(data, aes(x = df1_type, y = Rename)) +
    geom_point(aes(size = -log(pval)), colour = "grey", stroke = 1) +
    geom_point(aes(colour = Correlation2, size = -log(pval))) +
    geom_text(aes(size = -log(pval), label = Sign), nudge_y = 0.1) +
    geom_text(aes(label = Significance), nudge_x = 0.15, nudge_y = 0.1) +
    facet_grid(Metabolism~df1_name, scales = "free", space = "free") +
    labs(title = "Urinary bladder stone correlations (mapped)", x = "", colour = "Pearson (r)", y = "Metabolite", size = "-Log(p-val)") +
    theme +
    theme(strip.text.y = element_text(angle = 0, size = 6),
          legend.key.size = unit(0.4, 'cm'),
          legend.title = element_text(size = 8),
          legend.text = element_text(size = 6)) +
    guides(fill = FALSE) +
    scale_colour_gradient2(low = "#7DB8F3" , mid = "white", high = "#b08802")

ggsave(paste0(drive_path, "supplementary/selected-metabs-overlapping-correlations-with-area.tiff"), p, width = 19, height = 8, units = 'cm', dpi = 600) 
```

```{r}
## Individual correlation plots for urine
s <- metab_ur_df %>% 
  rownames_to_column("Molecule") %>% 
  filter(Molecule %in% selected_overlapping) %>%
  filter(Molecule %in% c("L-serine", "Glutathione thiol", "Ascorbic acid", "Cysteic acid", "Cysteineglutathione disulfide", "2-hydroxybutyric acid", "Indole-3-acetylglycine")) %>%
  pivot_longer(!Molecule, names_to = "SampleNumber", values_to = "Intensity") %>%
  right_join(metadata_urine)  %>%
  pivot_longer(c("Total_stone_volume", "Stone_number", "Stone_surface_area"), names_to = "Type", values_to = "Value") %>%
  dplyr::select(Molecule, Intensity, SampleID, Class, Type, Value) %>%
  mutate(Type = ifelse(Type == "Total_stone_volume", "Volume (mm\u00B3)",
                       ifelse(Type == "Stone_number", "Number", 
                              ifelse(Type == "Stone_surface_area", "Surface area (mm\u00B2)", Type))),
         Type = factor(Type, levels = c("Number", "Surface area (mm\u00B2)", "Volume (mm\u00B3)")))

for (i in 1){ #:length(unique(s$Molecule))
  
  # Number
  data <- s %>% filter(Molecule %in% unique(s$Molecule)[i], Type == "Number") 
  caption <- corr_table %>% left_join(unique_features ) %>% filter(Molecule %in% unique(s$Molecule)[i], df1_type == "Urine", df1_name == "Number")
  pval <- ifelse(caption$pval < 0.0005, signif(as.numeric(format(caption$pval, scientific = TRUE)), digits = 1), round(caption$pval, 3))
  annotation <- paste(paste("R =", round(caption$Correlation, 3)), paste("p =", pval), sep = ", ")
  
  p1 <- 
    ggplot(data, aes(x = Value, y = Intensity)) +
      geom_smooth(aes(group = Molecule), method = "lm", colour = "darkgrey", lty = "dashed") +
      geom_point(aes(fill = Class), shape = 21, size = 2) +
      theme +
      theme(legend.position = "none", plot.subtitle = element_blank(), plot.title = element_blank(), plot.margin = margin(t = 0, r = 0, b = 0, l = 0)) +
      labs(x = "Number", fill = "Group", caption = annotation) +
      scale_fill_manual(values = group_palette)
  
  # Surface area
  data <- s %>% filter(Molecule %in% unique(s$Molecule)[i], Type == "Surface area (mm\u00B2)") 
  caption <- corr_table %>% left_join(unique_features) %>% filter(Molecule %in% unique(s$Molecule)[i], df1_type == "Urine", df1_name == "Surface area (mm\u00B2)")
  pval <- ifelse(caption$pval < 0.0005, signif(as.numeric(format(caption$pval, scientific = TRUE)), digits = 1), round(caption$pval, 3))
  annotation <- paste(paste("R =", round(caption$Correlation, 3)), paste("p =", pval), sep = ", ")

  p2 <- 
    ggplot(data, aes(x = Value, y = Intensity)) +
      geom_smooth(aes(group = Molecule), method = "lm", colour = "darkgrey", lty = "dashed") +
      geom_point(aes(fill = Class), shape = 21, size = 2) +
      theme +
      theme(legend.position = "none", plot.subtitle = element_blank(), plot.title = element_blank(), plot.margin = margin(t = 0, r = 0, b = 0, l = 0)) +
      labs(x = expression(Area~(mm^2)), fill = "Group", caption = annotation) +
      scale_fill_manual(values = group_palette)
  
  # Volume
  data <- s %>% filter(Molecule %in% unique(s$Molecule)[i], Type == "Volume (mm\u00B3)") 
  caption <- corr_table %>% left_join(unique_features) %>% filter(Molecule %in% unique(s$Molecule)[i], df1_type == "Urine", df1_name == "Volume (mm\u00B3)")
  pval <- ifelse(caption$pval < 0.0005, signif(as.numeric(format(caption$pval, scientific = TRUE)), digits = 1), round(caption$pval, 3))
  annotation <- paste(paste("R =", round(caption$Correlation, 3)), paste("p =", pval), sep = ", ")

  p3 <- 
    ggplot(data, aes(x = Value, y = Intensity)) +
      geom_smooth(aes(group = Molecule), method = "lm", colour = "darkgrey", lty = "dashed") +
      geom_point(aes(fill = Class), shape = 21, size = 2) +
      theme +
      theme(legend.position = "none", plot.subtitle = element_blank(), plot.title = element_blank(), plot.margin = margin(t = 0, r = 0, b = 0, l = 0)) +
      labs(x = expression(Volume~(mm^3)), fill = "Group", caption = annotation) +
      scale_fill_manual(values = group_palette)
  
  p <- p1 + p2 + p3 + plot_annotation(title = unique(s$Molecule)[i], 
                           theme = theme(title = element_text(face = "bold", size = 8), 
                                         plot.subtitle = element_blank(),
                                         plot.margin = margin(t = 0, r = 3, b = 0, l = 0)))

ggsave(paste0(drive_path, "figure8/scatterplots/", unique(s$Molecule)[i], "-urine-with-area.tiff"), p, width = 9.5, height = 4, units = 'cm', dpi = 600)
}
```

```{r}
## Individual correlation plots for serum
s <- metab_se_df %>% 
  rownames_to_column("Molecule") %>% 
  filter(Molecule %in% selected_overlapping) %>%
  filter(!Molecule %in% c("L-serine", "Glutathione thiol", "Ascorbic acid", "Cysteic acid", "Cysteineglutathione disulfide")) %>%
  pivot_longer(!Molecule, names_to = "SampleNumber", values_to = "Intensity") %>%
  right_join(metadata_serum)  %>%
  pivot_longer(c("Total_stone_volume", "Stone_number", "Stone_surface_area"), names_to = "Type", values_to = "Value") %>%
  dplyr::select(Molecule, Intensity, SampleID, Class, Type, Value) %>%
  mutate(Type = ifelse(Type == "Total_stone_volume", "Volume (mm\u00B3)",
                       ifelse(Type == "Stone_number", "Number", 
                              ifelse(Type == "Stone_surface_area", "Surface area (mm\u00B2)", Type))),
         Type = factor(Type, levels = c("Number", "Surface area (mm\u00B2)", "Volume (mm\u00B3)")))

for (i in 2){ #:length(unique(s$Molecule))
  
  # Number
  data <- s %>% filter(Molecule %in% unique(s$Molecule)[i], Type == "Number") 
  caption <- corr_table %>% left_join(unique_features  ) %>% filter(Molecule %in% unique(s$Molecule)[i], df1_type == "Serum", df1_name == "Number")
  pval <- ifelse(caption$pval < 0.0005, signif(as.numeric(format(caption$pval, scientific = TRUE)), digits = 1), round(caption$pval, 3))
  annotation <- paste(paste("R =", round(caption$Correlation, 3)), paste("p =", pval), sep = ", ")
  
  p1 <- 
    ggplot(data, aes(x = Value, y = Intensity)) +
      geom_smooth(aes(group = Molecule), method = "lm", colour = "darkgrey", lty = "dashed") +
      geom_point(aes(fill = Class), shape = 21, size = 2) +
      theme +
      theme(legend.position = "none", plot.subtitle = element_blank(), plot.title = element_blank(), plot.margin = margin(t = 0, r = 0, b = 0, l = 0)) +
      labs(x = "Number", fill = "Group", caption = annotation) +
      scale_fill_manual(values = group_palette)
  
    # Surface area
  data <- s %>% filter(Molecule %in% unique(s$Molecule)[i], Type == "Surface area (mm\u00B2)") 
  caption <- corr_table %>% left_join(unique_features ) %>% filter(Molecule %in% unique(s$Molecule)[i], df1_type == "Serum", df1_name == "Surface area (mm\u00B2)")
  pval <- ifelse(caption$pval < 0.0005, signif(as.numeric(format(caption$pval, scientific = TRUE)), digits = 1), round(caption$pval, 3))
  annotation <- paste(paste("R =", round(caption$Correlation, 3)), paste("p =", pval), sep = ", ")

  p2 <- 
    ggplot(data, aes(x = Value, y = Intensity)) +
      geom_smooth(aes(group = Molecule), method = "lm", colour = "darkgrey", lty = "dashed") +
      geom_point(aes(fill = Class), shape = 21, size = 2) +
      theme +
      theme(legend.position = "none", plot.subtitle = element_blank(), plot.title = element_blank(), plot.margin = margin(t = 0, r = 0, b = 0, l = 0)) +
      labs(x = expression(Area~(mm^2)), fill = "Group", caption = annotation) +
      scale_fill_manual(values = group_palette)
  
  # Volume
  data <- s %>% filter(Molecule %in% unique(s$Molecule)[i], Type == "Volume (mm\u00B3)") 
  caption <- corr_table %>% left_join(unique_features ) %>% filter(Molecule %in% unique(s$Molecule)[i], df1_type == "Serum", df1_name == "Volume (mm\u00B3)")
  pval <- ifelse(caption$pval < 0.0005, signif(as.numeric(format(caption$pval, scientific = TRUE)), digits = 1), round(caption$pval, 3))
  annotation <- paste(paste("R =", round(caption$Correlation, 3)), paste("p =", pval), sep = ", ")
  
  p3 <- 
    ggplot(data, aes(x = Value, y = Intensity)) +
      geom_smooth(aes(group = Molecule), method = "lm", colour = "darkgrey", lty = "dashed") +
      geom_point(aes(fill = Class), shape = 21, size = 2) +
      theme +
      theme(legend.position = "none", plot.subtitle = element_blank(), plot.title = element_blank(), plot.margin = margin(t = 0, r = 0, b = 0, l = 0)) +
      labs(x = expression(Volume~(mm^3)), fill = "Group", caption = annotation) +
      scale_fill_manual(values = group_palette)
  
  p <- p1 + p2 + p3 + plot_annotation(title = unique(s$Molecule)[i], 
                           theme = theme(title = element_text(face = "bold", size = 8), 
                                         plot.subtitle = element_blank(),
                                         plot.margin = margin(t = 0, r = 3, b = 0, l = 0)))

ggsave(paste0(drive_path, "figure8/scatterplots/", unique(s$Molecule)[i], "-serum-with-area.tiff"), p, width = 9.5, height = 4, units = 'cm', dpi = 600)
}
```

```{r}
# All overlapping correlations
overlapping_correlations <- corr_table %>% 
  filter(pval < 0.05) %>% 
  arrange(Molecule) %>% 
  group_by(Molecule) %>% 
  dplyr::count(df1_name) %>%
  group_by(Molecule) %>% 
  dplyr::count() %>% 
  filter(n > 1) %>%
  filter(!Molecule %in% selected_overlapping)

data <- corr_table %>%
  left_join(unique_features) %>% 
  filter(Molecule %in% overlapping_correlations$Molecule) %>%
  mutate(Rename = ifelse(Rename == "", Molecule, Rename),
         Rename = gsub("*", "", Rename, fixed = TRUE),
         df1_name = factor(df1_name, levels = c("Number", "Surface area (mm\u00B2)", "Volume (mm\u00B3)")) )

p <- 
  ggplot(data, aes(x = df1_type, y = Rename)) +
    geom_point(aes(size = -log(pval)), colour = "grey", stroke = 1) +
    geom_point(aes(colour = Correlation2, size = -log(pval))) +
    geom_text(aes(size = -log(pval), label = Sign), nudge_y = 0.1) +
    geom_text(aes(label = Significance), nudge_x = 0.15, nudge_y = 0.1) +
    facet_grid(.~df1_name, scales = "free", space = "free") +
    labs(title = "Urinary bladder stone correlations (unmapped)", x = "", colour = "Pearson (r)", y = "Metabolite", size = "-Log(p-val)") +
    theme +
    theme(legend.key.size = unit(0.4, 'cm'),
          legend.title = element_text(size = 8),
          legend.text = element_text(size = 6)) +
    guides(fill = "none") +
    scale_colour_gradient2(low = "#7DB8F3" , mid = "white", high = "#b08802")

ggsave(paste0(drive_path, "supplementary/all-metabs-overlapping-correlations-with-area.tiff"), p, width = 26, height = 26, units = 'cm', dpi = 600) 
```


### Correlation cut off

```{r}
top_corr_metabolites <- corr_table %>% filter(abs(Correlation) >= 0.7, pval < 0.05) %>% pull(Molecule) %>% unique()
length(top_corr_metabolites) # 23 with area 24

data <- corr_table %>%
  left_join(unique_features) %>% 
  filter(Molecule %in% top_corr_metabolites, !Molecule %in% selected_overlapping) %>%
  mutate(Rename = gsub("*", "", Rename, fixed = TRUE),
         df1_name = factor(df1_name, levels = c("Number", "Surface area (mm\u00B2)", "Volume (mm\u00B3)")))

p <- 
  ggplot(data, aes(x = df1_type, y = Rename)) +
    geom_point(aes(size = -log(pval)), colour = "grey", stroke = 1) +
    geom_point(aes(colour = Correlation2, size = -log(pval))) +
    geom_text(aes(size = -log(pval), label = Sign), nudge_y = 0.1) +
    geom_text(aes(label = Significance), nudge_x = 0.15, nudge_y = 0.1) +
    facet_grid(.~df1_name, scales = "free", space = "free") +
    labs(title = "Urinary bladder stone correlations (r > 0.7)", x = "", colour = "Pearson (r)", y = "Metabolite", size = "-Log(p-val)") +
    theme +
    theme(legend.key.size = unit(0.4, 'cm'),
          legend.title = element_text(size = 8),
          legend.text = element_text(size = 6)) +
    guides(fill = FALSE) +
    scale_colour_gradient2(low = "#7DB8F3" , mid = "white", high = "#b08802")

ggsave(paste0(drive_path, "supplementary/r>0.7-metabs-correlations-with-area.tiff"), p, width = 22, height = 10, units = 'cm', dpi = 600) 
```

# Unique to area

```{r}
## Individual correlation plots for urine
s <- metab_ur_df %>% 
  rownames_to_column("Molecule") %>% 
  filter(Molecule %in% "4-methyl-n-[4-[(4-nitrophenyl)azaniumyl]phenyl]-4-oxidopiperazin-4-ium-1-carboximidothioate") %>%
  pivot_longer(!Molecule, names_to = "SampleNumber", values_to = "Intensity") %>%
  right_join(metadata_urine)  %>%
  pivot_longer(c("Stone_surface_area"), names_to = "Type", values_to = "Value") %>%
  dplyr::select(Molecule, Intensity, SampleID, Class, Type, Value)

for (i in 1:length(unique(s$Molecule))){
  
  data <- s #%>% filter(Molecule %in% unique(s$Molecule)[i]) 
  caption <- corr_table %>% left_join(unique_features) %>% filter(Molecule %in% unique(s$Molecule), df1_type == "Urine", df1_name == "Surface area (mm\u00B2)")
  pval <- ifelse(caption$pval < 0.0005, signif(as.numeric(format(caption$pval, scientific = TRUE)), digits = 1), round(caption$pval, 3))
  annotation <- paste(paste("R =", round(caption$Correlation, 3)), paste("p =", pval), sep = ", ")

  p1 <- 
    ggplot(data, aes(x = Value, y = Intensity)) +
      geom_smooth(aes(group = Molecule), method = "lm", colour = "darkgrey", lty = "dashed") +
      geom_point(aes(fill = Class), shape = 21, size = 2) +
      theme +
      theme(legend.position = "none", plot.subtitle = element_blank(), plot.title = element_text(size = 3), plot.margin = margin(t = 0, r = 0, b = 0, l = 0)) +
      labs(x = expression("Surface area"~(mm^2)), fill = "Group", caption = annotation, title = unique(s$Molecule)) +
      scale_fill_manual(values = group_palette)

ggsave(paste0(drive_path, "stone-area/scatterplots/", unique(s$Molecule)[i], "-urine.tiff"), p1, width = 4, height = 4, units = 'cm', dpi = 600)
}
```


# --------------------
## Supplementary
### PCA

```{r}
## Urine
set.seed(3)
a <- data.frame(adonis2(formula = as.formula(paste('t(metab_ur_df) ~', "Class")), 
                      data = metadata_urine, 
                      distance = "eu", 
                      permutations = 999,
                      na.action = na.omit,
                      parallel = 6))

annotation <- paste(paste("SumOfSqs =", round(a["Class",]$SumOfSqs, 3)),
              paste("R2 =", round(a["Class",]$R2, 3)),
              paste("p =", round(a["Class",]$Pr..F., 3)), sep = " ; " )

pca <- pca_process(data = metab_ur_df, metadata = metadata_urine)

p <- 

  ggplot(pca$data, aes(x = PC1, y = PC2)) +
  #stat_ellipse(aes(fill = Class), geom = 'polygon' , type = 't', level = 0.9, alpha = 0.2) +
  ggforce::geom_mark_ellipse(aes(fill = Class), expand = unit(2.5, "mm"), colour = NA, alpha = 0.2) +
  geom_point(aes(fill = Class), shape = 21) +
  scale_fill_manual(values = group_palette) +
  theme +
  labs(title = "Urine metabolomics",
       caption = annotation,
       fill = "Group",
       x = paste0('PC1 ', round(pca$varexp[1], 2), '%'),
       y = paste0('PC2 ', round(pca$varexp[2], 2), '%')) +
  scale_y_continuous(limits = c(-55, 40)) +
  scale_x_continuous(limits = c(-40, 70))

ggsave(paste0(drive_path, "supplementary/het-homo-ala-urine-pca.tiff"), p, width = 3, height = 2, dpi = 600)
```

```{r}
##Serum
set.seed(3)
a <- data.frame(adonis2(formula = as.formula(paste('t(metab_se_df) ~', "Class")), 
                      data = metadata_serum, 
                      distance = "eu", 
                      permutations = 999,
                      na.action = na.omit,
                      parallel = 6))

annotation <- paste(paste("SumOfSqs =", round(a["Class",]$SumOfSqs, 3)),
              paste("R2 =", round(a["Class",]$R2, 3)),
              paste("p =", round(a["Class",]$Pr..F., 3)), sep = " ; " )

pca <- pca_process(data = metab_se_df, metadata = metadata_serum)

p <- 

  ggplot(pca$data, aes(x = PC1, y = PC2)) +
  ggforce::geom_mark_ellipse(aes(fill = Class), expand = unit(2.5, "mm"), colour = NA, alpha = 0.2) +
  geom_point(aes(fill = Class), shape = 21) +
  scale_fill_manual(values = group_palette) +
  theme +
  theme(text = element_text(family = "Arial")) +
  labs(title = "Serum metabolomics",
       caption = annotation,
       fill = "Group",
       x = paste0('PC1 ', round(pca$varexp[1], 2), '%'),
       y = paste0('PC2 ', round(pca$varexp[2], 2), '%')) +
  scale_y_continuous(limits = c(-30, 30)) +
  scale_x_continuous(limits = c(-40, 40))

ggsave(paste0(drive_path, "supplementary/het-homo-ala-serum-pca.tiff"), p, width = 3, height = 2, dpi = 600)
```

```{r}
## Urine vs Serum 
set.seed(3)
a <- data.frame(adonis2(formula = as.formula(paste('t(metab_se_ur_df) ~', "Class")), 
                      data = metadata_ur_se, 
                      distance = "eu", 
                      permutations = 999,
                      na.action = na.omit,
                      parallel = 6))

annotation <- paste(paste("SumOfSqs =", round(a["Class",]$SumOfSqs, 3)),
              paste("R2 =", round(a["Class",]$R2, 3)),
              paste("p =", round(a["Class",]$Pr..F., 3)), sep = " ; " )

pca <- pca_process(data = metab_se_ur_df, metadata = metadata_ur_se)

p <- 

  ggplot(pca$data, aes(x = PC1, y = PC2)) +
  ggforce::geom_mark_ellipse(aes(fill = Class, group = Group_type), expand = unit(2.5, "mm"), colour = NA, alpha = 0.2) +
  geom_point(aes(fill = Class, shape = SampleType)) +
  scale_fill_manual(values = group_palette) +
  theme +
  labs(title = "Urine and Serum metabolomics",
       caption = annotation,
       fill = "Group",
       shape = "Sample",
       x = paste0('PC1 ', round(pca$varexp[1], 2), '%'),
       y = paste0('PC2 ', round(pca$varexp[2], 2), '%')) +
  scale_x_continuous(limits = c(-30, 30)) +
  scale_y_continuous(limits = c(-20, 25)) +
  scale_shape_manual(values = c("Urine" = 21, "Serum" = 22))

ggsave(paste0(drive_path, "supplementary/urine-serum-pca.tiff"), p, width = 3.5, height = 2.5, dpi = 600)
```

### Overlapping metabolites 

```{r}
length(intersect(c(limma_list$serum$HomoCtrl_vs_HeteroCtrl$sig_hits$Molecule, limma_list$urine$HomoCtrl_vs_HeteroCtrl$sig_hits$Molecule), 
          c(limma_list$serum$HomoALA_vs_HomoCtrl$sig_hits$Molecule, limma_list$urine$HomoALA_vs_HomoCtrl$sig_hits$Molecule) )) # 104
```

