---
title: "Cysteinuria Project"
author: "Giulia Iacono"
date: "30th June 2026"
output: 
  
---


```{r}
knitr::opts_chunk$set(error = FALSE, warning = FALSE, message = FALSE)
```

```{r libraries}
# Get the R version
# version$version.string

# Define a vector of required packages
pkgs <- c('here', 'data.table', 'tidyverse', 'kableExtra', 'ggplot2', 'pmp', 
          'SummarizedExperiment', 'S4Vectors', 'ComplexHeatmap', 'circlize', 'parallel',
          'ggsci', 'ggpubr', 'r2symbols', 'xml2',
          'doParallel', 'foreach', 'gridExtra', 'patchwork', 'readxl',
          'ggrepel', 'gplots', 'ggVennDiagram',
          'limma', 'FELLA', 'igraph', 'vegan', 'Hmisc', 'sjmisc')

# For each of the packages, load it and display the version number
for (pkg in pkgs) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
   print(paste0(pkg, ': ', packageVersion(pkg)))
}

#source(here::here("functions", "extra.functions.R"))
```

# Theme and palettes

```{r}
#group_type_palette <- c('#b2df8a','#33a02c','#fb9a99','#e31a1c','#fdbf6f','#ff7f00')
group_type_palette <- c('#7D83F8', "#1c2175", "#7DB8F3", '#3971a8', "grey", '#5C6378')
names(group_type_palette) <- c("Homo-ALA-Serum", "Homo-ALA-Urine", "Homo-Ctrl-Serum", "Homo-Ctrl-Urine", "Hetero-Ctrl-Serum","Hetero-Ctrl-Urine")

group_palette_serum <- c('#7D83F8', '#7DB8F3', 'grey')
names(group_palette_serum) <- c("Homo-ALA", "Homo-Ctrl", "Hetero-Ctrl")

group_palette_urine <- c('#1c2175', '#3971a8', '#5C6378')
names(group_palette_urine) <- c("Homo-ALA", "Homo-Ctrl", "Hetero-Ctrl")

group_palette <- c('#7D83F8', '#7DB8F3', 'grey')
names(group_palette) <- c("Homo-ALA", "Homo-Ctrl", "Hetero-Ctrl")

theme <- theme_classic() + 
         theme(plot.title = element_text(face = "bold"),
               strip.background = element_rect(fill = "lightgrey", linewidth = 0),
               strip.text = element_text(face = "bold", colour = "black"),
               legend.position = "right", legend.justification = "left",
               text = element_text(family = "Arial", size = 8)
)

metabolism_palette <- c("Tryptophan" = "#A8DCD1", 
                        "Lysine and carnitine" = "#995de3", 
                        "Arginine" = "#5F72A1", 
                        "Cysteine and methionine" = "#1c97ba", 
                        "Histidine" = "#083D77", 
                       # "Lipid and fatty acyl" = "#035443", 
                        "Nucleotide" = "#209179", 
                        "Glycolysis" = "#224ebd", 
                        "Vitamin, flavonoid and folate" = "#89e2fa", 
                        "Sulfur" = "#4a1f80",
                        "Other" = "grey")

```


# Import/Format Data

```{r}
metadata <- read.csv(here::here("data", "metadata.csv"),  strip.white = T, sep = "\t")

## Remove Serum and urine mixed QC
metadata <- metadata[!grepl("se_ur", metadata$SampleNumber),]

## edit names in Group for easier selection
metadata <- metadata %>% mutate(class = case_when(
  grepl("a-LA treated homozygote", Group) ~ "Homo-ALA",
  grepl("homozygote", Group) ~ "Homo-Ctrl",
  grepl("heterozygote", Group) ~ "Hetero-Ctrl",
  grepl("Blank", SampleNumber) ~ "blank",
  grepl("QC", SampleNumber) ~ "QC",
))

metadata$SampleType <- case_when(
  grepl("^se", metadata$SampleNumber) & !(grepl ("QC", metadata$SampleNumber)) ~ "Serum",
  grepl("^ur", metadata$SampleNumber) & !(grepl ("QC", metadata$SampleNumber)) & !(grepl ("blank", metadata$SampleNumber)) ~ "Urine",
  grepl("Blank", metadata$SampleNumber) ~ "blank",
  grepl("QC", metadata$SampleNumber) ~ "QC"
)

rownames(metadata) <- metadata$SampleNumber
```

```{r, results='hide'}
# Load the metabolomics height data
metab_pos <- read_csv(here::here('data', '2024_09_23_height_pos.csv'))
metab_neg <- read_csv(here::here('data', '2024_09_23_height_neg.csv'))

# Remove the MS/MS samples (not acquired in the same way)
metab_pos <- metab_pos[,-(grep('MSMS.*', names(metab_pos)))]
metab_neg <- metab_neg[,-(grep('MSMS.*', names(metab_neg)))]

identical(colnames(metab_pos), colnames(metab_neg)) # TRUE

# Separate into intensity and information data.frames for the SummarizedExperiment object
metab_pos_counts <- as.matrix(metab_pos[, c(36:79)]) #column numbers that contain Blanks, QCs, and samples
metab_neg_counts <- as.matrix(metab_neg[, c(36:79)])

metab_pos_info <- data.frame(metab_pos[, 1:35])
metab_neg_info <- data.frame(metab_neg[, 1:35])
metab_pos_info$ionisation <- "pos"
metab_neg_info$ionisation <- "neg"
# Rename the data to indicate ionisation mode (using the MS-DIAL alignment ID)
pos_rownames <- paste0(metab_pos_info$Alignment.ID, '_pos')
neg_rownames <- paste0(metab_neg_info$Alignment.ID, '_neg')

## add as rownames for easy selection
rownames(metab_pos_counts) <- pos_rownames
rownames(metab_pos_info) <- pos_rownames
metab_pos_info$MetabID <- pos_rownames
rownames(metab_neg_counts) <- neg_rownames
rownames(metab_neg_info) <- neg_rownames
metab_neg_info$MetabID <- neg_rownames

# Merge the positive and negative ionisation modes using rbind
metab_counts <- rbind(metab_pos_counts, metab_neg_counts)
metab_info <- rbind(metab_pos_info, metab_neg_info)
```

# SummarizedExperiment

```{r}
# Reorder so that sample name order matches
metab_counts <- metab_counts[,rownames(metadata)]
metab_pos_counts <- metab_pos_counts[,rownames(metadata)]
metab_neg_counts <- metab_neg_counts[,rownames(metadata)]
# Check that the metadata matches the samples
identical(rownames(metadata), colnames(metab_counts)) # should return 'TRUE'

## Split into urine and serum
metadata_urine <- metadata[grep("ur", metadata$SampleNumber), ]
metab_counts_ur <- metab_counts[, rownames(metadata_urine)]
metab_pos_ur <- metab_pos_counts[, rownames(metadata_urine)] ## create to check TIC
metab_neg_ur <- metab_neg_counts[, rownames(metadata_urine)]

metadata_serum <- metadata[grep("se", metadata$SampleNumber),]
metab_counts_se <- metab_counts[, rownames(metadata_serum)]
metab_pos_se <- metab_pos_counts[, rownames(metadata_serum)] ## create to check TIC
metab_neg_se <- metab_neg_counts[, rownames(metadata_serum)]

# Create the SummarizedExperiment (SE) object
metab_ur <- SummarizedExperiment(assays = list(counts = metab_counts_ur),
                                       metadata = metadata_urine,
                                       rowData = list(info = metab_info),
                                       colData = metadata_urine)

metab_se <- SummarizedExperiment(assays = list(counts = metab_counts_se),
                                       metadata = metadata_serum,
                                       rowData = list(info = metab_info),
                                       colData = metadata_serum)

metab_obj <- SummarizedExperiment(assays = list(counts = metab_counts),
                                       metadata = metadata,
                                       rowData = list(info = metab_info),
                                       colData = metadata)
```

```{r}
# Check TIC across samples
TIC <- function(intensity_table, metatdata){
  total <- data.frame(Total.intensity = colSums(intensity_table)) %>%
    mutate(SampleType = metatdata$SampleType) %>%
    rownames_to_column(., var = "SampleID") 
  total$SampleID <- factor(total$SampleID, levels = total$SampleID)
  
  ggplot(total, aes(x = SampleID, y = Total.intensity, fill = SampleType)) +
    geom_bar(stat = "identity") +
    labs(title = "Total intensity across samples") +
    theme(axis.text.x = element_text(angle = 90))
}

# split by sample type
TIC(metab_pos_se, metadata_serum)
TIC(metab_neg_se, metadata_serum)
TIC(metab_pos_ur, metadata_urine)
TIC(metab_neg_ur, metadata_urine)

# all together
TIC(metab_pos_counts, metadata)
TIC(metab_neg_counts, metadata)
```

```{r}
# Perform the PCA and retrieve the explained variance values
pca_function <- function(data){
  
  PCA <- prcomp(t(assay(data)), center = TRUE)
  varexp <- c(summary(PCA)$importance[2,1]*100,summary(PCA)$importance[2,2]*100)
  
  # Create a dataset for plotting
  data.PCA <- cbind(data.frame(Samples = rownames(PCA$x),
                                     PC1 = PCA$x[,1],
                                     PC2 = PCA$x[,2],
                                     SampleType = data$SampleType))
  
  # Plot results
  ggplot(data.PCA, aes(x = PC1, y = PC2)) +
    geom_point(aes(colour = SampleType)) +
    stat_ellipse(aes(group = SampleType), geom = 'polygon', type = 't', level = 0.95, alpha = 0.2) +
    labs(title = 'Serum - combined modes',
           x = paste0('PC1 ', round(varexp[1], 2), '%'),
           y = paste0('PC2 ', round(varexp[2], 2), '%'))
}

pca_function(metab_se)
pca_function(metab_ur)
pca_function(metab_obj)
# Extr_blank_2 of the serum blanks is falling within the area of the samples this blank will be removed and we will continue with the other 2
```

# Filtering and normalisation

```{r}
# Check the original number of features
dim(metab_ur) #87211    18 now 50740
dim(metab_se) #87211    23 now 50740

# Replace missing values with NA to be compatible with downstream filtering
assay(metab_ur) <- replace(assay(metab_ur), assay(metab_ur) == 0, NA)
assay(metab_se) <- replace(assay(metab_se), assay(metab_se) == 0, NA)
assay(metab_obj) <- replace(assay(metab_obj), assay(metab_obj) == 0, NA) # will not be filtered based on blanks, but based on retained features across sample types

# Filter peaks and optionally samples based on blanks
metab_filt_ur <- filter_peaks_by_blank(df = metab_ur,
                                           fold_change = 10, # # how many times more in samples over blanks
                                           classes = metab_ur$class,
                                           remove_samples = TRUE,
                                           remove_peaks = TRUE,
                                           blank_label = 'blank') # from the class vector

# Filter peaks and optionally samples based on blanks
metab_filt_se <- filter_peaks_by_blank(df = metab_se,
                                           fold_change = 10, 
                                           classes = metab_se$class,
                                           remove_samples = TRUE,
                                           remove_peaks = TRUE,
                                           blank_label = 'blank') # from the class vector

                                           
# Check the number of features/samples remaining
dim(metab_filt_ur) #72163 now 41961    16 14 + 2 QCs
dim(metab_filt_se) #63246 now 35667    20 17 + 3 QCs
```

## Filtering on missing values and QC intensity variation

```{r}
### Urine Filtering ###
# Filter samples based on the percentage of missing values
metab_filt_ur <- filter_samples_by_mv(df = metab_filt_ur,
                                           max_perc_mv = 0.8, 
                                      classes = metab_filt_ur$class) 
print(paste0("Filter samples based on the percentage of missing values. Results col: ", ncol(metab_filt_ur))  )  # 16 all samples retained

# Filter peaks based on fraction of missing values per feature within QC samples
metab_filt_ur <- filter_peaks_by_fraction(df=metab_filt_ur, 
                                       min_frac=0.9,
                                       classes=metab_filt_ur$class,
                                       method="QC",
                                       qc_label="QC")
print(paste0("Filter peaks based on fraction of missing values per feature within QC samples. Results features: ", nrow(metab_filt_ur)) )     #32999   now 25331      

## We will not use this filter because we expect metabolite variation between groups
# # Filter peaks based on missing values across all samples
# metab_filt_ur <- filter_peaks_by_fraction(df = metab_filt_ur,
#                                                min_frac = 0.20, # features should have values for >80% of samples
#                                                classes = metab_filt_ur$class,
#                                                method = 'across')
#   
# # Check the number of features/samples
# print(paste0("Filter peaks based on missing values across all samples. Results features: ", nrow(metab_filt_ur)) )            
  
# Filter peaks based on the percentage of variation in the QC samples
metab_filt_ur <- filter_peaks_by_rsd(df = metab_filt_ur, # was previously called metab_filt_ur_qc but not used later on
                                          max_rsd = 30,
                                          classes = metab_filt_ur$class,
                                          qc_label = 'QC')
 # Check the number of features/samples
print(paste0("Filter peaks based on the percentage of variation in the QC samples. Results features: ", nrow(metab_filt_ur)) ) #20309 now 19127

saveRDS(metab_filt_ur, paste0(drive_path, "data/rstudio/metab_filt_ur.rds") )

dim(metab_filt_ur) #20309 now 19127    16
```

```{r}
### Serum Filtering ###
# Filter samples based on the percentage of missing values
metab_filt_se <- filter_samples_by_mv(df = metab_filt_se,
                                           max_perc_mv = 0.8, 
                                        classes=metab_filt_se$class)

print(paste0("Filter samples based on the percentage of missing values. Results col: ", ncol(metab_filt_se))  )  # 30 all samples retained
  
# Filter peaks based on fraction of missing values per feature within QC samples
metab_filt_se <- filter_peaks_by_fraction(df=metab_filt_se, 
                                         min_frac=0.9,
                                         classes=metab_filt_se$class,
                                         method="QC",
                                         qc_label="QC")
print(paste0("Filter peaks based on fraction of missing values per feature within QC samples. Results features: ", nrow(metab_filt_se)) )      #11328   now 9523       
  
## We will not use this filter because we expect metabolite variation between groups
# # Filter peaks based on missing values across all samples
# metab_filt_se <- filter_peaks_by_fraction(df = metab_filt_se,
#                                              min_frac = 0.20, # features should have values for >80% of samples
#                                              classes = metab_filt_se$class,
#                                              method = 'across')
# 
# # Check the number of features/samples
# print(paste0("Filter peaks based on missing values across all samples. Results features: ", nrow(metab_filt_se)) )            

# Filter peaks based on the percentage of variation in the QC samples
metab_filt_se <- filter_peaks_by_rsd(df = metab_filt_se,
                                        max_rsd = 30, # previously 35
                                        classes = metab_filt_se$class,
                                        qc_label = 'QC')
# Check the number of features/samples
print(paste0("Filter peaks based on the percentage of variation in the QC samples. Results features: ", nrow(metab_filt_se)) ) #6057 now 6022

saveRDS(metab_filt_se, paste0(drive_path, "data/rstudio/metab_filt_se.rds") )

dim(metab_filt_se) #6057 now 6022 20 
```

## PQN normalisation and glog scaling

```{r, eval = F}
SE_list <- list(serum = metab_filt_se, urine = metab_filt_ur)

norm_scale_function <- function(metab_filt){
  # PQN data normalisation
  metab_norm <- pqn_normalisation(df = metab_filt,
                                        classes = metab_filt$class,
                                        qc_label = 'QC')
                                        
  # Missing values imputation
  metab_imp <- mv_imputation(df = metab_norm,
                                   method = 'rf') # or rf, bcpa, sv, mn, md.
                                   
  # Data scaling
  metab_glog <- glog_transformation(df = metab_imp,
                                          classes = metab_imp$class,
                                          qc_label = 'QC')
  
  
  opt_lambda <- processing_history(metab_glog)$glog_transformation$lambda_opt
  
  lambda <- glog_plot_optimised_lambda(df = metab_imp,
                             optimised_lambda = opt_lambda,
                             classes = metab_imp$class,
                             qc_label = 'QC')
  
  return(list(norm = metab_norm,
              imp = metab_imp,
              glog = metab_glog,
              opt_lambda = opt_lambda,
              lambda = lambda))
}

SE_list_proc <- mclapply(SE_list, norm_scale_function, mc.cores = 6)

table(rowData(SE_list_proc$serum$glog)@listData[['info.Metabolite.name']] != 'Unknown') #1846   395 now 5193   864 > 276   746
table(rowData(SE_list_proc$urine$glog)@listData[['info.Metabolite.name']] != 'Unknown') #8903  1678  now 17305  3004 > 17236  1891

dim(SE_list_proc$serum$glog) #6057   20 # previously 2241   20 > 6022   
dim(SE_list_proc$urine$glog) #20309  16 # previously 10581    16 > 19127

saveRDS(SE_list_proc, paste0(drive_path, "data/rstudio/SE_list_proc.rds") )

# SE_list_proc <- readRDS("data/new/SE_list_proc.rds")
```

# -----------------------
# Metabolite Annotation

```{r}
## create master info file with all remaining alignmentIDs
metIDs <- c(rownames(SE_list_proc$urine$glog), rownames(SE_list_proc$serum$glog) )
metIDs <- metIDs[!duplicated(metIDs)]
master_info <- metab_info[metIDs,] %>% arrange(ionisation, Alignment.ID) 
```

## HMDB

```{r}
# Load HMDB dataset
hmdb_df <- readRDS(here::here('data', 'hmdb_metabolites_detect_quant_v5_20231102.rds'))

# Search annotations in HMDB and add master_info 
master_info_annotated <- add_hmdb(metab_SE = master_info, hmdb = hmdb_df, mass_tol = 0.002, 6) # 1h with v5 database, 15min with v4 
dim(master_info_annotated) #23235    41 > 22140 44
saveRDS(master_info_annotated, paste0(drive_path, "data/rstudio/master_info_annotated_v5.rds") )

# master_info_annotated <- readRDS("data/new/master_info_annotated_v5.rds")
```

## Compare Annotatings

```{r}
# Prepare data.frame with alignment IDs and all four annotations and filter for at least one annotation
msdial_hmdb_glog <- compare_annotations(master_info_annotated)

# Keep only annotated rows and generate shortname column
master_info_annotated_filtered <- keep_annotated(master_info_annotated) #previously overwritten as master_info
dim(master_info_annotated_filtered) #6142   42 with v4 and 9943   42 with v5 > 7843   45

## Add column for those features found in Serumn and Urine
master_info_annotated_filtered$Serum <- ifelse(master_info_annotated_filtered$MetabID %in% rownames(SE_list_proc$serum$glog), TRUE, FALSE)
master_info_annotated_filtered$Urine <- ifelse(master_info_annotated_filtered$MetabID %in% rownames(SE_list_proc$urine$glog), TRUE, FALSE)

saveRDS(master_info_annotated_filtered, paste0(drive_path, "data/rstudio/master_info_annotated_filtered.rds") )

# master_info_annotated_filtered <- readRDS("data/new/master_info_annotated_filtered.rds")
```

# ----------------------
# Quality filtering

```{r}
# # Save the curation table
df <- data.frame('MetabID' = master_info_annotated_filtered$MetabID,
                  "RT" = master_info_annotated_filtered$Average.Rt.min.,
                  "MZ" = master_info_annotated_filtered$Average.Mz,
                  "SNratio" = master_info_annotated_filtered$S.N.average,
                  "Fill" = master_info_annotated_filtered$Fill..,
                  'MSDIAL_name' = master_info_annotated_filtered$Metabolite.name,
                  'HMDB_name' = master_info_annotated_filtered$HMDB,
                  'shortname' = master_info_annotated_filtered$shortname,
                  'Serum' = master_info_annotated_filtered$Serum,
                  'Urine' = master_info_annotated_filtered$Urine,
                 "Ionisation" = master_info_annotated_filtered$ionisation,
                 "AlignmentID" = master_info_annotated_filtered$Alignment.ID
                 )

# Flag duplicates
df$shortname <- gsub("_pos\\.\\d+|_neg\\.\\d+|_pos|_neg", "", df$shortname)
df$shortname <- gsub("\"", "", df$shortname) # remove quotation marks
df$shortname <- sub("^([a-z])", "\\U\\1", df$shortname, perl = TRUE) # capitalise first letter
df$shortname <- gsub("_", "/", df$shortname)
df$shortname <- tolower(df$shortname)
df$shortname <- str_to_sentence(df$shortname)
```

```{r}
# Merge tables based on previous filtering
#table1 <- read.csv("data/originals/manual_curation2_edit.csv") %>% mutate(Peak_quality = as.character(Peak_quality)) %>% dplyr::select(MetabID, Peak_quality, Shortname)

table <- df %>%
  #left_join(table1) %>%
  mutate(Peak_quality = ifelse(Fill > 0.5 & SNratio > 100, "Pre_assigned_TRUE", FALSE)) # pre-assign TRUE to good peaks

table(table$Peak_quality) # 7133 710 with v5

write_csv(table, "data/20240923_manual_curation_v5.csv")
```


## Edit XML file

```{r}
for (i in c("neg", "pos")){
  
check <- table %>% 
  filter(Ionisation == i) %>% 
  pull(AlignmentID) %>% 
  as.numeric()
  
xml_real <- read_xml(paste0("data/2024_09_23_AlignmentResult_", i, "_tags.xml")) 

# Create backup
write_xml(xml_real, file = paste0("data/2024_09_23_AlignmentResult_", i, "_tags_BACKUP.xml"), encoding = "utf-8")

xml <- xml_real
for (a in check) {
  # Find the parent node where you want to add the child
  parent_node <- xml_find_first(xml, "//Peaks")
  # Create the new child node
  val <- paste0('Peak Id="', a, '"', sep = "")
  new_child <- xml_add_child(parent_node, val)
  new_child2 <- xml_add_child(new_child, "Tag")
  # Set text content for the new child node
  xml_text(new_child2) <- "1"
  # Convert the XML back to a string
  modified_xml <- as.character(xml)
}

# Overwrite file
write_xml(xml, file = paste0("data/2024_09_23_AlignmentResult_", i, "_tags.xml"), encoding = "utf-8")

}

# Do manually for each ionisation in Notepad++ or apple XCode
# Do replace:   /Peak.* with /Peak>    - regular expression
#then replace \n with \r\n       - wrap around and extended
#Peaks and PeakSpotTags last two - remove CRLF at end
```

# Read XML (not used)

```{r}
#xml_Annotated <- xml2::read_xml("./Met_Lip/Exp18_lip/BAL/Neg/EXP18_CSL_Mouse_Lip_BAL_Neg_tags.xml")
#nodes <- xml_find_all(xml_Annotated, "//Peaks")
#rows <- xml_Annotated %>% xml_find_all("//Peaks")
#Ids <- xml_contents(rows) %>% xml_attrs() %>% as.character()
#Quality_Peaks <- xml_contents(rows) %>% as.character() %>% as.list()

#Peak_Data <- data.frame(Ids = paste0(Ids,"_neg"), A1 = FALSE, A2 = FALSE, A3 = FALSE, A4 = FALSE, A5 = FALSE)

#for (i in c(1:length(Quality_Peaks))) {
#  temp <- Quality_Peaks[[i]]
#  str_temp <- strsplit(temp,"  ")[[1]]
#  str_temp <- str_temp[2:length(str_temp)] %>% gsub("[[:punct:]]", "", .) %>% 
#    gsub("Tag", "", .) %>% substr(.,1,1) %>% as.numeric()
#  Vec <- rep(FALSE,5)
#  Vec[str_temp] <- TRUE
#  Peak_Data[i,c(2:6)] <- Vec
#}
```

```{r}
keep <- table[which(duplicated(table$shortname)),] %>% 
  dplyr::select(MetabID, Fill, shortname, Peak_quality) %>%
  arrange(desc(shortname), Peak_quality) %>%
  group_by(shortname) %>%
  filter(Peak_quality == TRUE) %>% 
  filter(sum(Peak_quality) >= 1) %>% # if 1 or multiple duplicates have been flagged as TRUE, keep the one with the highest fill
  slice_max(Fill) %>%
  slice_head(., n = 1) %>% 
  pull(MetabID)
length(keep) # 930

remove <- table[which(duplicated(table$shortname)),] %>% 
  filter(!MetabID %in% keep) %>%
  pull(MetabID)
length(remove) # 3336

dim(table[which(duplicated(table$shortname)),]) #5547

# Remove duplicates in quality table
table <- table %>% 
  mutate(Peak_quality = ifelse(MetabID %in% remove, FALSE, Peak_quality))
sum(is.na(table$Peak_quality)) # 2433 with v5
dim(table) #7843

# view(table[which(duplicated(table$shortname)),])
```

```{r}
ggplot(table, aes(x = Fill, y = log(SNratio))) +
  geom_point(aes(fill = Peak_quality), shape = 21) +
  geom_vline(xintercept = 0.25) +
  geom_hline(yintercept = 3)

ggplot(table, aes(x = RT, y = MZ)) +
  geom_point(aes(fill = Peak_quality), shape = 21)
```

## Curated table

```{r}
# Load quality data
metab_quality <- read.csv("data/20240923_manual_curation_v5_curated.csv")

## check identical so that we can replace shortname with non unique short name - this was a previous mistake
identical(metab_quality$MetabID, master_info_annotated_filtered$MetabID)
```

## Metab info table

```{r}
# Filter info table
metab_info_se_ur <- master_info_annotated_filtered[metab_quality$Peak_quality, ] %>%
  dplyr::select(MetabID, shortname, Fill.., S.N.average, Ontology, Metabolite.name, Formula, HMDB, KEGG, HMDB_accession, CHEBI_ID, PubChem_ID, BiGG_ID, INCHIKEY, SMILES, Serum, Urine) 

# Fix up shortnames
metab_info_se_ur$shortname <- gsub("_pos\\.\\d+|_neg\\.\\d+|_pos|_neg", "", metab_info_se_ur$shortname)
metab_info_se_ur$shortname <- gsub("\"", "", metab_info_se_ur$shortname) # remove quotation marks
metab_info_se_ur$shortname <- sub("^([a-z])", "\\U\\1", metab_info_se_ur$shortname, perl = TRUE) # capitalise first letter
metab_info_se_ur$shortname <- gsub("_", "/", metab_info_se_ur$shortname)
metab_info_se_ur$shortname <- tolower(metab_info_se_ur$shortname)
metab_info_se_ur$shortname <- gsub("no ms2: ", "", metab_info_se_ur$shortname, fixed = TRUE)
metab_info_se_ur$shortname <- gsub(" esa", "", metab_info_se_ur$shortname, fixed = TRUE)
metab_info_se_ur$shortname <- str_to_sentence(metab_info_se_ur$shortname)

dim(metab_info_se_ur) #3618

# Remove duplicates
metab_info_se_ur <- metab_info_se_ur %>% 
  group_by(shortname) %>%
  slice_max(Fill.., n = 1) %>%
  group_by(shortname) %>%
  slice_max(S.N.average, n = 1)

table(duplicated(metab_info_se_ur$shortname))

# view(metab_info_se_ur[which(duplicated(metab_info_se_ur$shortname)),] )

dim(metab_info_se_ur) #2757
```


```{r}
metab_info_se_ur <- metab_info_se_ur %>%
  left_join(hmdb_df[, c("name", "direct_parent", "class", "sub_class")] %>%
              dplyr::rename(shortname = name, CATEGORY = class, MAIN_CLASS = sub_class, SUB_CLASS = direct_parent)) %>%
  as.data.frame()

# Change names so can merge with lipid MSDial annotations
metab_info_se_ur <- metab_info_se_ur %>%
  mutate(SUB_CLASS = ifelse(grepl("erine|itol|glycerol|amine|choline", SUB_CLASS) & 
                              !grepl(paste0(paste0(c("erine", "itol", "glycerol", "amine", "choline"), "s"), collapse = "|"), SUB_CLASS), paste0(SUB_CLASS, "s"), SUB_CLASS),
         SUB_CLASS = gsub("ss", "s", SUB_CLASS),
         SUB_CLASS = gsub("Cholesterol and derivatives", "Cholesterols and derivatives", SUB_CLASS),
         Ontology = ifelse(Ontology == "Phosphatidylglycerols", "PG",
                                ifelse(Ontology == "Phosphatidylethanolamines", "PE",
                                       ifelse(Ontology == "Phosphatidylserines", "PS", 
                                              ifelse(Ontology == "Phosphatidylinositols", "PI", Ontology)))))

msdial_annotations <- read_excel("data/new/MSDIAL_annotations.xlsx", sheet = "Data") %>% distinct() %>% dplyr::rename(Ontology = info.Ontology) %>%
  filter(Ontology != "ST") %>%
  mutate(SUB_CLASS = ifelse(grepl("erine|itol|glycerol|amine|choline", SUB_CLASS) & 
                              !grepl(paste0(paste0(c("erine", "itol", "glycerol", "amine", "choline"), "s"), collapse = "|"), SUB_CLASS), paste0(SUB_CLASS, "s"), SUB_CLASS),
         SUB_CLASS = gsub("ss", "s", SUB_CLASS),
         SUB_CLASS = gsub("Cholesterol and derivatives", "Cholesterols and derivatives", SUB_CLASS)) %>% 
  as.data.frame()

metab_info_se_ur <- metab_info_se_ur %>% 
  mutate(Ontology = as.character(Ontology),
    Ontology = ifelse(Ontology == "null", NA, Ontology)) %>%
  left_join(msdial_annotations) %>%
    mutate(CATEGORY = str_squish(gsub("\\[.*?\\]", "", CATEGORY)),
         CATEGORY = gsub("Fatty Acyls", "Fatty acyls", CATEGORY),
         CATEGORY = gsub("Prenol Lipids", "Prenol lipids", CATEGORY),
         MAIN_CLASS = str_squish(gsub("\\[.*?\\]", "", MAIN_CLASS)),
         SUB_CLASS = str_squish(gsub("\\[.*?\\]", "", SUB_CLASS)) )

rownames(metab_info_se_ur) <- metab_info_se_ur$MetabID
dim(metab_info_se_ur) #2757
```


```{r}
# Separate info into sample type
a <- metab_quality[which(metab_quality$Urine == TRUE & metab_quality$Peak_quality == TRUE),] %>% pull(MetabID)
metab_info_ur <- metab_info_se_ur %>% filter(MetabID %in% a) %>% as.data.frame()
a <- metab_quality[which(metab_quality$Serum == TRUE & metab_quality$Peak_quality == TRUE),] %>% pull(MetabID)
metab_info_se <- metab_info_se_ur %>% filter(MetabID %in% a) %>% as.data.frame()

# Filter assay tables
metab_ur_df <- as.data.frame(assay(SE_list_proc$urine$glog))[metab_info_ur$MetabID, ]
metab_se_df <- as.data.frame(assay(SE_list_proc$serum$glog))[metab_info_se$MetabID, ]

identical(rownames(metab_ur_df), metab_info_ur$MetabID) # TRUE
identical(rownames(metab_se_df), metab_info_se$MetabID) # TRUE

rownames(metab_ur_df) <- metab_info_ur$shortname
rownames(metab_se_df) <- metab_info_se$shortname

rownames(metab_info_ur) <- metab_info_ur$shortname
rownames(metab_info_se) <- metab_info_se$shortname

dim(metab_ur_df) #previously 1379 now 2415
dim(metab_se_df) #previously 481 now 1034
```

## Combine Urine and Serum

```{r}
length(intersect(metab_info_ur$shortname, metab_info_se$shortname)) # 692 overlapping metabolites
```

```{r}
# Filter raw dataset by common metabolites between both sample types and include all samples across sample types, then scale and normalise
overlapping_metabolites <- metab_info_se_ur %>% filter(Serum == TRUE & Urine == TRUE) %>% pull(MetabID)
metab_obj_filt <- metab_obj[overlapping_metabolites, c(colnames(metab_ur_df), colnames(metab_se_df))] #metab_info_se_ur$MetabID
metab_obj_norm <- norm_scale_function(metab_obj_filt)
saveRDS(metab_obj_norm, paste0(drive_path, "data/rstudio/metab_obj_norm.rds") )

metab_se_ur_df <- data.frame(assay(metab_obj_norm$glog)) %>% dplyr::select(!contains("QC")) %>% dplyr::select(!contains("ur_13"))
metab_info_se_ur <- metab_info_se_ur %>% filter(MetabID %in% rownames(metab_se_ur_df))
identical(rownames(metab_se_ur_df), rownames(metab_info_se_ur)) # TRUE
rownames(metab_se_ur_df) <- metab_info_se_ur$shortname
dim(metab_se_ur_df) #692   30

# Remove QC samples
metab_ur_df <- metab_ur_df %>% dplyr::select(!contains("QC")) %>% dplyr::select(!contains("ur_13"))
metab_se_df <- metab_se_df %>% dplyr::select(!contains("QC"))

dim(metab_ur_df) # 2415 13
dim(metab_se_df) # 1034 17

length(unique(c(rownames(metab_ur_df), rownames(metab_se_df))))

# Remove QC samples
#SE_list_proc$serum$glog <- SummarizedExperiment::subset(SE_list_proc$serum$glog, select = colData(SE_list_proc$serum$glog)$class != "QC") 

metadata_urine <- metadata_urine %>% 
  filter(SampleNumber %in% colnames(metab_ur_df)) %>% 
  dplyr::select(SampleNumber, SampleID, Cage, Group, Weight..g., Vol..Collected..uL., Class) %>% 
  mutate(Class = factor(Class, levels = c("Homo-Ctrl", "Homo-ALA", "Hetero-Ctrl")) )
metadata_serum <- metadata_serum %>% 
  filter(SampleNumber %in% colnames(metab_se_df)) %>% 
  dplyr::select(SampleNumber, SampleID, Cage, Group, Weight..g., Vol..Collected..uL., Class) %>%
  mutate(Class = factor(Class, levels = c("Homo-Ctrl", "Homo-ALA", "Hetero-Ctrl")) )

# Add stones metadata
metadata_urine <- metadata_urine %>%
  left_join(read.csv(paste0(drive_path, "stones-metadata.csv")))
metadata_serum <- metadata_serum %>%
left_join(read.csv(paste0(drive_path, "stones-metadata.csv")))

metadata_ur_se <- metadata_urine %>% 
  mutate(SampleType = "Urine") %>% 
  full_join(metadata_serum %>% mutate(SampleType = "Serum")) %>%
  mutate(Class = as.character(Class),
         Group_type = paste0(Class, "-", SampleType),
         Group_type = factor(Group_type, levels = c("Homo-ALA-Serum", "Homo-ALA-Urine",
                                               "Homo-Ctrl-Serum", "Homo-Ctrl-Urine",
                                               "Hetero-Ctrl-Serum","Hetero-Ctrl-Urine") ),
         Class = factor(Class, levels = c("Hetero-Ctrl", "Homo-Ctrl", "Homo-ALA")))

saveRDS(metab_se_ur_df, paste0(drive_path, "rstudio/metab_se_ur_df.rds") )
saveRDS(metab_ur_df, paste0(drive_path, "rstudio/metab_ur_df.rds") )
saveRDS(metab_se_df, paste0(drive_path, "rstudio/metab_se_df.rds") )

write.csv(metab_ur_df, paste0(drive_path, "rstudio/metab_ur_df.csv") )
write.csv(metab_se_df, paste0(drive_path, "rstudio/metab_se_df.csv") )


saveRDS(metab_info_ur, paste0(drive_path, "rstudio/metab_info_ur.rds") )
saveRDS(metab_info_se, paste0(drive_path, "rstudio/metab_info_se.rds") )
saveRDS(metab_info_se_ur, paste0(drive_path, "rstudio/metab_info_se_ur.rds") )

write.csv(metab_info_ur, paste0(drive_path, "rstudio/metab_info_ur.csv"))
write.csv(metab_info_se, paste0(drive_path, "rstudio/metab_info_se.csv"))
write.csv(master_info_annotated_filtered, paste0(drive_path, "rstudio/master_info_annotated_filtered.csv"))

saveRDS(metadata_urine, paste0(drive_path, "rstudio/metadata_urine.rds") )
saveRDS(metadata_serum, paste0(drive_path, "rstudio/metadata_serum.rds") )
saveRDS(metadata_ur_se, paste0(drive_path, "rstudio/metadata_ur_se.rds") )
```

```{r}
#mmu <- buildGraphFromKEGGREST(organism = "mmu")
#buildDataFromGraph(keggdata.graph = mmu, matrices = c("hypergeom", "diffusion", "pagerank"), normality = c("diffusion", "pagerank"))
#mmu.hsa <- loadKEGGdata(loadMatrix = c("hypergeom", "diffusion", "pagerank"))
#saveRDS(mmu.hsa, "limma/mmu.hsa.rds")
```

# Check misannotation

```{r}
# Urine
## Homo vs Het 
a <- metab_info_ur %>% left_join(master_info_annotated_filtered[, c("MetabID", "Adduct.type")]) %>%
  relocate(Adduct.type) %>%
  filter(!Adduct.type %in% c("[M+H]+", "[M-H]-")) %>%
  dplyr::rename(Molecule = shortname) %>%
  left_join(limma_list$urine$HomoCtrl_vs_HeteroCtrl$sig_hits[, c("Molecule", "direction")]) %>%
  filter(!is.na(direction), !is.na(HMDB))
write.csv(a, "urine-homo-het.csv", row.names = FALSE)

## Homo vs ALA 
a <- metab_info_ur %>% left_join(master_info_annotated_filtered[, c("MetabID", "Adduct.type")]) %>%
  relocate(Adduct.type) %>%
  filter(!Adduct.type %in% c("[M+H]+", "[M-H]-")) %>%
  dplyr::rename(Molecule = shortname) %>%
  left_join(limma_list$urine$HomoALA_vs_HomoCtrl$sig_hits[, c("Molecule", "direction")]) %>%
  filter(!is.na(direction), !is.na(HMDB))
write.csv(a, "urine-ala-homo.csv", row.names = FALSE)

# Serum
## Homo vs Het 
a <- metab_info_se %>% left_join(master_info_annotated_filtered[, c("MetabID", "Adduct.type")]) %>%
  relocate(Adduct.type) %>%
  filter(!Adduct.type %in% c("[M+H]+", "[M-H]-")) %>%
  dplyr::rename(Molecule = shortname) %>%
  left_join(limma_list$serum$HomoCtrl_vs_HeteroCtrl$sig_hits[, c("Molecule", "direction")]) %>%
  filter(!is.na(direction), !is.na(HMDB))
write.csv(a, "serum-homo-het.csv", row.names = FALSE)

## Homo vs ALA 
a <- metab_info_se %>% left_join(master_info_annotated_filtered[, c("MetabID", "Adduct.type")]) %>%
  relocate(Adduct.type) %>%
  filter(!Adduct.type %in% c("[M+H]+", "[M-H]-")) %>%
  dplyr::rename(Molecule = shortname) %>%
  left_join(limma_list$serum$HomoALA_vs_HomoCtrl$sig_hits[, c("Molecule", "direction")]) %>%
  filter(!is.na(direction), !is.na(HMDB))
write.csv(a, "serum-ala-homo.csv", row.names = FALSE)
```


